---
title: "SEM diagnostic — sensitivity analysis"
author: "Emma Belaud"
format:
  html:
    css: css/styles.css  
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
execute: 
  warning: false
  message: false
  echo: true
---

# Overview

## Role in the pipeline

This script is the **second step** of the SEM pipeline, following:

1.  `data_preparation.qmd` — assembles the full SEM database

Its outputs are consumed directly by:

3.  `check_models_performance.qmd` — full rolling-window SEM execution
4.  `results_visualisation.qmd` — figure production and ecological interpretation

## What this script produces

| Output | File | Description |
|---------------------|------------------|---------------------------------|
| Full benchmark table | `benchmark_results_database.csv` | Diagnostic statistics for every parameter combination × response variable |
| Selected model parameters | `model_parameters.txt` | Optimal window duration, completeness threshold, and per-variable transformations, saved as JSON |

## Statistical approach

The causal model has two structural equations:

$$\text{Root} \leftarrow \beta_1 \cdot \text{microclimate}_1 + \beta_2 \cdot \text{microclimate}_2$$ $$\text{Fauna} \leftarrow \beta_3 \cdot \text{microclimate}_1 + \beta_4 \cdot \text{microclimate}_2 + \beta_5 \cdot \text{Root}$$

Both equations use a **nested random intercept** (`orientation / depth`) and an **AR(1) correlation structure** to account for temporal autocorrelation within scanner groups. Before fitting, all variables are z-score standardised so path coefficients are directly comparable across taxa and predictors.

## Sensitivity analysis rationale

Three structural choices must be validated before running the full SEM:

| Parameter | Question | Values tested |
|----------------------|--------------------|------------------------------|
| **Window duration** | How many days of data are needed for a stable fit? | 5, 7, 10, 14, 21, 28, 30 days |
| **Min. data completeness** | How many missing images can be tolerated per window? | 10 % – 50 % |
| **Transformation** | Which variance-stabilising transform best normalises counts? | none, Freeman-Tukey, log, Hellinger |

These three parameters are crossed in a **full factorial grid** (7 × 5 × 4 = 140 combinations). The same transformation is applied simultaneously to fauna counts, root growth flux, and microclimate indices within each run.

| Diagnostic | Metric | Threshold |
|---------------------------|-------------------|--------------------------|
| Heteroscedasticity | Spearman ρ between \|residuals\| and fitted values | \< 0.4 |
| Skewness | Absolute skewness of normalised residuals | \< 1.5 |
| Collinearity | Maximum VIF across predictors | \< 5 |
| Explanatory power | Marginal and conditional R² | — |

## Selection criterion

The optimal configuration is chosen by a **two-step procedure**:

1.  For each response variable (root, fauna) **independently**, select the transformation maximising `prop_valid_models × avg_R²_marginal` — balancing diagnostic validity with explanatory power.
2.  Across both responses **jointly**, find the (window, min_rate) pair maximising `combined_validity × mean_R²`, where `combined_validity = prod(prop_valid_root, prop_valid_fauna)`. Using the product rather than the mean ensures that neither response can mask a poor fit in the other.

------------------------------------------------------------------------

## 1. Setup

```{r setup}
#| message: false

# ── Locale ────────────────────────────────────────────────────────────────────
invisible(Sys.setlocale("LC_TIME", "C"))

# ── Libraries ─────────────────────────────────────────────────────────────────
libs <- c(
  "tidyverse",     # Data wrangling and visualisation
  "lubridate",     # Date/time manipulation
  "piecewiseSEM",  # Piecewise structural equation modelling
  "performance",   # Model performance metrics
  "lme4",          # Linear mixed-effects models
  "purrr",         # Functional iteration (map, keep, compact)
  "nlme",          # LME with correlation structures (corAR1)
  "moments",       # Skewness and kurtosis
  "car",           # VIF via variance inflation factors
  "knitr",         # Table rendering
  "kableExtra"     # Enhanced HTML tables
)
invisible(lapply(libs, require, character.only = TRUE))

# ── Global diagnostic thresholds ──────────────────────────────────────────────
# Define the statistical criteria for declaring a model window "valid"
HETERO_THRESHOLD_MODERATE <- 0.4   # Max Spearman rho (|residuals| vs fitted) for heteroscedasticity
SKEWNESS_THRESHOLD        <- 1.0   # Max absolute skewness of normalised residuals
VIF_THRESHOLD             <- 5.0   # Max variance inflation factor (collinearity limit)
MIN_RATE_DEFAULT          <- 0.3   # Default minimum data completeness per window

# ── Data loading ──────────────────────────────────────────────────────────────
data_sem_raw <- read.csv("../output/SEM_database.csv",
                         sep       = ",",
                         na.strings = c("", "NA")) %>%
  mutate(
    position = as.factor(position),
    depth    = as.factor(depth),
    land_use = as.numeric(position == "C"),
    date     = ymd_hms(date, tz = "UTC"),
    # Internal SEM aliases (avoids renaming columns throughout the pipeline)
    fauna = all_taxa,
    root  = root_growth_cm2_count_24h   # 24-hour root growth flux (cm²)
  )

# ── Metadata ──────────────────────────────────────────────────────────────────
# Taxa names and selected microclimate indices exported by data_preparation.qmd
fauna_vars         <- c("all_taxa", readLines("../output/fauna_vars.txt"))
microclimate_index <- readLines("../output/microclimate_index.txt")
```

------------------------------------------------------------------------

## 2. Functions

### 2.1 Rolling window splitter

```{r fn-rolling-windows}
#' Split a time series into overlapping rolling windows
#'
#' @param df          Dataframe with a `date` column (POSIXct).
#' @param window_days Integer. Duration of each window in days.
#' @param step_days   Integer. Step between consecutive window start dates.
#' @return Named list of dataframes, one per window start date.
create_rolling_windows <- function(df, window_days, step_days = 1) {
  start_dates <- seq(
    min(df$date),
    max(df$date) - days(window_days) + hours(6),
    by = paste(step_days, "days")
  )

  windows       <- map(start_dates, ~ filter(df, date >= .x & date < .x + days(window_days)))
  names(windows) <- format(start_dates, "%Y-%m-%d")
  return(windows)
}
```

### 2.2 SEM structure

The causal model has two equations:

-   **Root** ← microclimate₁ + microclimate₂
-   **Fauna** ← microclimate₁ + microclimate₂ + root

Both use `lme` with a nested random intercept `orientation / depth` and an AR(1) correlation structure to account for temporal autocorrelation within scanner groups.

```{r fn-fit-psem}
#' Fit the piecewise SEM for a single data window
#'
#' @param data      Dataframe with columns: `root`, `fauna`, `microclimate_1`,
#'                  `microclimate_2`, `orientation`, `depth`.
#' @param m_indices Character vector of length 2 (microclimate index names).
#'                  Used by the wrapper — internally mapped to `microclimate_1/2`.
#' @return A `psem` object.
fit_psem <- function(data, m_indices) {
  psem(
    # ── Tier 1: land use drives microclimate ──────────────────────────────
    lme(microclimate_1 ~ land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    lme(microclimate_2 ~ land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    # ── Tier 2: microclimate + land use drive root ─────────────────────────
    lme(root ~ microclimate_1 + microclimate_2 + land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    # ── Tier 3: everything drives fauna ───────────────────────────────────
    lme(fauna ~ microclimate_1 + microclimate_2 + root + land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim"))
  )
}
```

### 2.3 Data transformation functions

Four variance-stabilising transformations are tested against the untransformed baseline. All are applied to fauna counts and root growth values before modelling.

```{r fn-transforms}
#' Freeman-Tukey transform: sqrt(x) + sqrt(x + 1)
#' Preferred for count data with many zeros; stabilises variance across the mean.
transform_ft <- function(x) sqrt(x) + sqrt(x + 1)

#' Log transform: log(x + 1)
#' Classical approach for right-skewed data; log1p avoids issues with zero values.
transform_log <- function(x) log1p(x)

#' Hellinger transform: sqrt(x / sum(x))
#' Minimises the influence of outliers; applied locally within each window
#' so that the scale remains window-relative.
transform_hellinger <- function(x) sqrt(x / sum(x, na.rm = TRUE))
```

### 2.4 Main SEM pipeline

```{r fn-pipeline}
#' Run the rolling-window SEM pipeline for one taxon variable
#'
#' For each position and each valid time window, fits the SEM and extracts
#' R², heteroscedasticity, skewness, AR(1) phi, and VIF diagnostics.
#'
#' @param taxon_var   Character. Column name of the fauna variable to model.
#' @param data_raw    Dataframe. Full SEM database (already transformed if needed).
#' @param m_indices   Character vector of length 2. Microclimate index column names.
#' @param window_days Integer. Duration of each rolling window (days).
#' @param step_days   Integer. Step between window start dates (days).
#' @param min_rate    Numeric [0, 1]. Minimum data completeness ratio per window.
#'                    Windows below this threshold are discarded.
#'                    (Total possible slots = window_days × 16 readings per day.)
#' @param n_windows   Integer or NULL. If set, randomly subsample this many windows
#'                    per position (useful for benchmarking speed).
#' @return Dataframe with one row per model (window × position × response).
run_sem_pipeline <- function(taxon_var, data_raw, m_indices,
                             window_days = 14, step_days = 1,
                             min_rate = 0.3, n_windows = NULL) {

  # Rename dynamic columns to the fixed SEM internal names
  sem_data <- data_raw %>%
    mutate(
      fauna          = !!sym(taxon_var),
      microclimate_1 = !!sym(m_indices[1]),
      microclimate_2 = !!sym(m_indices[2]),
      depth          = as.factor(depth)
    ) %>%
    select(date, land_use, orientation, depth,
           fauna, root, microclimate_1, microclimate_2)

    windows <- create_rolling_windows(sem_data, window_days, step_days)

    # Discard windows with insufficient data density
    n_positions <- length(unique(sem_data$land_use)) 
    valid_windows <- keep(windows,
                          ~ (nrow(drop_na(.x)) / (window_days * 16)) >= min_rate)

    # Optional: random subsample for speed benchmarking
    if (!is.null(n_windows) && length(valid_windows) > n_windows) {
      valid_windows <- valid_windows[sample(seq_along(valid_windows), n_windows)]
    }

    results_list <- map(names(valid_windows), function(w_name) {
      win_data <- valid_windows[[w_name]] %>%
          drop_na() %>%
          arrange(land_use, orientation, depth, date)
      
      # Guard: skip windows where ANY modelled variable is constant
      # (zero variance causes Cholesky failures in lme and summary.psem)
      vars_to_check <- c("fauna", "root", "microclimate_1", "microclimate_2")
      if (any(sapply(win_data[vars_to_check], sd, na.rm = TRUE) == 0)) return(NULL)
      
      # Fit the SEM; return NULL silently on convergence failure
      model <- tryCatch(withCallingHandlers(
          fit_psem(win_data, m_indices),
          warning = function(w) invokeRestart("muffleWarning")),
        error = function(e) NULL)
      
      if (is.null(model)) return(NULL)

      # ── R² extraction ──────────────────────────────────────────────────────
      # summary.psem() can itself trigger chol.default() failures — wrap it too
      sm <- tryCatch(summary(model), error = function(e) NULL)
          if (is.null(sm)) return(NULL)
      r2_raw <- as.data.frame(sm$R2)

      # Column names vary across piecewiseSEM versions — detect dynamically
      col_cond <- grep("R.squared|Conditional", names(r2_raw), value = TRUE)[1]
      col_marg <- grep("Marginal",              names(r2_raw), value = TRUE)[1]

      r2_df <- r2_raw %>%
        mutate(r2_cond     = .[[col_cond]],
               r2_marginal = .[[col_marg]]) %>%
        select(response = Response, r2_cond, r2_marginal)

      # ── Residual diagnostics ───────────────────────────────────────────────
      lme_mods  <- keep(model, inherits, "lme")

      map_dfr(lme_mods, function(m) {
        res <- residuals(m, type = "normalized")
        fit <- fitted(m)

        # AR(1) phi: autocorrelation coefficient of the within-group error
        # tryCatch guards against non-positive-definite variance matrices
        phi_val <- tryCatch(
          coef(m$modelStruct$corStruct, unconstrained = FALSE),
          error = function(e) NA )

        # VIF: only meaningful when there are ≥ 2 predictors
        vif_val <- tryCatch({
          fixed_terms <- labels(terms(m))
          # land_use is exogenous by design — exclude from collinearity check
          non_treatment_terms <- setdiff(fixed_terms, "land_use")
          if (length(non_treatment_terms) > 1) {
            m_no_lu <- lm(formula(m), data = win_data)
            vif_sub  <- car::vif(m_no_lu)
            max(vif_sub[names(vif_sub) != "land_use"])
          } else 1
          }, error = function(e) NA)

        tibble(
          model_id        = paste(taxon_var,  w_name, sep = "_"),
          response        = all.vars(formula(m))[1],
          n_obs           = length(res),
          vif_max         = vif_val,
          hetero_strength = cor(abs(res), fit, method = "spearman",
                                use = "complete.obs"),
          skewness        = moments::skewness(res, na.rm = TRUE),
          ar1_phi         = as.numeric(phi_val)
        ) %>%
          left_join(r2_df, by = "response")
      })                            
    }) %>%                          
    compact()

  return(map_dfr(results_list, ~ .x))
}
```

------------------------------------------------------------------------

## 3. Benchmarking

### 3.1 Configuration tester

Applies one (window × min_rate × transformation) combination and returns aggregated diagnostic summaries across 50 randomly sampled windows.

```{r fn-test-config}
#' Test a single parameter configuration and summarise diagnostic outcomes
#'
#' @param w_days      Integer. Window duration (days).
#' @param m_rate_val  Numeric. Minimum data completeness ratio.
#' @param transfo_type Character. One of: "none", "sqrt_plus_one", "log", "hellinger".
#'                     Applied uniformly to fauna, root, AND microclimate indices.
#' @param data_raw    Full SEM dataframe.
#' @param taxon       Character. Fauna variable column name.
#' @param m_indices   Character vector of length 2. Microclimate index names.
#' @return Dataframe with per-response summary statistics, or NULL if no valid windows.
test_configuration <- function(w_days, m_rate_val, transfo_type,
                                data_raw, taxon, m_indices) {
  df_iter <- data_raw

  # ── A. Apply transformation ────────────────────────────────────────────────
  # Fauna and root: all four transformations are ecologically valid (count data)
  biological_vars <- c(fauna_vars, "root")
  df_iter <- df_iter %>%
    mutate(across(all_of(biological_vars), function(x) {
      switch(transfo_type,
        "sqrt_plus_one" = transform_ft(x),
        "log"           = transform_log(x),
        "hellinger"     = transform_hellinger(x),
        x
      )
    }))
  
  # Microclimate indices: only log or none are meaningful for continuous variables.
  # Freeman-Tukey and Hellinger are count-specific and are not applied here.
  microclimate_transfo <- if (transfo_type %in% c("log", "none")) transfo_type else "none"
  df_iter <- df_iter %>%
    mutate(across(all_of(intersect(m_indices, names(.))), function(x) {
      switch(microclimate_transfo,
        "log" = transform_log(x),
        x
      )
    }))


  # ── B. Z-score standardisation of all numeric variables ────────────────────
  df_iter <- df_iter %>%
    mutate(across(where(is.numeric) & !all_of("land_use"), ~ as.numeric(scale(.))))

  # ── C. Run pipeline on 50 randomly sampled windows ────────────────────────
  res_diag <- run_sem_pipeline(
    taxon, df_iter, m_indices,
    window_days = w_days,
    min_rate    = m_rate_val,
    n_windows   = 50
  )

  if (is.null(res_diag) || nrow(res_diag) == 0) return(NULL)

  # ── D. Aggregate and score each window ────────────────────────────────────
  res_diag %>%
    mutate(
      # A window is "valid" if it passes all three diagnostic thresholds
      is_ok = abs(hetero_strength) < HETERO_THRESHOLD_MODERATE &
              abs(skewness)        < (SKEWNESS_THRESHOLD + 0.5) &
              (vif_max < VIF_THRESHOLD | is.na(vif_max))
    ) %>%
    group_by(response) %>%
    summarise(
      window               = w_days,
      min_rate             = m_rate_val,
      transformation       = transfo_type,
      hetero_strength      = mean(hetero_strength, na.rm = TRUE),
      skewness             = mean(skewness, na.rm = TRUE),
      vif_max              = mean(vif_max, na.rm = TRUE),
      prop_valid_models    = mean(is_ok, na.rm = TRUE),
      avg_r2_marginal      = mean(r2_marginal, na.rm = TRUE),
      avg_r2_conditional   = mean(r2_cond, na.rm = TRUE),
      # delta_r2: variance explained by random effects (spatial + temporal noise)
      delta_r2             = mean(r2_cond - r2_marginal, na.rm = TRUE),
      spatial_random_ratio = mean(r2_cond - r2_marginal, na.rm = TRUE),
      .groups = "drop"
    )
}
```

### 3.2 Define the sensitivity grid and run

```{r benchmarking, cache=TRUE}
# Full factorial grid of parameter combinations to test
sensibility_grid <- expand.grid(
  w_days     = c(5, 7, 10, 14, 21, 28, 30),
  m_rate_val = c(0.1, 0.2, 0.3, 0.4, 0.5),
  transfo    = c("none", "sqrt_plus_one", "log", "hellinger"),
  stringsAsFactors = FALSE
)

# Run all configurations (pmap iterates over each row of the grid)
quiet_test <- purrr::quietly(test_configuration)

benchmark_results <- sensibility_grid %>%
  purrr::pmap_dfr(~ quiet_test(
    ..1, ..2, ..3,
    data_sem_raw, fauna_vars[1], microclimate_index
  )$result)
```

------------------------------------------------------------------------

## 4. Results

### 4.1 Global optimisation — independent transformations per response

The best transformation for each response variable is selected independently - root and fauna: all four transformations are candidates - microclimate_1 and microclimate_2: only "none" and "log" are meaningful (Freeman-Tukey and Hellinger are count-specific) A common structural configuration (window, min_rate) is then chosen based on the joint validity score across all four responses.

```{r global-optimization}
# Step 1: For each (window, min_rate, response), keep the transformation with
#         the best trade-off between validity rate and marginal R²
best_per_variable <- benchmark_results %>%
  group_by(window, min_rate, response) %>%
  slice_max(prop_valid_models * avg_r2_marginal, n = 1, with_ties = FALSE) %>%
  ungroup()

# Step 2: Aggregate across all four responses to find the best common (window, min_rate)
# Both microclimate indices receive the same transformation (log or none),
# so we extract it from microclimate_1 only and apply it to both.
global_optimization <- best_per_variable %>%
  group_by(window, min_rate) %>%
  summarise(
   # Biological validity: product over root + fauna only
    combined_validity    = min(prop_valid_models[response %in% c("root", "fauna")]),
    # Microclimate validity: reported separately, not penalising the score
    validity_microclimate = mean(prop_valid_models[response %in% 
                                   c("microclimate_1", "microclimate_2")]),
    mean_r2              = mean(avg_r2_marginal[response %in% c("root", "fauna")]),
    transfo_root         = transformation[response == "root"],
    transfo_fauna        = transformation[response == "fauna"],
    transfo_microclimate = {
      raw <- transformation[response == "microclimate_1"]
      if (raw %in% c("log", "none")) raw else "none"
    },
    .groups = "drop"
  ) %>%
  mutate(global_score = combined_validity * mean_r2) %>%
  arrange(desc(global_score))


# Step 3: Display top 5 configurations
global_optimization %>%
  head(5) %>%
  mutate(
    across(c(combined_validity, mean_r2, global_score), ~ round(., 3))
  ) %>%
  select(
    "Window (days)"        = window,
    "Max missing data"     = min_rate,
    "Transfo root"         = transfo_root,
    "Transfo fauna"        = transfo_fauna,
    "Transfo microclimate" = transfo_microclimate,
    "Validity (joint)"     = combined_validity,
    "R² Marginal (mean)"   = mean_r2,
    "Score"                = global_score
  ) %>%
  kable(caption = "Top 5 global configurations (common window/rate, response-specific transformations)") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position   = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  row_spec(1, bold = TRUE, background = "#FEC288")
```

------------------------------------------------------------------------

## 5. Export

### 5.1 Full benchmark results

```{r results-writing}
write.csv(benchmark_results,
          "../output/benchmark_results_database.csv",
          row.names = FALSE,
          quote     = TRUE)
```

### 5.2 Selected model parameters

The best configuration is saved as JSON for use in downstream modelling scripts. `transform_microclimate` applies to both indices — they are transformed identically.

```{r export-parameters}
final_selection <- global_optimization %>%
  slice_max(global_score, n = 1)

params_to_save <- list(
  selected_window   = final_selection$window,
  selected_min_rate = final_selection$min_rate,
  transform_root    = final_selection$transfo_root,
  transform_fauna   = final_selection$transfo_fauna,
  transform_microclimate = final_selection$transfo_microclimate
)

jsonlite::write_json(params_to_save,
                     "../output/model_parameters.txt",
                     pretty = TRUE)
```
