---
title: "SEM diagnostic — sensitivity analysis"
author: "Emma Belaud"
format:
  html:
    css: css/styles.css
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
execute:
  warning: false
  message: false
  echo: true
---

# Overview

## Role in the pipeline

This script is the **second step** of the SEM pipeline, following:

1.  `data_preparation.qmd` — assembles the full SEM database

Its outputs are consumed directly by:

3.  `check_models_performance.qmd` — full rolling-window SEM execution
4.  `results_visualisation.qmd` — figure production and ecological interpretation

## What this script produces

| Output | File | Description |
|------------------------|------------------------|------------------------|
| Full benchmark table | `benchmark_results_database.csv` | Diagnostic statistics for every parameter combination x response variable |
| Selected model parameters | `model_parameters.txt` | Optimal window duration, completeness threshold, and transformation, saved as JSON |

## Statistical approach

The causal model has four structural equations organised in three tiers, with
`land_use` (0 = unmanaged, 1 = managed) as the exogenous binary driver.
At the diagnostic stage, each equation is fitted **independently as a classical
linear mixed model** (LME) rather than within the full piecewise SEM structure.
This isolates the residual behaviour of each response variable and avoids
confounding model-fitting issues across equations. Interaction terms are excluded
at this stage; the full SEM is additive throughout — no interaction terms are used in any script.
the optimal window and transformation have been determined.

**Tier 1 — land use shapes microclimate** (additive):

$$\text{microclimate}_1 \leftarrow \beta_1 \cdot \text{land\_use}$$
$$\text{microclimate}_2 \leftarrow \beta_2 \cdot \text{land\_use}$$

**Tier 2 — microclimate drives root:**

$$\text{Root} \leftarrow \beta_3 \cdot \text{mc}_1 + \beta_4 \cdot \text{mc}_2 + \beta_5 \cdot \text{lu}$$

**Tier 3 — microclimate and root drive fauna:**

$$\text{Fauna} \leftarrow \beta_6 \cdot \text{mc}_1 + \beta_7 \cdot \text{mc}_2 + \beta_8 \cdot \text{Root} + \beta_9 \cdot \text{lu}$$

*where mc = microclimate index, lu = land_use.*

All equations use a **nested random intercept** (`orientation / depth`) and an
**AR(1) correlation structure** to account for temporal autocorrelation within
scanner groups.

## Sensitivity analysis rationale

Three structural choices must be validated before running the full SEM:

| Parameter | Question | Values tested |
|------------------------|------------------------|------------------------|
| **Window duration** | How many days are needed for a stable fit? | 5, 7, 10, 14, 21 days |
| **Min. data completeness** | How many missing images can be tolerated? | 10% to 50% |
| **Transformation** | Which transform best normalises the variables? | none, sqrt, Freeman-Tukey, log, Hellinger |

These parameters are crossed in a **full factorial grid** (5 × 4 × 5 = 100
combinations). The same transformation type is applied uniformly to all modelled
variables within each configuration, but **z-score standardisation is performed
independently within each rolling window** after the transformation. This ensures
that standardised coefficients reflect within-window variation rather than the
global distribution of the full time series — which would mix seasonal and
inter-annual variance in an uncontrolled way.

| Diagnostic | Metric | Threshold |
|------------------------|------------------------|------------------------|
| Heteroscedasticity | Spearman rho between abs(residuals) and fitted values | \< 0.4 |
| Skewness | Absolute skewness of normalised residuals | \< 1.5 |
| Collinearity | Maximum VIF across predictors | \< 5 |
| Explanatory power | Marginal and conditional R-squared | — |

> **Note on VIF.** With no interaction terms in the diagnostic models, VIF is
> computed directly on the fitted model formula. `land_use` is excluded from VIF
> calculation as a binary exogenous driver; the check targets collinearity among
> continuous predictors only.

## Selection criterion

The optimal configuration is chosen by a **two-step procedure**:

1.  For each response variable **independently**, select the transformation
    maximising `prop_valid_models × avg_R2_marginal`.
2.  Across biological responses **jointly** (root and fauna), find the
    (window, min_rate) pair maximising `combined_validity × mean_R2`, where
    `combined_validity = min(prop_valid_root, prop_valid_fauna)`. Using the
    minimum ensures that neither response can mask a poor fit in the other.

------------------------------------------------------------------------

## 1. Setup

```{r setup}
#| message: false

invisible(Sys.setlocale("LC_TIME", "C"))

libs <- c(
  "tidyverse", "lubridate",
  "lme4", "purrr", "nlme", "moments", "car",
  "knitr", "kableExtra"
)
invisible(lapply(libs, require, character.only = TRUE))

# ── Diagnostic thresholds ──────────────────────────────────────────────────────
HETERO_THRESHOLD   <- 0.4
SKEWNESS_THRESHOLD <- 1.5
VIF_THRESHOLD      <- 5.0

# ── Data loading ───────────────────────────────────────────────────────────────
data_sem_raw <- read.csv("../output/SEM_database.csv",
                         sep        = ",",
                         na.strings = c("", "NA")) %>%
  mutate(
    position = as.factor(position),
    depth    = as.factor(depth),
    land_use = as.numeric(position == "C"),
    date     = ymd_hms(date, tz = "UTC"),
    fauna    = all_taxa,
    root     = root_growth_cm2_count_24h
  )

fauna_vars         <- c("all_taxa", readLines("../output/fauna_vars.txt"))
microclimate_index <- readLines("../output/microclimate_index.txt")
```

------------------------------------------------------------------------

## 2. Functions

### 2.1 Rolling window splitter

```{r fn-rolling-windows}
#' Split a time series into overlapping rolling windows
#'
#' @param df          Dataframe with a `date` column (POSIXct).
#' @param window_days Integer. Duration of each window in days.
#' @param step_days   Integer. Step between consecutive window start dates.
#' @return Named list of dataframes, one per window start date.
create_rolling_windows <- function(df, window_days, step_days = 1) {
  start_dates <- seq(
    min(df$date),
    max(df$date) - days(window_days) + hours(6),
    by = paste(step_days, "days")
  )
  windows        <- map(start_dates, ~ filter(df, date >= .x & date < .x + days(window_days)))
  names(windows) <- format(start_dates, "%Y-%m-%d")
  return(windows)
}
```

### 2.2 Independent model fitting

Each structural equation is fitted as a standalone LME, without the piecewise
SEM wrapper. This approach is appropriate at the diagnostic stage: residuals
from each equation can be inspected in isolation, and convergence failures in
one model do not propagate to others. No interaction terms are included here —
the diagnostic targets the additive structure only.

```{r fn-fit-models}
#' Fit the four SEM equations independently as classical LME models
#'
#' Each model is fitted separately without the psem() wrapper.
#' No interaction terms — additive structure only for diagnostics.
#'
#' @param data Dataframe with columns: `land_use`, `root`, `fauna`,
#'             `microclimate_1`, `microclimate_2`, `orientation`, `depth`.
#'             Variables must already be transformed and z-scored within window.
#' @return Named list of four `lme` objects (NULL entries on convergence failure).
fit_models_independently <- function(data) {

  lme_ctrl <- lmeControl(msMaxIter = 200, opt = "optim")

  safe_lme <- function(formula) {
    tryCatch(
      withCallingHandlers(
        lme(formula,
            data        = data,
            random      = ~ 1 | orientation / depth,
            correlation = corAR1(form = ~ 1 | orientation / depth),
            control     = lme_ctrl),
        warning = function(w) invokeRestart("muffleWarning")
      ),
      error = function(e) NULL
    )
  }

  list(
    microclimate_1 = safe_lme(microclimate_1 ~ land_use),
    microclimate_2 = safe_lme(microclimate_2 ~ land_use),
    root           = safe_lme(root ~ microclimate_1 + microclimate_2 + land_use),
    fauna          = safe_lme(fauna ~ microclimate_1 + microclimate_2 + root + land_use)
  )
}
```

### 2.3 Transformation functions

```{r fn-transforms}
#' Shift a vector to be strictly positive before a transform that requires x > 0.
#' If min(x) <= 0, shifts the whole vector by |min(x)| + 0.001 and emits a message.
shift_positive <- function(x, label = "") {
  x_min <- min(x, na.rm = TRUE)
  if (!is.na(x_min) && x_min <= 0) {
    shift <- abs(x_min) + 0.001
    message(sprintf("  [shift] '%s' shifted by +%.4f (min was %.4f)", label, shift, x_min))
    x <- x + shift
  }
  x
}

#' Freeman-Tukey: sqrt(x) + sqrt(x + 1). Stabilises variance in count data with zeros.
transform_ft <- function(x) sqrt(x) + sqrt(x + 1)

#' Log: log(x + 1) after shifting to positive if needed.
transform_log <- function(x, label = "") log1p(shift_positive(x, label))

#' Sqrt after shifting to positive if needed.
transform_sqrt <- function(x, label = "") sqrt(shift_positive(x, label))

#' Hellinger: sqrt(x / sum(x)). Window-relative — must be applied per window.
transform_hellinger <- function(x) sqrt(x / sum(x, na.rm = TRUE))

#' Compute marginal and conditional R² for an nlme::lme model.
#'
#' Uses the fitted-value variance approach: R²m = var(fixed-effect predictions)
#' / var(observed); R²c = var(full predictions including random) / var(observed).
#' This avoids parsing the VarCorr character matrix, which is fragile across
#' nlme versions. fitted(m, level = 0) returns population-level predictions
#' (fixed effects only); fitted(m, level = k) adds all k levels of random effects.
#'
#' @param m A fitted nlme::lme object.
#' @return Named numeric vector c(R2m = ..., R2c = ...).
r2_lme <- function(m) {
  tryCatch({
    y_obs <- as.numeric(nlme::getResponse(m))
    y_fix <- as.numeric(fitted(m, level = 0))                # fixed effects only
    y_all <- as.numeric(fitted(m, level = length(m$groups))) # fixed + all random

    ss_obs <- var(y_obs, na.rm = TRUE)
    if (is.na(ss_obs) || ss_obs == 0)
      return(c(R2m = NA_real_, R2c = NA_real_))

    r2m <- var(y_fix, na.rm = TRUE) / ss_obs
    r2c <- var(y_all, na.rm = TRUE) / ss_obs

    # Clamp to [0, 1]: numerical noise can produce tiny negative or >1 values
    c(R2m = min(max(r2m, 0), 1),
      R2c = min(max(r2c, 0), 1))
  }, error = function(e) c(R2m = NA_real_, R2c = NA_real_))
}
```

### 2.4 Main diagnostic pipeline

```{r fn-pipeline}
#' Run the rolling-window diagnostic pipeline for one taxon variable
#'
#' Each window is processed independently:
#'   1. Subset and drop NAs
#'   2. Apply transformation
#'   3. Z-score standardise within the window (excludes land_use)
#'   4. Fit four LME models independently
#'   5. Extract residual diagnostics and R² per model
#'
#' @param taxon_var   Character. Column name of the fauna variable.
#' @param data_raw    Dataframe. Full SEM database (raw, untransformed).
#' @param m_indices   Character vector of length 2. Microclimate index names.
#' @param transfo_type Character. One of "none", "sqrt", "sqrt_plus_one",
#'                    "log", "hellinger".
#' @param window_days Integer. Rolling window duration (days).
#' @param step_days   Integer. Step between window start dates.
#' @param min_rate    Numeric [0, 1]. Minimum completeness ratio.
#' @param n_windows   Integer or NULL. Random subsample size for benchmarking.
#' @return Dataframe with one row per model x window, or NULL.
run_sem_pipeline <- function(taxon_var, data_raw, m_indices, transfo_type,
                             window_days = 14, step_days = 1,
                             min_rate = 0.3, n_windows = NULL) {

  # Rename fauna and microclimate columns to generic names
  sem_data <- data_raw %>%
    mutate(
      fauna          = !!sym(taxon_var),
      microclimate_1 = !!sym(m_indices[1]),
      microclimate_2 = !!sym(m_indices[2]),
      depth          = as.factor(depth)
    ) %>%
    select(date, land_use, orientation, depth,
           fauna, root, microclimate_1, microclimate_2)

  # Split into overlapping windows and retain only sufficiently complete ones
  windows <- create_rolling_windows(sem_data, window_days, step_days)

  valid_windows <- keep(windows,
                        ~ (nrow(drop_na(.x)) / (window_days * 16 * 2)) >= min_rate)

  if (!is.null(n_windows) && length(valid_windows) > n_windows) {
    valid_windows <- valid_windows[sample(seq_along(valid_windows), n_windows)]
  }

  vars_to_process <- c("fauna", "root", "microclimate_1", "microclimate_2")

  results_list <- map(names(valid_windows), function(w_name) {

    win_data <- valid_windows[[w_name]] %>%
      drop_na() %>%
      arrange(land_use, orientation, depth, date)

    # ── A. Apply transformation within this window ───────────────────────────
    # Hellinger is inherently window-relative (sums to 1 within window).
    # Other transforms are element-wise but applied here for consistency.
    win_data <- win_data %>%
      mutate(across(all_of(vars_to_process), function(x) {
        col <- cur_column()
        switch(transfo_type,
          "sqrt"          = transform_sqrt(x, col),
          "sqrt_plus_one" = transform_ft(x),
          "log"           = transform_log(x, col),
          "hellinger"     = transform_hellinger(x),
          x
        )
      }))

    # ── B. Z-score standardise within this window ────────────────────────────
    # land_use is excluded: it is binary and its scale must remain 0/1
    # for coefficient interpretation.
    win_data <- win_data %>%
      mutate(across(all_of(vars_to_process), ~ as.numeric(scale(.))))

    # ── Guards ────────────────────────────────────────────────────────────────
    # sd() returns NA (not 0) when all values in a column are NA after drop_na()
    # (can occur for sparse taxa). is.na() check catches both NA and zero-SD cases.
    stdevs <- sapply(win_data[vars_to_process], sd, na.rm = TRUE)
    if (any(is.na(stdevs) | stdevs == 0)) return(NULL)
    if (length(unique(win_data$land_use)) < 2) return(NULL)
    if (any(!is.finite(as.matrix(win_data[vars_to_process])))) return(NULL)

    # ── C. Fit the four equations independently ───────────────────────────────
    models <- fit_models_independently(win_data)

    # ── D. Extract diagnostics per model ──────────────────────────────────────
    map_dfr(names(models), function(resp_name) {

      m <- models[[resp_name]]
      if (is.null(m)) return(NULL)

      res <- residuals(m, type = "normalized")
      fit <- fitted(m)

      # AR(1) autocorrelation parameter
      phi_val <- tryCatch(
        coef(m$modelStruct$corStruct, unconstrained = FALSE),
        error = function(e) NA_real_
      )

      # R² (marginal = fixed effects only; conditional = fixed + random)
      # r2_lme() uses manual variance partitioning — avoids the MuMIn matrix
      # indexing issue where r.squaredGLMM() returns a matrix for lme objects
      # and named-vector access silently returns NA.
      r2_vals <- r2_lme(m)

      # VIF: computed on continuous predictors only (land_use excluded as binary)
      vif_val <- tryCatch({
        pred_terms <- labels(terms(m))
        cont_terms <- setdiff(pred_terms, "land_use")
        if (length(cont_terms) > 1) {
          f_vif <- as.formula(
            paste(all.vars(formula(m))[1], "~",
                  paste(cont_terms, collapse = " + "))
          )
          max(car::vif(lm(f_vif, data = win_data)))
        } else NA_real_
      }, error = function(e) NA_real_)

      tibble(
        model_id        = paste(taxon_var, w_name, sep = "_"),
        response        = resp_name,
        n_obs           = length(res),
        vif_max         = vif_val,
        hetero_strength = cor(abs(res), fit, method = "spearman",
                              use = "complete.obs"),
        skewness        = moments::skewness(res, na.rm = TRUE),
        ar1_phi         = as.numeric(phi_val),
        r2_marginal     = as.numeric(r2_vals["R2m"]),
        r2_cond         = as.numeric(r2_vals["R2c"])
      )
    })
  }) %>%
    compact()

  if (length(results_list) == 0) return(NULL)
  return(map_dfr(results_list, ~ .x))
}
```

------------------------------------------------------------------------

## 3. Benchmarking

### 3.1 Configuration tester

```{r fn-test-config}
#' Test a single parameter configuration and summarise diagnostic outcomes
#'
#' Transformation type and window parameters are passed to run_sem_pipeline(),
#' which applies them within each window independently. No global scaling
#' is performed at this level.
#'
#' @param w_days       Integer. Window duration (days).
#' @param m_rate_val   Numeric. Minimum data completeness ratio.
#' @param transfo_type Character. One of: "none", "sqrt", "sqrt_plus_one",
#'                    "log", "hellinger".
#' @param data_raw     Full SEM dataframe (raw, untransformed).
#' @param taxon        Character. Fauna variable column name.
#' @param m_indices    Character vector of length 2. Microclimate index names.
#' @return Dataframe with per-response summary statistics, or NULL.
test_configuration <- function(w_days, m_rate_val, transfo_type,
                               data_raw, taxon, m_indices) {

  # run_sem_pipeline handles all transformation and per-window scaling
  res_diag <- run_sem_pipeline(
    taxon_var   = taxon,
    data_raw    = data_raw,
    m_indices   = m_indices,
    transfo_type = transfo_type,
    window_days = w_days,
    min_rate    = m_rate_val,
    n_windows   = 50
  )

  if (is.null(res_diag) || nrow(res_diag) == 0) return(NULL)

  res_diag %>%
    mutate(
      is_ok = abs(hetero_strength) < HETERO_THRESHOLD &
              abs(skewness)        < SKEWNESS_THRESHOLD &
              (is.na(vif_max) | vif_max < VIF_THRESHOLD)
    ) %>%
    group_by(response) %>%
    summarise(
      window             = w_days,
      min_rate           = m_rate_val,
      transformation     = transfo_type,
      hetero_strength    = mean(hetero_strength,       na.rm = TRUE),
      skewness           = mean(skewness,              na.rm = TRUE),
      vif_max            = mean(vif_max,               na.rm = TRUE),
      prop_valid_models  = mean(is_ok,                 na.rm = TRUE),
      avg_r2_marginal    = mean(r2_marginal,           na.rm = TRUE),
      avg_r2_conditional = mean(r2_cond,               na.rm = TRUE),
      delta_r2           = mean(r2_cond - r2_marginal, na.rm = TRUE),
      .groups = "drop"
    )
}
```

### 3.2 Sensitivity grid and execution

```{r benchmarking, cache=TRUE}
sensibility_grid <- expand.grid(
  w_days     = c(5, 7, 10, 14, 21),
  m_rate_val = c(0.1, 0.2, 0.3, 0.4),
  transfo    = c("none", "sqrt", "sqrt_plus_one", "log", "hellinger"),
  stringsAsFactors = FALSE
)

quiet_test <- purrr::safely(test_configuration)

# Run all configurations; safely() captures errors without stopping execution
safe_results <- sensibility_grid %>%
  purrr::pmap(~ quiet_test(
    ..1, ..2, ..3,
    data_sem_raw, fauna_vars[1], microclimate_index
  ))

# Report how many configurations failed (NULL result = no valid windows)
n_failed <- sum(map_lgl(safe_results, ~ !is.null(.x$error)))
if (n_failed > 0) message(n_failed, " configuration(s) returned an error.")

benchmark_results <- map_dfr(safe_results, "result")
```

------------------------------------------------------------------------

## 4. Results

### 4.1 Global optimisation — independent transformation per response variable

The benchmark grid tests one transformation per run, but because diagnostics
are extracted per equation, each response can be evaluated independently. We
select the best transformation for each of the four responses separately using
a lexicographic criterion — validity first, R² second — then find the common
(window, min_rate) that maximises joint validity across all responses.

```{r global-optimization}
EXPECTED_RESPONSES <- c("fauna", "root", "microclimate_1", "microclimate_2")

# ── Step 1: per (window, min_rate, response), best transformation ─────────────
best_per_variable <- benchmark_results %>%
  group_by(window, min_rate, response) %>%
  arrange(desc(prop_valid_models), desc(avg_r2_marginal)) %>%
  slice(1) %>%
  ungroup()

# ── Step 2: best common (window, min_rate) across all four responses ──────────
# Only include (window, min_rate) pairs where all four responses converged.
# If any response is missing entirely, transformation[response == "root"] returns
# character(0) inside summarise(), which crashes or silently drops the row.
# combined_validity = min across responses so a poor fit in one cannot be masked.
global_optimization <- best_per_variable %>%
  group_by(window, min_rate) %>%
  filter(all(EXPECTED_RESPONSES %in% response)) %>%
  summarise(
    combined_validity      = min(prop_valid_models,  na.rm = TRUE),
    mean_r2                = mean(avg_r2_marginal,   na.rm = TRUE),
    transfo_root           = transformation[response == "root"],
    transfo_fauna          = transformation[response == "fauna"],
    transfo_microclimate_1 = transformation[response == "microclimate_1"],
    transfo_microclimate_2 = transformation[response == "microclimate_2"],
    .groups = "drop"
  ) %>%
  # NaN from mean() over all-NA R² — replace with NA so arrange() works cleanly
  mutate(mean_r2 = ifelse(is.nan(mean_r2), NA_real_, mean_r2)) %>%
  arrange(desc(combined_validity), desc(mean_r2)) %>%
  mutate(global_score = combined_validity * mean_r2)

# ── Step 3: display top 5 ─────────────────────────────────────────────────────
global_optimization %>%
  head(5) %>%
  mutate(across(c(combined_validity, mean_r2, global_score), ~ round(., 3))) %>%
  select(
    "Window (days)"          = window,
    "Max missing data"       = min_rate,
    "Transfo root"           = transfo_root,
    "Transfo fauna"          = transfo_fauna,
    "Transfo microclimate 1" = transfo_microclimate_1,
    "Transfo microclimate 2" = transfo_microclimate_2,
    "Validity (joint)"       = combined_validity,
    "R2 Marginal (mean)"     = mean_r2,
    "Score"                  = global_score
  ) %>%
  kable(caption = "Top 5 configurations — ranked by validity then R² (lexicographic)") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE, position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  row_spec(1, bold = TRUE, background = "#FEC288")

final_selection <- global_optimization %>%
  slice(1)
```

------------------------------------------------------------------------

## 4.2 Per-equation performance summary — selected configuration

The table below zooms into the **selected (window, min_rate)** pair and
shows, for each structural equation independently, the diagnostic
statistics obtained with its own best transformation. This makes it
immediately clear which equations drive the joint validity ceiling and
why.

Colour legend:

-   **Validity** — proportion of windows passing all three diagnostic
    thresholds (heteroscedasticity \< 0.4, \|skewness\| \< 1.5, VIF \<
    5). Green ≥ 0.7 / orange 0.4–0.69 / red \< 0.4.
-   **Skewness** — mean absolute skewness of normalised residuals across
    valid windows. Flagged orange when \> 1.5 (threshold exceeded on
    average).
-   **Heteroscedasticity** — mean Spearman ρ between \|residuals\| and
    fitted values. Flagged orange when \> 0.4.
-   **R² marginal** — variance explained by fixed effects only. Low
    values indicate the predictors lack explanatory power beyond random
    structure. No threshold — shown for context.
-   **ΔR²** — conditional minus marginal R². Large values indicate
    strong random effects (random structure absorbs most of the
    variance); small values indicate random structure contributes little.

```{r per-equation-summary}
# ── Retrieve the selected (window, min_rate) ──────────────────────────────────
sel_window   <- final_selection$window
sel_min_rate <- final_selection$min_rate

# Map of response → its selected transformation (from global_optimization row 1)
sel_transfo <- c(
  fauna          = final_selection$transfo_fauna,
  root           = final_selection$transfo_root,
  microclimate_1 = final_selection$transfo_microclimate_1,
  microclimate_2 = final_selection$transfo_microclimate_2
)

# Pull the matching rows from the full benchmark for each response + its transfo
per_eq_stats <- imap_dfr(sel_transfo, function(transfo, resp) {
  benchmark_results %>%
    filter(
      response       == resp,
      window         == sel_window,
      min_rate       == sel_min_rate,
      transformation == transfo
    )
}) %>%
  mutate(
    # ── Display labels ────────────────────────────────────────────────────────
    response_label = case_match(response,
      "fauna"          ~ "Faunal activity",
      "root"           ~ "Root growth",
      "microclimate_1" ~ microclimate_index[1],
      "microclimate_2" ~ microclimate_index[2],
      .default = response
    )
  ) %>%
  arrange(match(response,
                c("fauna", "root", "microclimate_1", "microclimate_2")))

# ── Pre-extract vectors for column_spec (avoids double evaluation) ────────────
validity_vals  <- per_eq_stats$prop_valid_models
skewness_vals  <- abs(per_eq_stats$skewness)
hetero_vals    <- per_eq_stats$hetero_strength
r2_vals_marg   <- per_eq_stats$avg_r2_marginal

validity_col <- case_when(
  validity_vals >= 0.70 ~ "#2d6a4f",
  validity_vals >= 0.40 ~ "#c1603a",
  TRUE                  ~ "#b00020"
)
skewness_col  <- if_else(skewness_vals  > SKEWNESS_THRESHOLD, "#c1603a", "black")
hetero_col    <- if_else(hetero_vals    > HETERO_THRESHOLD,   "#c1603a", "black")

# ── Render table ──────────────────────────────────────────────────────────────
per_eq_stats %>%
  mutate(
    across(c(prop_valid_models, avg_r2_marginal, avg_r2_conditional,
             delta_r2, hetero_strength, skewness), ~ round(., 3)),
    vif_max = round(vif_max, 2)
  ) %>%
  select(
    "Equation"            = response_label,
    "Transform"           = transformation,
    "Validity"            = prop_valid_models,
    "R² marg."            = avg_r2_marginal,
    "R² cond."            = avg_r2_conditional,
    "ΔR²"                 = delta_r2,
    "Skewness"            = skewness,
    "Heterosced."         = hetero_strength,
    "VIF max"             = vif_max
  ) %>%
  kable(
    align = c("l", "c", "c", "c", "c", "c", "c", "c", "c")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE, position = "center", font_size = 13
  ) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  column_spec(3,  color = validity_col, bold = TRUE) %>%
  column_spec(7,  color = skewness_col, bold = skewness_vals > SKEWNESS_THRESHOLD) %>%
  column_spec(8,  color = hetero_col,   bold = hetero_vals   > HETERO_THRESHOLD) 
```


```{r per-equation-barplot}
#| fig-cap: >
#|   Proportion of valid windows per equation under the selected configuration.
#|   The dashed line marks the 0.7 acceptability threshold. Colour encodes the
#|   dominant diagnostic failure mode when validity < 0.7.
#| fig-width: 8
#| fig-height: 3.5

bar_dat <- per_eq_stats %>%
  mutate(
    failure_mode = case_when(
      prop_valid_models >= 0.70                    ~ "Acceptable",
      abs(skewness)     > SKEWNESS_THRESHOLD       ~ "Skewness",
      hetero_strength   > HETERO_THRESHOLD         ~ "Heteroscedasticity",
      !is.na(vif_max) & vif_max > VIF_THRESHOLD    ~ "Collinearity",
      TRUE                                         ~ "Mixed / other"
    ),
    response_label = factor(response_label,
                            levels = rev(response_label))
  )

failure_colours <- c(
  "Acceptable"        = "#2d6a4f",
  "Skewness"          = "#6b9ac4",
  "Heteroscedasticity"= "#c1603a",
  "Collinearity"      = "#9b5de5",
  "Mixed / other"     = "#adb5bd"
)

ggplot(bar_dat, aes(x = prop_valid_models, y = response_label,
                    fill = failure_mode)) +
  geom_col(width = 0.55, colour = "white", linewidth = 0.3) +
  geom_vline(xintercept = 0.70, linetype = "dashed",
             colour = "grey30", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.0f%%", prop_valid_models * 100)),
            hjust = -0.15, size = 3.2, colour = "grey20") +
  scale_x_continuous(
    limits = c(0, 1.08),
    labels = scales::percent_format(accuracy = 1),
    expand = c(0, 0)
  ) +
  scale_fill_manual(values = failure_colours, name = "Dominant issue") +
  labs(
    x = "Proportion of windows passing all diagnostics",
    y = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    legend.position  = "right",
    legend.key.size  = unit(0.4, "cm"),
    panel.grid.major.x = element_line(colour = "grey92", linewidth = 0.3),
    plot.subtitle    = element_text(colour = "grey50", size = 9)
  )
```

------------------------------------------------------------------------

## 5. Export

### 5.1 Full benchmark results

```{r results-writing}
write.csv(benchmark_results,
          "../output/benchmark_results_database.csv",
          row.names = FALSE,
          quote     = TRUE)
```

### 5.2 Selected model parameters

Four separate transform keys are exported, one per response variable.
`check_models_performance.qmd` reads each key and applies the corresponding
transformation before z-scoring within each rolling window. The selected
configuration is row 1 of `global_optimization`, sorted lexicographically
(highest validity, then highest R²).

```{r export-parameters}
params_to_save <- list(
  selected_window          = final_selection$window,
  selected_min_rate        = final_selection$min_rate,
  transform_root           = final_selection$transfo_root,
  transform_fauna          = final_selection$transfo_fauna,
  transform_microclimate_1 = final_selection$transfo_microclimate_1,
  transform_microclimate_2 = final_selection$transfo_microclimate_2
)

jsonlite::write_json(params_to_save,
                     "../output/model_parameters.txt",
                     pretty = TRUE)

message(sprintf(
  paste0("Best config: window = %d days | min_rate = %.1f\n",
         "  validity = %.3f | R2 = %.3f | score = %.3f\n",
         "  root: %s | fauna: %s | mc1: %s | mc2: %s"),
  as.integer(final_selection$window),
  final_selection$min_rate,
  final_selection$combined_validity,
  final_selection$mean_r2,
  final_selection$global_score,
  final_selection$transfo_root,
  final_selection$transfo_fauna,
  final_selection$transfo_microclimate_1,
  final_selection$transfo_microclimate_2
))
```
