---
title: "SEM diagnostic — sensitivity analysis"
author: "Emma Belaud"
format:
  html:
    css: css/styles.css
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
execute:
  warning: false
  message: false
  echo: true
---

# Overview

## Role in the pipeline

This script is the **second step** of the SEM pipeline, following:

1.  `data_preparation.qmd` — assembles the full SEM database

Its outputs are consumed directly by:

3.  `check_models_performance.qmd` — full rolling-window SEM execution
4.  `results_visualisation.qmd` — figure production and ecological interpretation

## What this script produces

| Output | File | Description |
|----|----|----|
| Full benchmark table | `benchmark_results_database.csv` | Diagnostic statistics for every parameter combination x response variable |
| Selected model parameters | `model_parameters.txt` | Optimal window duration, completeness threshold, and transformation, saved as JSON |

## Statistical approach

The causal model has four structural equations organised in three tiers, with `land_use` (0 = unmanaged, 1 = managed) as the exogenous binary driver. **Tiers 2 and 3 include interaction terms** so that the sensitivity analysis evaluates the same model complexity used in `check_models_performance.qmd`.

**Tier 1 - land use shapes microclimate** (additive):

$$\text{microclimate}_1 \leftarrow \beta_1 \cdot \text{land\_use}$$ $$\text{microclimate}_2 \leftarrow \beta_2 \cdot \text{land\_use}$$

**Tier 2 - microclimate x land use drives root**:

$$\text{Root} \leftarrow \beta_3 \cdot \text{mc}_1 + \beta_4 \cdot \text{mc}_2 + \beta_5 \cdot \text{lu} + \beta_6 \cdot (\text{mc}_1 \times \text{lu}) + \beta_7 \cdot (\text{mc}_2 \times \text{lu})$$

**Tier 3 - (microclimate + root) x land use drives fauna**:

$$\text{Fauna} \leftarrow \beta_8 \cdot \text{mc}_1 + \beta_9 \cdot \text{mc}_2 + \beta_{10} \cdot \text{Root} + \beta_{11} \cdot \text{lu} + \beta_{12} \cdot (\text{mc}_1 \times \text{lu}) + \beta_{13} \cdot (\text{mc}_2 \times \text{lu}) + \beta_{14} \cdot (\text{Root} \times \text{lu})$$

*where mc = microclimate index, lu = land_use.*

All equations use a **nested random intercept** (`orientation / depth`) and an **AR(1) correlation structure** to account for temporal autocorrelation within scanner groups. The residual covariance between the two microclimate indices is modelled with a `%~~%` term.

## Sensitivity analysis rationale

Three structural choices must be validated before running the full SEM:

| Parameter | Question | Values tested |
|----|----|----|
| **Window duration** | How many days are needed for a stable fit? | 5, 7, 10, 14, 21, 28, 30 days |
| **Min. data completeness** | How many missing images can be tolerated? | 10% to 50% |
| **Transformation** | Which transform best normalises the variables? | none, Freeman-Tukey, log, Hellinger |

These parameters are crossed in a **full factorial grid** (7 x 5 x 4 = 140 combinations). The same transformation is applied uniformly to all modelled variables (fauna, root, and both microclimate indices) within each configuration. This simplifies the parameter space at this diagnostic stage; variable-specific transforms can be investigated later if residual diagnostics remain unsatisfactory.

| Diagnostic | Metric | Threshold |
|----|----|----|
| Heteroscedasticity | Spearman rho between abs(residuals) and fitted values | \< 0.4 |
| Skewness | Absolute skewness of normalised residuals | \< 1.5 |
| Collinearity | Maximum VIF across main-effect predictors | \< 5 |
| Explanatory power | Marginal and conditional R-squared | -- |

> **Note on VIF with interaction models.** VIF is computed on a **main-effects-only** version of each formula, stripping interaction terms. Collinearity between X and X:land_use is expected by construction and is not informative about data problems; computing VIF on the full interaction formula would always return inflated values.

## Selection criterion

The optimal configuration is chosen by a **two-step procedure**:

1.  For each response variable **independently**, select the transformation maximising `prop_valid_models x avg_R2_marginal`.
2.  Across biological responses **jointly** (root and fauna), find the (window, min_rate) pair maximising `combined_validity x mean_R2`, where `combined_validity = min(prop_valid_root, prop_valid_fauna)`. Using the minimum ensures that neither response can mask a poor fit in the other.

------------------------------------------------------------------------

## 1. Setup

```{r setup}
#| message: false

invisible(Sys.setlocale("LC_TIME", "C"))

libs <- c(
  "tidyverse", "lubridate", "piecewiseSEM", "performance",
  "lme4", "purrr", "nlme", "moments", "car",
  "knitr", "kableExtra"
)
invisible(lapply(libs, require, character.only = TRUE))

# ── Diagnostic thresholds ──────────────────────────────────────────────────────
HETERO_THRESHOLD   <- 0.4
SKEWNESS_THRESHOLD <- 1.5
VIF_THRESHOLD      <- 5.0

# ── Data loading ───────────────────────────────────────────────────────────────
data_sem_raw <- read.csv("../output/SEM_database.csv",
                         sep        = ",",
                         na.strings = c("", "NA")) %>%
  mutate(
    position = as.factor(position),
    depth    = as.factor(depth),
    land_use = as.numeric(position == "C"),
    date     = ymd_hms(date, tz = "UTC"),
    fauna    = all_taxa,
    root     = root_growth_cm2_count_24h
  )

fauna_vars         <- c("all_taxa", readLines("../output/fauna_vars.txt"))
microclimate_index <- readLines("../output/microclimate_index.txt")
```

------------------------------------------------------------------------

## 2. Functions

### 2.1 Rolling window splitter

```{r fn-rolling-windows}
#' Split a time series into overlapping rolling windows
#'
#' @param df          Dataframe with a `date` column (POSIXct).
#' @param window_days Integer. Duration of each window in days.
#' @param step_days   Integer. Step between consecutive window start dates.
#' @return Named list of dataframes, one per window start date.
create_rolling_windows <- function(df, window_days, step_days = 1) {
  start_dates <- seq(
    min(df$date),
    max(df$date) - days(window_days) + hours(6),
    by = paste(step_days, "days")
  )
  windows        <- map(start_dates, ~ filter(df, date >= .x & date < .x + days(window_days)))
  names(windows) <- format(start_dates, "%Y-%m-%d")
  return(windows)
}
```

### 2.2 SEM structure

Tiers 2 and 3 use `* land_use` to expand into main effects plus interactions. Tier 1 is additive. The `%~~%` term accounts for shared physical drivers of both microclimate indices.

```{r fn-fit-psem}
#' Fit the piecewise SEM with interaction terms for a single data window
#'
#' @param data Dataframe with columns: `land_use`, `root`, `fauna`,
#'             `microclimate_1`, `microclimate_2`, `orientation`, `depth`.
#' @return A `psem` object.
fit_psem <- function(data) {
  psem(

    lme(microclimate_1 ~ land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    lme(microclimate_2 ~ land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    lme(root ~ (microclimate_1 + microclimate_2) * land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    lme(fauna ~ (microclimate_1 + microclimate_2 + root) * land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    microclimate_1 %~~% microclimate_2
  )
}
```

### 2.3 Transformation functions

```{r fn-transforms}
#' Freeman-Tukey: sqrt(x) + sqrt(x + 1). Stabilises variance in count data with zeros.
transform_ft <- function(x) sqrt(x) + sqrt(x + 1)

#' Log: log(x + 1). Right-skew correction; log1p handles zero values.
transform_log <- function(x) log1p(x)

#' Hellinger: sqrt(x / sum(x)). Reduces outlier influence; window-relative scale.
transform_hellinger <- function(x) sqrt(x / sum(x, na.rm = TRUE))
```

### 2.4 Main SEM pipeline

```{r fn-pipeline}
#' Run the rolling-window SEM pipeline for one taxon variable
#'
#' @param taxon_var   Character. Column name of the fauna variable.
#' @param data_raw    Dataframe. Full SEM database (already transformed).
#' @param m_indices   Character vector of length 2. Microclimate index names.
#' @param window_days Integer. Rolling window duration (days).
#' @param step_days   Integer. Step between window start dates.
#' @param min_rate    Numeric [0, 1]. Minimum completeness ratio.
#'                    Denominator = window_days x 16 slots x 2 positions.
#' @param n_windows   Integer or NULL. Random subsample size for benchmarking.
#' @return Dataframe with one row per model x response, or NULL.
run_sem_pipeline <- function(taxon_var, data_raw, m_indices,
                             window_days = 14, step_days = 1,
                             min_rate = 0.3, n_windows = NULL) {

  sem_data <- data_raw %>%
    mutate(
      fauna          = !!sym(taxon_var),
      microclimate_1 = !!sym(m_indices[1]),
      microclimate_2 = !!sym(m_indices[2]),
      depth          = as.factor(depth)
    ) %>%
    select(date, land_use, orientation, depth,
           fauna, root, microclimate_1, microclimate_2)

  windows <- create_rolling_windows(sem_data, window_days, step_days)

  # Both positions are present in every window, so denominator is x2
  valid_windows <- keep(windows,
                        ~ (nrow(drop_na(.x)) / (window_days * 16 * 2)) >= min_rate)

  if (!is.null(n_windows) && length(valid_windows) > n_windows) {
    valid_windows <- valid_windows[sample(seq_along(valid_windows), n_windows)]
  }

  results_list <- map(names(valid_windows), function(w_name) {

    win_data <- valid_windows[[w_name]] %>%
      drop_na() %>%
      arrange(land_use, orientation, depth, date)

    vars_to_check <- c("fauna", "root", "microclimate_1", "microclimate_2")

    # Guard 1: zero variance causes Cholesky failures
    if (any(sapply(win_data[vars_to_check], sd, na.rm = TRUE) == 0)) return(NULL)

    # Guard 2: interactions require both land_use levels
    if (length(unique(win_data$land_use)) < 2) return(NULL)

    # Guard 3: non-finite values from transforms or scaling
    if (any(!is.finite(as.matrix(win_data[vars_to_check])))) return(NULL)

    model <- tryCatch(
      withCallingHandlers(
        fit_psem(win_data),
        warning = function(w) invokeRestart("muffleWarning")
      ),
      error = function(e) NULL
    )
    if (is.null(model)) return(NULL)

    sm <- tryCatch(summary(model), error = function(e) NULL)
    if (is.null(sm)) return(NULL)

    # ── R-squared ──────────────────────────────────────────────────────────
    r2_raw   <- as.data.frame(sm$R2)
    col_cond <- grep("R.squared|Conditional", names(r2_raw), value = TRUE)[1]
    col_marg <- grep("Marginal",              names(r2_raw), value = TRUE)[1]

    r2_df <- r2_raw %>%
      mutate(r2_cond     = .[[col_cond]],
             r2_marginal = .[[col_marg]]) %>%
      select(response = Response, r2_cond, r2_marginal)

    # ── Residual diagnostics ───────────────────────────────────────────────
    lme_mods <- keep(model, inherits, "lme")

    map_dfr(lme_mods, function(m) {
      res <- residuals(m, type = "normalized")
      fit <- fitted(m)

      phi_val <- tryCatch(
        coef(m$modelStruct$corStruct, unconstrained = FALSE),
        error = function(e) NA
      )

      # VIF on main-effects-only formula (strip interaction terms and land_use)
      # Interaction terms inflate VIF by construction — not a data problem
      vif_val <- tryCatch({
        all_terms  <- labels(terms(m))
        main_terms <- all_terms[!str_detect(all_terms, ":")]
        main_terms <- setdiff(main_terms, "land_use")
        if (length(main_terms) > 1) {
          f_main <- as.formula(
            paste(all.vars(formula(m))[1], "~",
                  paste(main_terms, collapse = " + "))
          )
          max(car::vif(lm(f_main, data = win_data)))
        } else 1
      }, error = function(e) NA)

      tibble(
        model_id        = paste(taxon_var, w_name, sep = "_"),
        response        = all.vars(formula(m))[1],
        n_obs           = length(res),
        vif_max         = vif_val,
        hetero_strength = cor(abs(res), fit, method = "spearman",
                              use = "complete.obs"),
        skewness        = moments::skewness(res, na.rm = TRUE),
        ar1_phi         = as.numeric(phi_val)
      ) %>%
        left_join(r2_df, by = "response")
    })
  }) %>%
    compact()

  return(map_dfr(results_list, ~ .x))
}
```

------------------------------------------------------------------------

## 3. Benchmarking

### 3.1 Configuration tester

```{r fn-test-config}
#' Test a single parameter configuration and summarise diagnostic outcomes
#'
#' The same transformation is applied uniformly to all modelled variables
#' (fauna, root, microclimate_1, microclimate_2). This keeps the sensitivity
#' grid simple at this stage.
#'
#' @param w_days       Integer. Window duration (days).
#' @param m_rate_val   Numeric. Minimum data completeness ratio.
#' @param transfo_type Character. One of: "none", "sqrt","sqrt_plus_one", "log", "hellinger".
#' @param data_raw     Full SEM dataframe.
#' @param taxon        Character. Fauna variable column name.
#' @param m_indices    Character vector of length 2. Microclimate index names.
#' @return Dataframe with per-response summary statistics, or NULL.
test_configuration <- function(w_days, m_rate_val, transfo_type,
                               data_raw, taxon, m_indices) {

  df_iter <- data_raw

  # ── A. Apply transformation uniformly to all modelled variables ───────────
  all_target_vars <- c(fauna_vars, "root", m_indices)

  df_iter <- df_iter %>%
    mutate(across(all_of(intersect(all_target_vars, names(.))), function(x) {
      switch(transfo_type,
        "sqrt"          = sqrt(x),
        "sqrt_plus_one" = transform_ft(x),
        "log"           = transform_log(x),
        "hellinger"     = transform_hellinger(x),
        x
      )
    }))

  # ── B. Z-score standardisation — all numeric variables except land_use ────
  df_iter <- df_iter %>%
    mutate(across(
      where(is.numeric) & !all_of("land_use"),
      ~ as.numeric(scale(.))
    ))

  # ── C. Run pipeline on 50 randomly sampled windows ───────────────────────
  res_diag <- run_sem_pipeline(
    taxon, df_iter, m_indices,
    window_days = w_days,
    min_rate    = m_rate_val,
    n_windows   = 50
  )

  if (is.null(res_diag) || nrow(res_diag) == 0) return(NULL)

  # ── D. Score and aggregate ────────────────────────────────────────────────
  # NA vif_max = not applicable (single predictor or interaction model) -> pass
  res_diag %>%
    mutate(
      is_ok = abs(hetero_strength) < HETERO_THRESHOLD &
              abs(skewness)        < SKEWNESS_THRESHOLD &
              (is.na(vif_max) | vif_max < VIF_THRESHOLD)
    ) %>%
    group_by(response) %>%
    summarise(
      window             = w_days,
      min_rate           = m_rate_val,
      transformation     = transfo_type,
      hetero_strength    = mean(hetero_strength,       na.rm = TRUE),
      skewness           = mean(skewness,              na.rm = TRUE),
      vif_max            = mean(vif_max,               na.rm = TRUE),
      prop_valid_models  = mean(is_ok,                 na.rm = TRUE),
      avg_r2_marginal    = mean(r2_marginal,           na.rm = TRUE),
      avg_r2_conditional = mean(r2_cond,               na.rm = TRUE),
      delta_r2           = mean(r2_cond - r2_marginal, na.rm = TRUE),
      .groups = "drop"
    )
}
```

### 3.2 Sensitivity grid and execution

```{r benchmarking, cache=TRUE}
sensibility_grid <- expand.grid(
  w_days     = c(7, 10, 14, 21, 28, 30),
  m_rate_val = c(0.1, 0.2, 0.3, 0.4),
  transfo    = c("none", "sqrt", "sqrt_plus_one", "log", "hellinger"),
  stringsAsFactors = FALSE
)

quiet_test <- purrr::quietly(test_configuration)

benchmark_results <- sensibility_grid %>%
  purrr::pmap_dfr(~ quiet_test(
    ..1, ..2, ..3,
    data_sem_raw, fauna_vars[1], microclimate_index
  )$result)
```

------------------------------------------------------------------------

### 4. Results

### 4.1 Global optimisation — independent transformation per response variable

The benchmark grid tested one transformation uniformly per run, but because diagnostics are extracted per-equation, each response can be evaluated independently. We select the best transformation for each of the four responses separately using a lexicographic criterion — validity first, R² second — then find the common (window, min_rate) that maximises joint validity across all responses by the same logic.

```{r global-optimization}
# ── Step 1: per (window, min_rate, response), best transformation ─────────────
# Lexicographic selection: maximise prop_valid_models first, break ties by R².
# This ensures scientific rigour is never traded for explanatory power.
best_per_variable <- benchmark_results %>%
  group_by(window, min_rate, response) %>%
  arrange(desc(prop_valid_models), desc(avg_r2_marginal)) %>%
  slice(1) %>%
  ungroup()

# ── Step 2: best common (window, min_rate) across all four responses ──────────
# combined_validity = min across all four responses: a poor fit in any single
# response cannot be masked by good fits in the others.
# Lexicographic sort applied again at this level: validity first, R² second.
global_optimization <- best_per_variable %>%
  group_by(window, min_rate) %>%
  summarise(
    combined_validity      = min(prop_valid_models),
    mean_r2                = mean(avg_r2_marginal),
    transfo_root           = transformation[response == "root"],
    transfo_fauna          = transformation[response == "fauna"],
    transfo_microclimate_1 = transformation[response == "microclimate_1"],
    transfo_microclimate_2 = transformation[response == "microclimate_2"],
    .groups = "drop"
  ) %>%
  arrange(desc(combined_validity), desc(mean_r2)) %>%
  # global_score retained for reporting only — selection is lexicographic
  mutate(global_score = combined_validity * mean_r2)

# ── Step 3: display top 5 ─────────────────────────────────────────────────────
global_optimization %>%
  head(5) %>%
  mutate(across(c(combined_validity, mean_r2, global_score), ~ round(., 3))) %>%
  select(
    "Window (days)"          = window,
    "Max missing data"       = min_rate,
    "Transfo root"           = transfo_root,
    "Transfo fauna"          = transfo_fauna,
    "Transfo microclimate 1" = transfo_microclimate_1,
    "Transfo microclimate 2" = transfo_microclimate_2,
    "Validity (joint)"       = combined_validity,
    "R2 Marginal (mean)"     = mean_r2,
    "Score"                  = global_score
  ) %>%
  kable(caption = "Top 5 configurations — ranked by validity then R² (lexicographic)") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE, position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  row_spec(1, bold = TRUE, background = "#FEC288")
```

------------------------------------------------------------------------

## 5. Export

### 5.1 Full benchmark results

```{r results-writing}
write.csv(benchmark_results,
          "../output/benchmark_results_database.csv",
          row.names = FALSE,
          quote     = TRUE)
```

### 5.2 Selected model parameters

Four separate transform keys are exported, one per response variable. `check_models_performance.qmd` reads each key and applies the corresponding transformation before z-scoring. The selected configuration is row 1 of `global_optimization`, which is already sorted lexicographically (highest validity, then highest R²).

```{r export-parameters}
# Row 1 is the winner: highest combined validity, then highest R² among ties
final_selection <- global_optimization %>%
  slice(1)

params_to_save <- list(
  selected_window          = final_selection$window,
  selected_min_rate        = final_selection$min_rate,
  transform_root           = final_selection$transfo_root,
  transform_fauna          = final_selection$transfo_fauna,
  transform_microclimate_1 = final_selection$transfo_microclimate_1,
  transform_microclimate_2 = final_selection$transfo_microclimate_2
)

jsonlite::write_json(params_to_save,
                     "../output/model_parameters.txt",
                     pretty = TRUE)

message(sprintf(
  paste0("Best config: window = %d days | min_rate = %.1f\n",
         "  validity = %.3f | R2 = %.3f | score = %.3f\n",
         "  root: %s | fauna: %s | mc1: %s | mc2: %s"),
  final_selection$window,
  final_selection$min_rate,
  final_selection$combined_validity,
  final_selection$mean_r2,
  final_selection$global_score,
  final_selection$transfo_root,
  final_selection$transfo_fauna,
  final_selection$transfo_microclimate_1,
  final_selection$transfo_microclimate_2
))
```
