---
title: |
   SEM model execution and output database generation
author: "Emma Belaud"
format:
  html:
    css: css/styles.css
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
    number-sections: true
execute:
  warning: false
  message: false
  echo: true
---

::: {.callout-note appearance="simple"}
## Purpose of this document

This script is the **third step** of the SEM pipeline, following:

-   `1_database_edition.qmd` — assembles and quality-controls the SEM database
-   `2_SEM_diagnostic.qmd` — selects the optimal rolling-window duration, variance-stabilising transformation, and set of diagnostically valid windows

This script performs **all modelling** and writes the result databases that `4_SEM_results_analysis.qmd` reads for figures and tables. No figures are produced here — keeping computation and presentation strictly separated makes it easy to re-run the visualisation without refitting any model.

| Output file                | Contents                                       |
|----------------------------|------------------------------------------------|
| `H1_adjusted_values.csv`   | Per-observation RE-adjusted faunal counts      |
| `H1_model_summary.csv`     | Fixed-effect estimate, Hedge's g, p-value      |
| `DAG_coefficients.csv`     | Full-dataset psem path coefficients (all_taxa) |
| `DAG_model_fit.csv`        | Fisher's C, AIC for the full-dataset psem      |
| `SEM_results_database.csv` | Rolling-window path coefficients, all taxa     |
| `indirect_effects.csv`     | Computed indirect effects per window × taxon   |
| `H1_results_database.csv`  | Rolling-window H1 estimates, all responses     |
:::

------------------------------------------------------------------------

# SEM structure

The four structural equations encode the hypothesised causal hierarchy from land-use type through soil microclimate and root growth to faunal activity:

$$\text{microclimate}_i \leftarrow \beta_i \cdot \text{land\_use}, \quad i=1,2$$

$$\text{Root} \leftarrow \beta_3 \cdot \text{mc}_1 + \beta_4 \cdot \text{mc}_2 + \beta_5 \cdot \text{lu}$$

$$\text{Fauna} \leftarrow \beta_6 \cdot \text{mc}_1 + \beta_7 \cdot \text{mc}_2 + \beta_8 \cdot \text{Root} + \beta_9 \cdot \text{lu}$$

The residual covariance `microclimate_1 %~~% microclimate_2` is included explicitly in the psem to acknowledge that the two microclimate indices share physical drivers beyond land use (e.g. weather events, aspect). Omitting it would force any shared variance into the land_use path coefficients.

## Random structures

| Equation | Random structure | Rationale |
|----|----|----|
| microclimate \~ land_use | `~ 1 \| depth` | No orientation recorded for sensors |
| root / fauna \~ … | `~ 1 \| orientation/depth` | Depth nested in orientation |

The biological equations fall back to `~ 1 | depth` if the nested structure does not converge (e.g. too few levels per group in short windows).

------------------------------------------------------------------------

# Setup

```{r setup}
invisible(Sys.setlocale("LC_TIME", "C"))

libs <- c(
  "tidyverse", "lubridate", "piecewiseSEM",
  "purrr", "nlme", "jsonlite",
  "knitr", "kableExtra"
)
invisible(lapply(libs, require, character.only = TRUE))

season_lookup <- tibble(
  month  = 1:12,
  season = c("Winter", "Winter",
             "Spring", "Spring", "Spring",
             "Summer", "Summer", "Summer",
             "Autumn", "Autumn", "Autumn",
             "Winter")
)
```

------------------------------------------------------------------------

# Parameter loading

Parameters are read from the JSON file written by S2 (`2_SEM_diagnostic.qmd`). Hardcoded fallback values are used if the file is absent.

```{r load-params}
if (file.exists("../output/model_parameters.txt")) {
  best_params  <- jsonlite::fromJSON("../output/model_parameters.txt")
  WIN_DAYS               <- as.integer(best_params$selected_window)
  MIN_RATE               <- as.numeric(best_params$selected_min_rate)
  TRANSFO_ROOT           <- best_params$transform_root
  TRANSFO_FAUNA          <- best_params$transform_fauna
  TRANSFO_MICROCLIMATE_1 <- best_params$transform_microclimate_1
  TRANSFO_MICROCLIMATE_2 <- best_params$transform_microclimate_2
} else {
  WIN_DAYS               <- 14
  MIN_RATE               <- 0.3
  TRANSFO_ROOT           <- "log"
  TRANSFO_FAUNA          <- "log"
  TRANSFO_MICROCLIMATE_1 <- "none"
  TRANSFO_MICROCLIMATE_2 <- "none"
  warning("model_parameters.txt not found — using hardcoded fallback values.")
}

message(sprintf(
  "Config | window: %d days | min_rate: %.2f | root: %s | fauna: %s | mc1: %s | mc2: %s",
  WIN_DAYS, MIN_RATE,
  TRANSFO_ROOT, TRANSFO_FAUNA,
  TRANSFO_MICROCLIMATE_1, TRANSFO_MICROCLIMATE_2
))
```

------------------------------------------------------------------------

# Data loading and preprocessing

```{r data-preparation}
data_sem_raw <- read.csv("../output/SEM_database.csv",
                         sep = ",", na.strings = c("", "NA")) %>%
  mutate(
    position    = as.factor(position),
    depth       = as.factor(depth),
    orientation = as.factor(orientation),
    land_use    = as.numeric(position == "C"),
    date        = ymd_hms(date, tz = "UTC"),
    root        = root_cm2_count
  )

fauna_vars         <- c("all_taxa", readLines("../output/fauna_vars.txt"))
microclimate_index <- readLines("../output/microclimate_index.txt")

# Windows pre-validated by S2 (2_SEM_diagnostic.qmd).
# NULL = no filter (all completeness-passing windows are used).
selected_windows <- if (file.exists("../output/selected_windows.txt")) {
  as.Date(readLines("../output/selected_windows.txt"))
} else {
  NULL
}

message(sprintf(
  "Data loaded: %d rows | %d taxa | %d orientations | %d depth levels | %s valid windows",
  nrow(data_sem_raw),
  length(fauna_vars),
  n_distinct(data_sem_raw$orientation),
  n_distinct(data_sem_raw$depth),
  if (is.null(selected_windows)) "no filter" else as.character(length(selected_windows))
))
```

```{r apply-transformations}
# ── Apply variance-stabilising transformations ────────────────────────────────
# Applied ONCE on the full dataset before any z-scoring.
# Z-scoring is done within each rolling window independently.
# Exception: Hellinger is window-relative and is handled inside run_sem_pipeline().

shift_pos <- function(x, label = "") {
  x_min <- min(x, na.rm = TRUE)
  if (!is.na(x_min) && x_min <= 0) {
    shift <- abs(x_min) + 0.001
    message(sprintf("  [shift] '%s' shifted by +%.4f", label, shift))
    x <- x + shift
  }
  x
}

apply_transfo <- function(x, method, label = "") {
  switch(method,
    sqrt_plus_one = sqrt(x) + sqrt(x + 1),
    sqrt          = sqrt(shift_pos(x, label)),
    log           = log1p(shift_pos(x, label)),
    hellinger     = x,   # handled per-window
    x
  )
}

data_transformed <- data_sem_raw %>%
  mutate(root = apply_transfo(root, TRANSFO_ROOT, "root")) %>%
  mutate(across(all_of(intersect(fauna_vars, names(.))),
                ~ apply_transfo(., TRANSFO_FAUNA, cur_column()))) %>%
  mutate(across(all_of(microclimate_index[1]),
                ~ apply_transfo(., TRANSFO_MICROCLIMATE_1, microclimate_index[1]))) %>%
  mutate(across(all_of(microclimate_index[2]),
                ~ apply_transfo(., TRANSFO_MICROCLIMATE_2, microclimate_index[2])))
```

------------------------------------------------------------------------

# Shared helper functions

```{r helpers}
# ── LME controls ──────────────────────────────────────────────────────────────
lme_ctrl <- nlme::lmeControl(msMaxIter = 200, opt = "optim")

# ── R² for nlme::lme (variance-based) ────────────────────────────────────────
# R²m (marginal)    = var(fixed predictions) / var(observed)
# R²c (conditional) = var(full predictions)  / var(observed)
r2_lme <- function(m) {
  if (is.null(m)) return(c(R2m = NA_real_, R2c = NA_real_))
  tryCatch({
    y   <- as.numeric(nlme::getResponse(m))
    ss  <- var(y, na.rm = TRUE)
    if (is.na(ss) || ss == 0) return(c(R2m = NA_real_, R2c = NA_real_))
    r2m <- var(as.numeric(fitted(m, level = 0)),                na.rm = TRUE) / ss
    r2c <- var(as.numeric(fitted(m, level = length(m$groups))), na.rm = TRUE) / ss
    c(R2m = min(max(r2m, 0), 1), R2c = min(max(r2c, 0), 1))
  }, error = \(e) c(R2m = NA_real_, R2c = NA_real_))
}

# ── Rolling window splitter ───────────────────────────────────────────────────
create_rolling_windows <- function(df, window_days, step_days = 1) {
  starts  <- seq(min(df$date),
                 max(df$date) - days(window_days) + hours(6),
                 by = paste(step_days, "days"))
  windows <- map(starts, ~ filter(df, date >= .x & date < .x + days(window_days)))
  setNames(windows, format(starts, "%Y-%m-%d"))
}

# ── Safe LME fitter for rolling-window models ─────────────────────────────────
# NOT passed to psem() — safe to use for rolling-window loop and H1 models.
fit_lme_safe <- function(formula, data, rand_form) {
  tryCatch(
    withCallingHandlers(
      nlme::lme(formula, data = data,
                random      = rand_form,
                correlation = nlme::corAR1(form = rand_form),
                control     = lme_ctrl),
      warning = \(w) invokeRestart("muffleWarning")
    ),
    error = \(e) NULL
  )
}
```

------------------------------------------------------------------------

# Two-step model: land-use effect on faunal activity (H1)

The approach is a two-step sequential modelling procedure:

**Step 1 — Null model** (`y ~ 1 + (1 | orientation/depth) + AR(1)`): fitted with no land_use predictor. Residuals represent faunal activity after removing variation attributable to scanner orientation, depth, and temporal autocorrelation.

**Step 2 — Land-use test on residuals** (`residual_step1 ~ land_use`): a simple OLS testing whether Step 1 residuals differ between treed (A) and cultivated (C). β = mean(residuals_C) − mean(residuals_A). Effect size is reported as Hedge's g ≈ β / σ_pooled.

```{r h1-global-model}
h1_model_data <- data_sem_raw %>%
  select(date, position, orientation, depth, land_use, all_taxa) %>%
  drop_na() %>%
  mutate(
    y_trans        = apply_transfo(all_taxa, TRANSFO_FAUNA, "all_taxa"),
    position_label = factor(
      if_else(land_use == 0, "Treed (A)", "Cultivated (C)"),
      levels = c("Treed (A)", "Cultivated (C)")
    ),
    month  = month(as.Date(date)),
    season = factor(season_lookup$season[month],
                    levels = c("Spring", "Summer", "Autumn", "Winter"))
  )

# Step 1: null model (intercept only + random effects)
m_null <- suppressWarnings(
  nlme::lme(
    y_trans ~ 1,
    data        = h1_model_data,
    random      = ~ 1 | orientation / depth,
    correlation = nlme::corAR1(form = ~ 1 | orientation / depth),
    control     = lme_ctrl
  )
)

h1_resid <- residuals(m_null, type = "response")

# Step 2: OLS on residuals
m_h1_step2 <- lm(h1_resid ~ h1_model_data$land_use)
step2_summ  <- summary(m_h1_step2)
beta_lu     <- coef(m_h1_step2)["h1_model_data$land_use"]
se_lu       <- step2_summ$coefficients["h1_model_data$land_use", "Std. Error"]
t_lu        <- step2_summ$coefficients["h1_model_data$land_use", "t value"]
p_lu        <- step2_summ$coefficients["h1_model_data$land_use", "Pr(>|t|)"]

sigma_pooled <- sd(h1_resid, na.rm = TRUE)
hedges_g     <- beta_lu / sigma_pooled
r2_null      <- r2_lme(m_null)

h1_adj <- h1_model_data %>%
  mutate(adj_value = as.numeric(h1_resid), transformation = TRANSFO_FAUNA)
write.csv(h1_adj, "../output/H1_adjusted_values.csv", row.names = FALSE)

h1_summary <- tibble(
  beta_land_use       = beta_lu,
  se_land_use         = se_lu,
  t_value             = t_lu,
  p_value             = p_lu,
  hedges_g            = hedges_g,
  r2_null_marginal    = as.numeric(r2_null["R2m"]),
  r2_null_conditional = as.numeric(r2_null["R2c"]),
  r2_step2            = step2_summ$r.squared,
  n_obs               = nrow(h1_model_data),
  transformation      = TRANSFO_FAUNA
)
write.csv(h1_summary, "../output/H1_model_summary.csv", row.names = FALSE)

message(sprintf(
  "H1 | β = %.3f | g = %.2f | p %s | R²(lu) = %.3f | R²c(null) = %.3f",
  beta_lu, hedges_g,
  if (p_lu < 0.001) "< 0.001" else sprintf("= %.3f", p_lu),
  step2_summ$r.squared, as.numeric(r2_null["R2c"])
))
```

------------------------------------------------------------------------

# Full-dataset piecewise SEM

```{r h2a-full-psem}
library(piecewiseSEM)

# Z-scored globally (full time-series SEM, not rolling window)
full_data <- data_transformed %>%
  select(date, land_use, orientation, depth, all_taxa,
         root, all_of(microclimate_index)) %>%
  rename(microclimate_1 = microclimate_index[1],
         microclimate_2 = microclimate_index[2],
         fauna          = all_taxa) %>%
  drop_na() %>%
  mutate(across(c(fauna, root, microclimate_1, microclimate_2),
                ~ as.numeric(scale(.))))

# ── Fit four LMEs with literal random formulas ────────────────────────────────
lme_mc1 <- suppressWarnings(tryCatch(
  nlme::lme(microclimate_1 ~ land_use,
            data = full_data, random = ~ 1 | depth,
            correlation = nlme::corAR1(form = ~ 1 | depth), control = lme_ctrl),
  error = \(e) NULL
))

lme_mc2 <- suppressWarnings(tryCatch(
  nlme::lme(microclimate_2 ~ land_use,
            data = full_data, random = ~ 1 | depth,
            correlation = nlme::corAR1(form = ~ 1 | depth), control = lme_ctrl),
  error = \(e) NULL
))

lme_root <- suppressWarnings(tryCatch(
  nlme::lme(root ~ microclimate_1 + microclimate_2 + land_use,
            data = full_data, random = ~ 1 | orientation / depth,
            correlation = nlme::corAR1(form = ~ 1 | orientation / depth),
            control = lme_ctrl),
  error = \(e) suppressWarnings(tryCatch(
    nlme::lme(root ~ microclimate_1 + microclimate_2 + land_use,
              data = full_data, random = ~ 1 | depth,
              correlation = nlme::corAR1(form = ~ 1 | depth), control = lme_ctrl),
    error = \(e2) NULL
  ))
))

lme_fauna <- suppressWarnings(tryCatch(
  nlme::lme(fauna ~ microclimate_1 + microclimate_2 + root + land_use,
            data = full_data, random = ~ 1 | orientation / depth,
            correlation = nlme::corAR1(form = ~ 1 | orientation / depth),
            control = lme_ctrl),
  error = \(e) suppressWarnings(tryCatch(
    nlme::lme(fauna ~ microclimate_1 + microclimate_2 + root + land_use,
              data = full_data, random = ~ 1 | depth,
              correlation = nlme::corAR1(form = ~ 1 | depth), control = lme_ctrl),
    error = \(e2) NULL
  ))
))

# ── Assemble and summarise psem ───────────────────────────────────────────────
full_psem   <- suppressWarnings(
  psem(lme_mc1, lme_mc2, lme_root, lme_fauna,
       microclimate_1 %~~% microclimate_2, data = full_data)
)
sem_summary <- summary(full_psem)

# ── Model fit ─────────────────────────────────────────────────────────────────
dag_fit <- tibble(
  fishers_C = sem_summary$Cstat$Fisher.C,
  df        = sem_summary$Cstat$df,
  p_value   = sem_summary$Cstat$P.Value,
  AIC       = sem_summary$AIC$AIC,
  AICc      = sem_summary$AIC$AICc
)
write.csv(dag_fit, "../output/DAG_model_fit.csv", row.names = FALSE)

# ── Path coefficients ─────────────────────────────────────────────────────────
r2_cols     <- names(sem_summary$R2)
r2_marg_col <- r2_cols[grep("Marginal|R.squared$",   r2_cols)[1]]
r2_cond_col <- r2_cols[grep("Conditional|R.squared.1", r2_cols)[1]]

dag_table <- as.data.frame(sem_summary$coefficients) %>%
  setNames(make.names(names(.), unique = TRUE)) %>%
  filter(!is.na(Response), Response != "", Response != "~~") %>%
  transmute(
    equation     = paste0(Response, " ~ ", Predictor),
    response     = Response,
    predictor    = Predictor,
    estimate     = Estimate,
    std_estimate = Std.Estimate,
    se           = Std.Error,
    t_value      = if ("Crit.Value" %in% names(.)) Crit.Value else NA_real_,
    p_value      = suppressWarnings(as.numeric(P.Value)),
    sig          = case_when(
      p_value < 0.001 ~ "***", p_value < 0.01 ~ "**",
      p_value < 0.05  ~ "*",   p_value < 0.10 ~ ".", TRUE ~ "")
  ) %>%
  left_join(
    as.data.frame(sem_summary$R2) %>%
      transmute(response       = Response,
                r2_marginal    = .[[r2_marg_col]],
                r2_conditional = .[[r2_cond_col]]),
    by = "response"
  )

write.csv(dag_table, "../output/DAG_coefficients.csv", row.names = FALSE)

message(sprintf(
  "Full psem — Fisher's C = %.2f (df = %d, p = %.3f) | AIC = %.1f",
  dag_fit$fishers_C, dag_fit$df, dag_fit$p_value, dag_fit$AIC
))
```

------------------------------------------------------------------------

# Rolling-window SEMs

## piecewise SEM structure per window

```{r sem-structure}
#' Fit the piecewise SEM for one rolling window.
#'
#' Uses literal formula objects (not variables) so psem::summary() can
#' re-evaluate $call for d-separation tests. Falls back to depth-only random
#' structure if the nested structure fails.
fit_psem_window <- function(data) {
  lc   <- nlme::lmeControl(msMaxIter = 200, opt = "optim")
  safe <- function(expr) {
    tryCatch(
      withCallingHandlers(expr, warning = \(w) invokeRestart("muffleWarning")),
      error = \(e) NULL
    )
  }

  m_mc1 <- safe(nlme::lme(microclimate_1 ~ land_use, data = data,
                           random = ~ 1 | depth,
                           correlation = nlme::corAR1(form = ~ 1 | depth), control = lc))
  m_mc2 <- safe(nlme::lme(microclimate_2 ~ land_use, data = data,
                           random = ~ 1 | depth,
                           correlation = nlme::corAR1(form = ~ 1 | depth), control = lc))

  m_root <- safe(nlme::lme(root ~ microclimate_1 + microclimate_2 + land_use, data = data,
                            random = ~ 1 | orientation / depth,
                            correlation = nlme::corAR1(form = ~ 1 | orientation / depth),
                            control = lc))
  if (is.null(m_root))
    m_root <- safe(nlme::lme(root ~ microclimate_1 + microclimate_2 + land_use, data = data,
                              random = ~ 1 | depth,
                              correlation = nlme::corAR1(form = ~ 1 | depth), control = lc))

  m_fauna <- safe(nlme::lme(fauna ~ microclimate_1 + microclimate_2 + root + land_use,
                             data = data,
                             random = ~ 1 | orientation / depth,
                             correlation = nlme::corAR1(form = ~ 1 | orientation / depth),
                             control = lc))
  if (is.null(m_fauna))
    m_fauna <- safe(nlme::lme(fauna ~ microclimate_1 + microclimate_2 + root + land_use,
                               data = data,
                               random = ~ 1 | depth,
                               correlation = nlme::corAR1(form = ~ 1 | depth), control = lc))

  if (any(sapply(list(m_mc1, m_mc2, m_root, m_fauna), is.null))) return(NULL)
  safe(psem(m_mc1, m_mc2, m_root, m_fauna,
            microclimate_1 %~~% microclimate_2, data = data))
}
```

## Rolling-window pipeline

```{r sem-pipeline}
#' Rolling-window SEM pipeline for one taxon.
#'
#' Per-window steps: (1) subset + drop NAs, (2) apply Hellinger if selected,
#' (3) z-score all modelled variables, (4) guard checks, (5) fit psem,
#' (6) extract coefficients and R².
run_sem_pipeline <- function(taxon_var, data_raw, m_indices,
                             window_days, min_rate) {
  cat("\n--- Taxon:", taxon_var, "---\n")
  vars_to_scale <- c("fauna", "root", "microclimate_1", "microclimate_2")

  sem_data <- data_raw %>%
    mutate(fauna          = !!sym(taxon_var),
           microclimate_1 = !!sym(m_indices[1]),
           microclimate_2 = !!sym(m_indices[2])) %>%
    select(date, land_use, orientation, depth,
           fauna, root, microclimate_1, microclimate_2)

  windows    <- create_rolling_windows(sem_data, window_days)
  valid_wins <- keep(windows, function(w) {
    n <- nrow(w); nc <- nrow(drop_na(w))
    n > 0 && (nc / n) >= min_rate
  })

  if (!is.null(selected_windows))
    valid_wins <- valid_wins[as.Date(names(valid_wins)) %in% selected_windows]

  cat(sprintf("  %d / %d windows retained.\n", length(valid_wins), length(windows)))
  if (length(valid_wins) == 0) return(NULL)

  results <- suppressWarnings(map(names(valid_wins), function(w_name) {
    win_data <- valid_wins[[w_name]] %>%
      drop_na() %>%
      arrange(land_use, orientation, depth, date)

    if (TRANSFO_FAUNA == "hellinger") {
      if (mean(win_data$fauna == 0, na.rm = TRUE) > 0.8) return(NULL)
      win_data <- mutate(win_data, fauna = sqrt(fauna / sum(fauna, na.rm = TRUE)))
    }

    win_data <- win_data %>%
      mutate(across(all_of(vars_to_scale), ~ as.numeric(scale(.))))

    stdevs <- sapply(win_data[vars_to_scale], sd, na.rm = TRUE)
    if (any(is.na(stdevs) | stdevs == 0))                   return(NULL)
    if (length(unique(win_data$land_use)) < 2)               return(NULL)
    if (any(!is.finite(as.matrix(win_data[vars_to_scale])))) return(NULL)

    model <- tryCatch(fit_psem_window(win_data), error = \(e) NULL)
    if (is.null(model)) return(NULL)
    sm <- tryCatch(summary(model), error = \(e) NULL)
    if (is.null(sm)) return(NULL)

    coefs <- as.data.frame(sm$coefficients) %>%
      setNames(make.names(names(.), unique = TRUE)) %>%
      filter(!is.na(Response), Response != "", Response != "~~")

    r2_df     <- as.data.frame(sm$R2)
    r2_cols_w <- names(r2_df)
    r2m_col   <- r2_cols_w[grep("Marginal|R.squared$",    r2_cols_w)[1]]
    r2c_col   <- r2_cols_w[grep("Conditional|R.squared.1", r2_cols_w)[1]]

    r2_tbl <- r2_df %>%
      transmute(response = Response, r2_marg = .[[r2m_col]], r2_cond = .[[r2c_col]])

    tibble(
      taxon        = taxon_var,
      date_start   = as.Date(w_name),
      n_obs        = sm$n,
      response     = coefs$Response,
      predictor    = coefs$Predictor,
      estimate     = coefs$Estimate,
      std_estimate = coefs$Std.Estimate,
      p_value      = suppressWarnings(as.numeric(gsub("[^0-9.e-]", "", coefs$P.Value)))
    ) %>%
      left_join(r2_tbl, by = "response")
  })) %>% compact()

  if (length(results) == 0) return(NULL)
  map_dfr(results, ~ .x)
}
```

## Execution

```{r sem-execution}
message(">>> Running rolling-window SEMs for all taxa...")

sem_results_raw <- suppressWarnings(
  map(fauna_vars, ~ run_sem_pipeline(
    .x, data_transformed, microclimate_index, WIN_DAYS, MIN_RATE
  )) %>%
    compact() %>%
    map_dfr(~ .x)
)

message(sprintf("SEMs complete: %d coefficient rows across %d taxa.",
                nrow(sem_results_raw), n_distinct(sem_results_raw$taxon)))
```

------------------------------------------------------------------------

# Rolling-window H1 models

Per-window simple LME (`y ~ land_use`) for all response variables. These are total land-use effects estimated in isolation (not SEM partial effects), saved for the rolling-window H1 visualisation in S4.

```{r h1-rolling}
fit_h1_windows <- function(response_var, response_group, data_raw,
                           win_days, min_rate) {
  rand_primary  <- if (response_group == "microclimate") ~ 1 | depth
                   else ~ 1 | orientation / depth
  rand_fallback <- ~ 1 | depth

  df <- data_raw %>%
    mutate(y = !!sym(response_var)) %>%
    select(date, land_use, orientation, depth, y) %>%
    drop_na()

  windows    <- create_rolling_windows(df, win_days)
  valid_wins <- keep(windows, function(w) {
    n <- nrow(w); nc <- nrow(drop_na(w))
    n > 0 && (nc / n) >= min_rate
  })

  if (!is.null(selected_windows))
    valid_wins <- valid_wins[as.Date(names(valid_wins)) %in% selected_windows]
  if (length(valid_wins) == 0) return(NULL)

  map_dfr(names(valid_wins), function(w_name) {
    wd <- valid_wins[[w_name]] %>%
      drop_na() %>%
      arrange(land_use, orientation, depth, date) %>%
      mutate(y = as.numeric(scale(y)))

    if (is.na(sd(wd$y, na.rm = TRUE)) || sd(wd$y, na.rm = TRUE) == 0) return(NULL)
    if (length(unique(wd$land_use)) < 2) return(NULL)

    m <- fit_lme_safe(y ~ land_use, wd, rand_primary)
    if (is.null(m) && response_group != "microclimate")
      m <- fit_lme_safe(y ~ land_use, wd, rand_fallback)
    if (is.null(m)) return(NULL)

    tt  <- summary(m)$tTable
    r2  <- r2_lme(m)
    phi <- tryCatch(
      as.numeric(coef(m$modelStruct$corStruct, unconstrained = FALSE)),
      error = \(e) NA_real_
    )

    tibble(
      response       = response_var,
      response_group = response_group,
      date_start     = as.Date(w_name),
      season         = season_lookup$season[month(as.Date(w_name))],
      n_obs          = nrow(wd),
      beta_land_use  = tt["land_use", "Value"],
      se_land_use    = tt["land_use", "Std.Error"],
      t_value        = tt["land_use", "t-value"],
      p_value        = tt["land_use", "p-value"],
      ar1_phi        = phi,
      r2_marginal    = as.numeric(r2["R2m"]),
      r2_conditional = as.numeric(r2["R2c"]),
      delta_r2       = as.numeric(r2["R2c"]) - as.numeric(r2["R2m"])
    )
  })
}

message(">>> Running H1 rolling-window models...")

h1_response_map <- c(
  setNames(rep("microclimate", 2),            microclimate_index),
  root_growth_cm2_count_24h = "root",
  setNames(rep("fauna", length(fauna_vars)),  fauna_vars)
)

h1_results <- suppressWarnings(
  imap_dfr(h1_response_map,
           ~ fit_h1_windows(.y, .x, data_sem_raw, WIN_DAYS, MIN_RATE))
) %>%
  mutate(season = factor(season,
                         levels = c("Spring", "Summer", "Autumn", "Winter")))

message(sprintf("H1 rolling-window complete: %d rows | %d responses",
                nrow(h1_results), n_distinct(h1_results$response)))
```

------------------------------------------------------------------------

# Indirect effect computation

```{r indirect-effects}
# ── Path-product indirect effects ─────────────────────────────────────────────
# Indirect effects = products of standardised β values along each causal route.

mc1_resp <- "microclimate_1"
mc2_resp <- "microclimate_2"

coefs_wide <- sem_results_raw %>%
  filter(!str_detect(predictor, "~~"), !is.na(std_estimate)) %>%
  mutate(coef_id = paste0(predictor, "__", response)) %>%
  select(taxon, date_start, coef_id, std_estimate) %>%
  pivot_wider(names_from = coef_id, values_from = std_estimate, values_fn = mean)

get_col <- \(df, col) if (col %in% names(df)) df[[col]] else rep(NA_real_, nrow(df))

indirect_effects <- coefs_wide %>%
  mutate(
    b1 = get_col(., paste0("land_use__", mc1_resp)),  # lu → mc1 (β₁)
    b2 = get_col(., paste0("land_use__", mc2_resp)),  # lu → mc2 (β₂)
    b3 = get_col(., paste0(mc1_resp,     "__root")),  # mc1 → root (β₃)
    b4 = get_col(., paste0(mc2_resp,     "__root")),  # mc2 → root (β₄)
    b5 = get_col(., "land_use__root"),                # lu → root (β₅)
    b6 = get_col(., paste0(mc1_resp,     "__fauna")), # mc1 → fauna (β₆)
    b7 = get_col(., paste0(mc2_resp,     "__fauna")), # mc2 → fauna (β₇)
    b8 = get_col(., "root__fauna"),                   # root → fauna (β₈)
    b9 = get_col(., "land_use__fauna"),               # lu → fauna direct (β₉)

    direct         = b9,
    via_mc1        = b1 * b6,
    via_mc2        = b2 * b7,
    via_root       = b5 * b8,
    via_mc1_root   = b1 * b3 * b8,
    via_mc2_root   = b2 * b4 * b8,
    total_indirect = via_mc1 + via_mc2 + via_root + via_mc1_root + via_mc2_root,
    total_effect   = direct + total_indirect
  ) %>%
  select(taxon, date_start,
         direct, via_mc1, via_mc2, via_root,
         via_mc1_root, via_mc2_root, total_indirect, total_effect) %>%
  mutate(
    month  = month(as.Date(date_start)),
    season = case_when(
      month %in% 3:5  ~ "Spring",
      month %in% 6:8  ~ "Summer",
      month %in% 9:11 ~ "Autumn",
      TRUE            ~ "Winter"
    )
  ) %>%
  select(-month)

write.csv(indirect_effects, "../output/indirect_effects.csv", row.names = FALSE)
```

------------------------------------------------------------------------

# Export

```{r export}
# ── Add season and path_id to SEM results ────────────────────────────────────
sem_results <- sem_results_raw %>%
  filter(!str_detect(predictor, "~~")) %>%
  mutate(path_id = paste(predictor, "\u2192", response),
         month   = month(date_start)) %>%
  left_join(season_lookup, by = "month") %>%
  mutate(season = factor(season,
                         levels = c("Spring", "Summer", "Autumn", "Winter")))

write.csv(sem_results, "../output/SEM_results_database.csv", row.names = FALSE)
write.csv(h1_results,  "../output/H1_results_database.csv",  row.names = FALSE)

message(sprintf(
  "Export complete:
  SEM results : %d rows | %d taxa | %d unique paths
  H1 results  : %d rows | %d responses
  Indirect eff: %d rows | %d taxa
  Full psem   : DAG_coefficients.csv + DAG_model_fit.csv
  H1 global   : H1_adjusted_values.csv + H1_model_summary.csv",
  nrow(sem_results),      n_distinct(sem_results$taxon),
  n_distinct(sem_results$path_id),
  nrow(h1_results),       n_distinct(h1_results$response),
  nrow(indirect_effects), n_distinct(indirect_effects$taxon)
))
```

| Output file | Contents |
|----|----|
| `H1_adjusted_values.csv` | Per-observation null-model residuals |
| `H1_model_summary.csv` | β, SE, t, p, Hedge's g, R² (two-step H1 model) |
| `H1_results_database.csv` | Rolling-window H1 estimates (all responses) |
| `DAG_coefficients.csv` | Full-dataset psem path coefficients |
| `DAG_model_fit.csv` | Fisher's C, AIC |
| `SEM_results_database.csv` | Rolling-window path coefficients (all taxa) |
| `indirect_effects.csv` | Rolling-window indirect effect products (all taxa) |
