---
title: "SEM — model execution and path coefficient extraction"
author: "Emma Belaud"
format:
  html:
    css: css/styles.css  
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
execute: 
  warning: false
  message: false
  echo: true
---

# Overview

## Role in the pipeline

This script is the **third step** of the SEM pipeline, following:

1. `data_preparation.qmd` — builds the full SEM database
2. `sem_diagnostic.qmd` — selects the optimal window, completeness threshold, and transformations

Here, those validated parameters are applied to run the **full rolling-window pSEM** across all taxa, and to generate a **null distribution via permutation** for inferential comparison.

## What this script produces

| Output | File | Description |
|--------|------|-------------|
| Observed path coefficients | `SEM_results_database.csv` | Standardised estimates, p-values, R², season labels for all taxon × window combinations |
| Null distribution summary | *(in memory)* | Mean and SD of path coefficients under random permutation of all causal variables |

## Statistical approach

The causal model has four structural equations organised in three tiers, with `land_use` (0 = unmanaged, 1 = managed) as the exogenous driver:

$$\text{Tier 1} \quad \text{microclimate}_1 \leftarrow \beta_1 \cdot \text{land\_use}$$
$$\phantom{\text{Tier 1}} \quad \text{microclimate}_2 \leftarrow \beta_2 \cdot \text{land\_use}$$
$$\text{Tier 2} \quad \text{Root} \leftarrow \beta_3 \cdot \text{microclimate}_1 + \beta_4 \cdot \text{microclimate}_2 + \beta_5 \cdot \text{land\_use}$$
$$\text{Tier 3} \quad \text{Fauna} \leftarrow \beta_6 \cdot \text{microclimate}_1 + \beta_7 \cdot \text{microclimate}_2 + \beta_8 \cdot \text{Root} + \beta_9 \cdot \text{land\_use}$$

Each equation uses a **nested random intercept** (`orientation / depth`) and an **AR(1) correlation structure** to account for temporal autocorrelation within scanner groups.

**Transformations** are applied differentially before z-score standardisation:
- `root` and `fauna` variables: transformation selected during sensitivity analysis (any of none, Freeman-Tukey, log, Hellinger)
- `microclimate` indices: only `"log"` or `"none"` — count-specific transforms are not meaningful for continuous environmental variables
- `land_use`: excluded from z-scoring to preserve its interpretation as a binary 0/1 contrast

**Null distribution**: all causal variables (fauna, root, microclimate indices) are permuted independently across 30 iterations, breaking every directed link while preserving marginal distributions. Observed coefficients with |z| > 1.96 relative to the null are considered non-random.

---

## 1. Setup

```{r setup}
#| message: false

# ── Locale ────────────────────────────────────────────────────────────────────
invisible(Sys.setlocale("LC_TIME", "C"))

# ── Libraries ─────────────────────────────────────────────────────────────────
libs <- c(
  "tidyverse",    # Data wrangling and visualisation
  "lubridate",    # Date/time manipulation
  "piecewiseSEM", # Piecewise structural equation modelling
  "performance",  # Model performance metrics
  "lme4",         # Linear mixed-effects models
  "purrr",        # Functional iteration
  "nlme",         # LME with correlation structures (corAR1)
  "moments",      # Skewness and kurtosis
  "car",          # VIF
  "jsonlite",     # Reading the JSON parameter file
  "knitr"         # Table rendering
)
invisible(lapply(libs, require, character.only = TRUE))
```

---

## 2. Parameter loading

Parameters were determined in `sem_diagnostic.qmd` and saved to `model_parameters.txt`.
If the file is missing, safe fallback defaults are used with a warning.

```{r load-params}
if (file.exists("../output/model_parameters.txt")) {

  best_params          <- read_json("../output/model_parameters.txt")
  WIN_DAYS             <- as.numeric(best_params$selected_window)
  MIN_RATE             <- as.numeric(best_params$selected_min_rate)
  TRANSFO_ROOT         <- as.character(best_params$transform_root)
  TRANSFO_FAUNA        <- as.character(best_params$transform_fauna)
  TRANSFO_MICROCLIMATE <- as.character(best_params$transform_microclimate)

} else {
  # Fallback: safe defaults used when the config file is unavailable
  WIN_DAYS             <- 14
  MIN_RATE             <- 0.3
  TRANSFO_ROOT         <- "sqrt_plus_one"
  TRANSFO_FAUNA        <- "sqrt_plus_one"
  TRANSFO_MICROCLIMATE <- "none"
  warning("Configuration file not found. Using default fallback parameters.")
}

message(sprintf(
  "Active config — Window: %d days | Min. rate: %.1f | Transfo root: %s | Transfo fauna: %s | Transfo microclimate: %s",
  WIN_DAYS, MIN_RATE, TRANSFO_ROOT, TRANSFO_FAUNA, TRANSFO_MICROCLIMATE
))
```

---

## 3. Data loading and preprocessing

```{r data-preparation}

# ── Load raw database ─────────────────────────────────────────────────────────
data_sem_raw <- read.csv("../output/SEM_database.csv",
                         sep        = ",",
                         na.strings = c("", "NA")) %>%
  mutate(
    position = as.factor(position),
    depth    = as.factor(depth),
    # Binary land-use contrast: 0 = unmanaged (A), 1 = managed (C)
    land_use = as.numeric(position == "C"),
    date     = ymd_hms(date, tz = "UTC"),
    root     = root_growth_cm2_count_24h   # 24-hour root growth flux (cm²)
  )

# ── Load metadata ─────────────────────────────────────────────────────────────
# Taxa names and microclimate index column names, exported by data_preparation.qmd
fauna_vars         <- c("all_taxa", readLines("../output/fauna_vars.txt"))
microclimate_index <- readLines("../output/microclimate_index.txt")

# ── Transformation helper ─────────────────────────────────────────────────────
#' Apply a named variance-stabilising transformation to a numeric vector
#'
#' @param x       Numeric vector (counts or densities).
#' @param method  Character. One of: "none", "sqrt_plus_one", "log", "hellinger".
#' @return Transformed numeric vector.
apply_transfo <- function(x, method) {
  switch(method,
    "sqrt_plus_one" = sqrt(x) + sqrt(x + 1),          # Freeman-Tukey
    "log"           = log1p(x),                         # log(x + 1)
    "hellinger"     = sqrt(x / sum(x, na.rm = TRUE)),  # window-relative scale
    x                                                   # "none" or NA: identity
  )
}

# ── Apply differential transformations ────────────────────────────────────────
data_transformed <- data_sem_raw %>%

  # Root: biological count variable — all transforms are valid candidates
  mutate(root = apply_transfo(root, TRANSFO_ROOT)) %>%

  # Fauna variables: same biological rationale as root
  mutate(across(
    all_of(intersect(fauna_vars, names(.))),
    ~ apply_transfo(., TRANSFO_FAUNA)
  )) %>%

  # Microclimate indices: only "log" or "none" are meaningful for continuous
  # environmental variables; count-specific transforms (Freeman-Tukey, Hellinger)
  # are not applied even if selected — TRANSFO_MICROCLIMATE is always "log"/"none"
  mutate(across(
    all_of(intersect(microclimate_index, names(.))),
    ~ apply_transfo(., TRANSFO_MICROCLIMATE)
  )) %>%

  # Z-score standardisation: applied to all numeric variables EXCEPT land_use,
  # which is a binary 0/1 contrast and must not be rescaled
  mutate(across(
    where(is.numeric) & !all_of("land_use"),
    ~ as.numeric(scale(.))
  ))
```

---

## 4. Analytical functions

### 4.1 Rolling window splitter

```{r rolling-window-generator}
#' Split a time series into overlapping rolling windows
#'
#' @param df          Dataframe with a `date` column (POSIXct).
#' @param window_days Integer. Window duration in days.
#' @param step_days   Integer. Step between consecutive window start dates.
#' @return Named list of dataframes, one per window start date.
create_rolling_windows <- function(df, window_days, step_days = 1) {
  start_dates <- seq(
    min(df$date),
    max(df$date) - days(window_days) + hours(6),
    by = paste(step_days, "days")
  )

  windows        <- map(start_dates, ~ filter(df, date >= .x & date < .x + days(window_days)))
  names(windows) <- format(start_dates, "%Y-%m-%d")
  return(windows)
}
```

### 4.2 SEM structure

The four-equation causal model is fixed across all windows and taxa.
`land_use` enters all three tiers as the exogenous management driver.

```{r SEM-structure}
#' Fit the piecewise SEM for a single data window
#'
#' @param data      Dataframe with columns: `land_use`, `root`, `fauna`,
#'                  `microclimate_1`, `microclimate_2`, `orientation`, `depth`.
#' @param m_indices Character vector of length 2 (microclimate index names).
#'                  Used by the pipeline wrapper; internally mapped to
#'                  `microclimate_1` and `microclimate_2`.
#' @return A `psem` object.
fit_psem <- function(data, m_indices) {
  psem(

    # ── Tier 1: land use drives microclimate ──────────────────────────────────
    lme(microclimate_1 ~ land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    lme(microclimate_2 ~ land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    # ── Tier 2: microclimate + land use drive root ────────────────────────────
    lme(root ~ microclimate_1 + microclimate_2 + land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    # ── Tier 3: everything drives fauna ──────────────────────────────────────
    lme(fauna ~ microclimate_1 + microclimate_2 + root + land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim"))
  )
}
```

### 4.3 Main pipeline

```{r fixed-pipeline}
#' Run the full rolling-window SEM pipeline for one taxon variable
#'
#' Iterates over all valid time windows, fits the four-equation pSEM, and
#' extracts standardised path coefficients and R² values.
#' Both management positions (A and C) are modelled jointly via `land_use`;
#' there is no longer a per-position split.
#'
#' @param taxon_var   Character. Column name of the fauna variable to model.
#' @param data_raw    Dataframe. Transformed and z-score standardised SEM database.
#' @param m_indices   Character vector of length 2. Microclimate index column names.
#' @param window_days Integer. Window duration (days).
#' @param min_rate    Numeric [0, 1]. Minimum data completeness ratio.
#'                    Windows below this threshold are skipped.
#'                    (Max possible rows per window = window_days × 16:
#'                     4 time steps/day × 2 depths × 2 orientations.)
#' @return Dataframe with one row per path coefficient
#'         (taxon × window × predictor → response).
run_sem_pipeline <- function(taxon_var, data_raw, m_indices,
                             window_days, min_rate) {

  cat("\n--- Processing taxon:", taxon_var, "---\n")

  # Rename dynamic columns to fixed SEM internal names
  sem_data <- data_raw %>%
    mutate(
      fauna          = !!sym(taxon_var),
      microclimate_1 = !!sym(m_indices[1]),
      microclimate_2 = !!sym(m_indices[2])
    ) %>%
    select(date, land_use, orientation, depth,
           fauna, root, microclimate_1, microclimate_2)

  windows <- create_rolling_windows(sem_data, window_days, step_days = 1)

  if (length(windows) == 0) return(NULL)

  # Discard windows with insufficient data density
  # Both positions combined: max slots = window_days × 16 per position × 2 positions
  valid_wins <- keep(windows,
                     ~ (nrow(drop_na(.x)) / (window_days * 16 * 2)) >= min_rate)

  cat(sprintf("  %d / %d windows passed filter.\n",
              length(valid_wins), length(windows)))

  if (length(valid_wins) == 0) return(NULL)

  results_list <- map(names(valid_wins), function(w_name) {
    win_data <- valid_wins[[w_name]] %>%
      drop_na() %>%
      arrange(land_use, orientation, depth, date)

    # Guard: skip windows where ANY modelled variable is constant
    # (zero variance causes Cholesky failures in lme and summary.psem)
    vars_to_check <- c("fauna", "root", "microclimate_1", "microclimate_2")
    if (any(sapply(win_data[vars_to_check], sd, na.rm = TRUE) == 0)) return(NULL)

    # Guard: skip windows where land_use has only one level
    # (both positions must be present for land_use to be estimable)
    if (length(unique(win_data$land_use)) < 2) return(NULL)

    # Fit the SEM; mute warnings from lme optimisation; catch convergence errors
    model <- tryCatch(
      withCallingHandlers(
        fit_psem(win_data, m_indices),
        warning = function(w) invokeRestart("muffleWarning")
      ),
      error = function(e) NULL
    )
    if (is.null(model)) return(NULL)

    # summary.psem() can itself trigger chol.default() failures — wrap it too
    sm <- tryCatch(summary(model), error = function(e) NULL)
    if (is.null(sm)) return(NULL)

    # ── Extract path coefficients ──────────────────────────────────────────────
    coefs <- as.data.frame(sm$coefficients)

    # ── Extract R² ────────────────────────────────────────────────────────────
    # marginal = fixed effects only; conditional = fixed + random
    # Column names vary across piecewiseSEM versions — detected dynamically
    r2_raw <- as.data.frame(sm$R2)
    r2_df  <- r2_raw %>%
      mutate(
        r2_cond = .[[grep("R.squared|Conditional", names(.))[1]]],
        r2_marg = .[[grep("Marginal",              names(.))[1]]]
      ) %>%
      select(response = Response, r2_cond, r2_marg)

    tibble(
      taxon        = taxon_var,
      date_start   = as.Date(w_name),
      n_obs        = sm$n,
      response     = coefs$Response,
      predictor    = coefs$Predictor,
      estimate     = coefs$Estimate,
      std_estimate = coefs$Std.Estimate,
      # "<0.001" strings are coerced to numeric by stripping the "<" prefix
      p_value      = as.numeric(str_replace(coefs$P.Value, "<", ""))
    ) %>%
      left_join(r2_df, by = "response")
  }) %>%
    compact()

  return(map_dfr(results_list, ~ .x))
}
```

---

## 5. Model execution

### 5.1 Observed models

```{r observed-models}
message(">>> Running observed models across all taxa...")

sem_results_raw <- map_dfr(
  fauna_vars,
  ~ run_sem_pipeline(.x, data_transformed, microclimate_index, WIN_DAYS, MIN_RATE)
)
```

### 5.2 Null distribution via permutation

The null distribution is built by independently permuting all causal variables
(fauna, root, microclimate indices) across `n_perm = 30` iterations. `land_use`
is **not** permuted — it is the exogenous management driver and its observed
assignment to scanners must be preserved.

```{r null-distribution}
n_perm <- 30
message(sprintf(">>> Generating null distribution (n = %d permutations)...", n_perm))

sem_results_null <- map_dfr(1:n_perm, function(i) {
  data_null <- data_transformed %>%
    mutate(across(
      # Permute all endogenous causal variables; leave land_use intact
      all_of(c(fauna_vars, "root", microclimate_index)),
      ~ sample(.)
    ))

  run_sem_pipeline(fauna_vars[1], data_null, microclimate_index,
                   WIN_DAYS, MIN_RATE) %>%
    mutate(iteration = i)
})

# Summarise null: mean and SD of each path coefficient under H₀
sem_null_summary <- sem_results_null %>%
  group_by(response, predictor) %>%
  summarise(
    null_mean_estimate = mean(estimate, na.rm = TRUE),
    null_sd_estimate   = sd(estimate,   na.rm = TRUE),
    .groups = "drop"
  )
```

---

## 6. Consolidation and export

```{r consolidation}

# ── Merge observed results with null summary ───────────────────────────────────
sem_results <- sem_results_raw %>%
  mutate(type = "observed") %>%
  left_join(sem_null_summary, by = c("response", "predictor")) %>%
  # Keep only directed causal paths (→); remove correlated error terms (~~)
  filter(!str_detect(predictor, "~~")) %>%
  mutate(path_id = paste(predictor, "->", response))

# ── Season labels ──────────────────────────────────────────────────────────────
season_lookup <- tibble(
  month  = 1:12,
  season = c("Winter", "Winter",
             "Spring", "Spring", "Spring",
             "Summer", "Summer", "Summer",
             "Autumn", "Autumn", "Autumn",
             "Winter")
)

sem_results <- sem_results %>%
  mutate(month = month(date_start)) %>%
  left_join(season_lookup, by = "month")

# ── Export ─────────────────────────────────────────────────────────────────────
write.csv(sem_results, "../output/SEM_results_database.csv", row.names = FALSE)

message(sprintf(
  "Export complete: %d rows | %d unique taxa | %d unique paths",
  nrow(sem_results),
  n_distinct(sem_results$taxon),
  n_distinct(sem_results$path_id)
))
```
