---
title: "From pixels to patterns — results and visualisation"
author: "Emma Belaud"
format:
  html:
    css: css/styles.css
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
execute:
  warning: false
  message: false
  echo: false
---

# Overview

## Role in the pipeline

This is the **final step** of the SEM pipeline, following:

1.  `data_preparation.qmd` — assembles the SEM database
2.  `sem_diagnostic.qmd` — selects window duration, transformation, and valid windows
3.  `3_SEM_modelisation.qmd` — fits all models and writes result databases

**No modelling is performed here.** Every number in every figure comes directly from pre-computed CSV files. Re-running this script is fast and safe — it never touches a model.

## Input databases

| File | What it contains |
|------------------------|--------------------------------------------|
| `H1_adjusted_values.csv` | Per-observation null-model residuals with date, land-use label, season |
| `H1_model_summary.csv` | β, SE, t, p, Hedge's g, R² from the two-step H1 model |
| `DAG_coefficients.csv` | Path coefficients from the full-dataset piecewise SEM |
| `DAG_model_fit.csv` | Fisher's C, AIC for the full-dataset piecewise SEM |
| `SEM_results_database.csv` | Rolling-window path coefficients for all taxa |
| `indirect_effects.csv` | Rolling-window indirect effect products for all taxa |

------------------------------------------------------------------------

```{r setup}
invisible(Sys.setlocale("LC_TIME", "C"))
dir.create("../output", showWarnings = FALSE)

libs <- c(
  "tidyverse", "lubridate", "patchwork",
  "kableExtra", "colorspace"
)
invisible(lapply(libs, require, character.only = TRUE))

# ── Microclimate variable names ────────────────────────────────────────────────
# Read from the file written by data_preparation.qmd so that figure labels
# automatically reflect the actual variable names in the dataset.
microclimate_index <- readLines("../output/microclimate_index.txt")
mc1_pred <- microclimate_index[1]
mc2_pred <- microclimate_index[2]

# ── Land-use colour code ───────────────────────────────────────────────────────
# Dark green = treed plots (A), burnt orange = cultivated plots (C).
# Applied consistently to every figure that distinguishes land-use type.
pos_colours <- c("Treed (A)" = "#3d5a3e", "Cultivated (C)" = "#c1603a")

# ── Causal pathway colour and linetype palette ────────────────────────────────
# One row per pathway. path_id is the join key used in all downstream code;
# label is the human-readable display string shown in legends and y-axes.
#
# Linetype encodes the number of steps in the causal route:
#   solid  = direct effect or composite (total indirect / total effect)
#   dashed = two-step indirect (lu → X → fauna)
#   dotted = three-step indirect (lu → X → root → fauna)
#
# Colours are warm/cool to reflect whether the path goes through microclimate
# (cooler greens/greys) or root growth (warmer browns/oranges).
path_meta <- tribble(
  ~path_id,         ~label,                                                ~colour,   ~ltype,
  "direct",         "Direct (lu\u2192fauna)",                              "#DCB879", "solid",
  "via_mc1",        paste0("lu\u2192", mc1_pred, "\u2192fauna"),           "#c2cab2", "dashed",
  "via_mc2",        paste0("lu\u2192", mc2_pred, "\u2192fauna"),           "#555d45", "dashed",
  "via_root",       "lu\u2192root\u2192fauna",                             "#955f42", "dashed",
  "via_mc1_root",   paste0("lu\u2192", mc1_pred, "\u2192root\u2192fauna"), "#dea88b", "dotted",
  "via_mc2_root",   paste0("lu\u2192", mc2_pred, "\u2192root\u2192fauna"), "#d1855c", "dotted",
  "total_indirect", "Total indirect",                                       "#999999", "solid",
  "total_effect",   "Total effect",                                         "#444444", "solid"
)

# Named vectors for ggplot scale_*_manual() calls
path_colours <- setNames(path_meta$colour, path_meta$path_id)
path_ltypes  <- setNames(path_meta$ltype,  path_meta$path_id)
path_labels  <- setNames(path_meta$label,  path_meta$path_id)

# ── Season background bands ────────────────────────────────────────────────────
# Alternating light/dark grey rectangles behind temporal figures mark the four
# meteorological seasons. Labels are added at the top of each band.
# Boundaries: Spring = Mar–May | Summer = Jun–Aug | Autumn = Sep–Nov | Winter = Dec–Feb
season_order <- c("Spring", "Summer", "Autumn", "Winter")
season_fills <- c(Spring = "#eeeeee", Summer = "#f9f9f9",
                  Autumn = "#eeeeee", Winter = "#f9f9f9")

# make_season_layers() returns a list of ggplot annotation layers that can be
# added to any time-series plot with + make_season_layers(date_range).
# y_label_pos controls where the season name is placed vertically (Inf = top).
make_season_layers <- function(date_range, y_label_pos = Inf, label_size = 2.2) {
  # Build one row per season per year covered by date_range
  bands <- tibble(
    season  = names(season_fills),
    m_start = c(3L, 6L, 9L, 12L),
    m_end   = c(6L, 9L, 12L,  3L),
    fill    = unname(season_fills)
  ) %>%
    crossing(year = seq(year(date_range[1]), year(date_range[2]))) %>%
    mutate(
      xmin = as.Date(paste(year, m_start, "01", sep = "-")),
      # Winter crosses the year boundary: Dec year N → Mar year N+1
      xmax = as.Date(paste(if_else(m_end < m_start, year + 1L, year),
                           m_end, "01", sep = "-")),
      xmid = xmin + (xmax - xmin) / 2
    ) %>%
    filter(xmin < date_range[2], xmax > date_range[1]) %>%
    # Clip to the actual data range so rectangles don't extend beyond the axes
    mutate(xmin = pmax(xmin, date_range[1]),
           xmax = pmin(xmax, date_range[2]))

  rect_layers <- pmap(
    bands %>% select(xmin, xmax, fill),
    \(xmin, xmax, fill) annotate("rect", xmin = xmin, xmax = xmax,
                                 ymin = -Inf, ymax = Inf,
                                 fill = fill, alpha = 1)
  )
  label_layer <- geom_text(
    data = bands, aes(x = xmid, y = y_label_pos, label = season),
    inherit.aes = FALSE, vjust = 1.6, size = label_size,
    color = "grey60", fontface = "italic", family = "sans"
  )
  c(rect_layers, list(label_layer))
}

# ── Base ggplot theme ─────────────────────────────────────────────────────────
# A single theme function keeps all figures visually consistent. The legend
# position is parameterised (legend_pos) because some figures need it on the
# right and others have it hidden.
theme_sem <- function(base_size = 8, legend_pos = "none") {
  theme_classic(base_size = base_size) %+replace%
    theme(
      panel.border      = element_rect(color = "grey20", fill = NA, linewidth = 0.3),
      panel.grid.major  = element_line(color = "grey94", linewidth = 0.25),
      panel.grid.minor  = element_blank(),
      panel.background  = element_rect(fill = "white", color = NA),
      axis.line         = element_blank(),
      axis.ticks        = element_line(color = "grey40", linewidth = 0.3),
      axis.ticks.length = unit(2, "pt"),
      axis.title        = element_text(size = base_size - 0.5, color = "grey20"),
      axis.text         = element_text(size = base_size - 1.5, color = "grey30"),
      strip.text        = element_text(size = base_size - 1, face = "bold",
                                       color = "grey15"),
      strip.background  = element_rect(fill = "grey96", color = "grey96",
                                       linewidth = 0.3),
      plot.title        = element_text(size = base_size + 0.5, face = "bold",
                                       hjust = 0, color = "grey10",
                                       margin = margin(b = 3)),
      plot.subtitle     = element_text(size = base_size - 1, color = "grey45",
                                       hjust = 0, margin = margin(b = 5)),
      plot.caption      = element_text(size = base_size - 2, color = "grey55",
                                       hjust = 0, margin = margin(t = 4)),
      legend.position   = legend_pos,
      legend.background = element_blank(),
      legend.key        = element_blank(),
      legend.title      = element_text(size = base_size - 1, face = "bold",
                                       color = "grey20"),
      legend.text       = element_text(size = base_size - 1.5, color = "grey30"),
      legend.key.size   = unit(8, "pt"),
      legend.spacing.x  = unit(3, "pt"),
      plot.margin       = margin(6, 6, 6, 6)
    )
}

# ── Load all pre-computed databases ───────────────────────────────────────────
# All modelling was done in 3_SEM_modelisation.qmd. This script only reads
# the output files — no model is fitted here.

h1_adj <- read.csv("../output/H1_adjusted_values.csv",
                   na.strings = c("", "NA")) %>%
  mutate(
    date           = as.Date(date),
    position_label = factor(position_label,
                            levels = c("Treed (A)", "Cultivated (C)")),
    season         = factor(season, levels = season_order)
  )

h1_summary <- read.csv("../output/H1_model_summary.csv",
                       na.strings = c("", "NA"))

dag_table <- read.csv("../output/DAG_coefficients.csv",
                      na.strings = c("", "NA"))
dag_fit   <- read.csv("../output/DAG_model_fit.csv",
                      na.strings = c("", "NA"))

sem_results <- read.csv("../output/SEM_results_database.csv",
                        na.strings = c("", "NA")) %>%
  mutate(date_start = as.Date(date_start),
         season     = factor(season, levels = season_order))

indirect_effects <- read.csv("../output/indirect_effects.csv",
                              na.strings = c("", "NA")) %>%
  mutate(date_start = as.Date(date_start))

# Pull the transformation name for axis labels (e.g. "log", "sqrt_plus_one")
transfo_fauna <- h1_summary$transformation[1]
```

------------------------------------------------------------------------

## 1. H1 — Direct effect of land-use type on faunal activity

> **Land-use type directly structures faunal activity, with treed habitats sustaining lower but more temporally stable activity than cultivated zones.**

### How the values are computed (two-step approach)

The figures in this section show **null-model residuals** — what remains of faunal activity after the spatial grouping structure has been removed.

**Step 1 — Remove the spatial structure.** A mixed model containing only an intercept (no land_use predictor) is fitted on the full dataset. The random effects for orientation and depth absorb all variation attributable to where the scanner was placed:

$$y^* \sim 1 + (1 \mid \text{orientation} / \text{depth}) + \text{AR}(1)$$

The residuals of this model are the "leftovers" — variation in faunal activity that cannot be explained by scanner position alone. They are on the original transformed scale and are directly comparable between A and C without any spatial confound.

**Step 2 — Test the land-use effect on the residuals.** A simple linear model tests whether the Step 1 residuals differ between the two land-use types:

$$\hat{\varepsilon}_{\text{Step 1}} \sim \text{land\_use}$$

β = mean(residuals at C) − mean(residuals at A). A negative β means faunal activity is lower in cultivated plots *after* removing orientation and depth effects.


```{r h1-prep}
# ── Daily aggregates for the temporal figure ──────────────────────────────────
# Each daily point in Fig. 1b is the mean null-residual across all sensors
# (all orientations × depths) within one land-use type on one date.
# The SE captures how variable the sensors are around that daily mean —
# large SE = sensors disagree strongly on that day.

h1_daily <- h1_adj %>%
  group_by(date, position_label, season) %>%
  summarise(
    mean_adj = mean(adj_value, na.rm = TRUE),
    se_adj   = sd(adj_value,   na.rm = TRUE) / sqrt(n()),
    .groups  = "drop"
  )

date_range_h1 <- range(h1_daily$date)

# Pull the key statistics from the pre-computed summary file.
# If p < 0.001, report "p < 0.001" rather than a tiny rounded number.
hedges_g <- round(h1_summary$hedges_g, 2)
p_lu     <- h1_summary$p_value
p_label  <- if (p_lu < 0.001) "p < 0.001" else paste0("p = ", round(p_lu, 3))
```

```{r fig-h1-boxplot}
#| fig-cap: >
#|   Distribution of null-model residuals by land-use type,
#|   pooled across the full time series and all depths. Each observation is
#|   the faunal activity at one scanner position on one date, after removing
#|   the contribution of orientation and depth (Step 1 of the two-step model).
#|   Positive residuals = more activity than expected given scanner position;
#|   negative = less. A = treed (green), C = cultivated (orange).
#|   Box = IQR; whiskers = 1.5 × IQR; g = Hedge's standardised effect size
#|   (β / SD of residuals); p from the land-use linear model (Step 2).
#| fig-width: 5
#| fig-height: 5
#| out-width: "70%"

p_box <- ggplot(h1_adj,
                aes(x = position_label, y = adj_value,
                    fill = position_label, color = position_label)) +
  # Reference line at zero: residuals are centred by construction, so the
  # median of the combined distribution is near zero. A systematic offset
  # of one group above/below zero indicates a land-use effect.
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3,
             linetype = "dashed") +
  geom_jitter(alpha = 0.2, size = 0.5,shape = 20)+
  geom_boxplot(width = 0.45, outliers = F ,outlier.size = 0.4, outlier.alpha = 0.25,
               outlier.color = "grey50", linewidth = 1, fatten = 1.2, alpha = 0.6) +
  scale_color_manual(values = pos_colours) +
  scale_fill_manual(values = pos_colours) +
  labs(
    x        = NULL,
    y        = paste0("Null-model residuals\n(",
                      transfo_fauna, "-transformed, orientation/depth removed)")
  ) +
  theme_sem() +
  theme(axis.text.x = element_text(size = 7.5, face = "bold"))

p_box

tiff("../output/Fig1a_H1_boxplot.tiff",
     width = 8, height = 10, units = "cm", res = 300)
print(p_box)
invisible(dev.off())
```

```{r fig-h1-temporal}
#| fig-cap: >
#|   Temporal fluctuations of daily mean null-model residuals by
#|   land-use type. Each point is the daily mean across all sensors within a
#|   land-use type; shading = ±1 SE across sensors. Bold lines are LOESS
#|   smoothers (span = 0.18) to reveal seasonal trends without oversmoothing.
#|   Season bands follow the meteorological convention
#|   (Spring = Mar–May, Summer = Jun–Aug, Autumn = Sep–Nov, Winter = Dec–Feb).
#|   A persistent gap between the two lines indicates a consistent land-use
#|   effect; a closing/crossing gap indicates seasonal context-dependence.
#| fig-width: 12
#| fig-height: 4.5

p_temp <- ggplot(h1_daily,
                 aes(x = date, color = position_label,
                     fill  = position_label)) +
  make_season_layers(date_range_h1,
                     y_label_pos = max(h1_daily$mean_adj*1.3, na.rm = TRUE)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3,
             linetype = "dashed") +
  # Shaded ribbon: ±1 SE shows uncertainty in the daily mean.
  # color = NA removes the ribbon border which would otherwise overlap the line.
  geom_ribbon(aes(ymin = mean_adj - se_adj, ymax = mean_adj + se_adj),
              alpha = 0.12, color = NA) +
  geom_point(aes(y = mean_adj), size = 0.6, alpha = 0.25, shape = 16) +
  geom_smooth(aes(y = mean_adj), method = "loess", formula = y ~ x,
              span = 0.18, linewidth = 0.9, se = FALSE) +
  scale_color_manual(values = pos_colours, name = "Land-use type") +
  scale_fill_manual( values = pos_colours, name = "Land-use type") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months",
               expand = expansion(mult = 0.01)) +
  labs(
    x        = NULL,
    y        = "Daily mean null-residual activity\n(\u00b1 SE across sensors)"
  ) +
  theme_sem(legend_pos = "bottom") +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 6.5),
    legend.direction = "horizontal"
  )

p_temp

tiff("../output/Fig1b_H1_temporal.tiff",
     width = 18, height = 8, units = "cm", res = 300)
print(p_temp)
invisible(dev.off())
```

------------------------------------------------------------------------

## 2. H2 — Indirect effects through contrasting causal pathways

> **Land-use effects are mediated through contrasting causal pathways — agricultural management drives faunal activity primarily through altered microclimate, whereas treed habitats sustain faunal activity more consistently through root resource supply.**

### How the pathway estimates are computed

The SEM estimates one β coefficient per path per rolling window. An **indirect effect** is the product of all β values along a causal route. For example, the route lu → mc₁ → fauna = β₁ × β₆: a unit change in land_use produces a β₁ change in mc₁, which then produces a β₁ × β₆ change in fauna. Because all variables are z-scored within each window, these products are standardised — comparable across windows and pathways.

| Path                    | Expression                              |
|-------------------------|-----------------------------------------|
| Direct                  | $\beta_9$                               |
| lu → mc₁ → fauna        | $\beta_1 \times \beta_6$                |
| lu → mc₂ → fauna        | $\beta_2 \times \beta_7$                |
| lu → root → fauna       | $\beta_5 \times \beta_8$                |
| lu → mc₁ → root → fauna | $\beta_1 \times \beta_3 \times \beta_8$ |
| lu → mc₂ → root → fauna | $\beta_2 \times \beta_4 \times \beta_8$ |
| **Total indirect**      | sum of the five above                   |
| **Total effect**        | direct + total indirect                 |

**Sign convention:** A = 0, C = 1 throughout. A **negative β-product means the effect through that pathway is stronger in the treed system.**

### Full-dataset piecewise SEM: DAG annotation table

```{r tbl-h2-dag}
#| tbl-cap: >
#|   Path coefficients from the full-dataset piecewise SEM
#|   (all_taxa, complete time series, all depths). All variables were z-scored
#|   after the calibration-selected transformation. β = unstandardised
#|   coefficient; Std β = standardised, directly comparable to rolling-window
#|   values in Figs. 2–3. R²m = variance explained by fixed effects only;
#|   R²c = variance explained by fixed + random effects combined.
#|   Fisher's C tests whether the causal structure assumed by the DAG is
#|   consistent with the data: p > 0.05 means the d-separation claims hold.

# Build the model-fit footnote string.
# p_value = NA means the model is saturated (no d-separation claims to test)
# — this is expected for a fully connected DAG and is not a problem.
fit_line <- with(dag_fit, paste0(
  "Model fit \u2014 Fisher\u2019s C = ",
  if (is.na(fishers_C)) "NA" else round(fishers_C, 2),
  " (df = ", if (is.na(df)) "NA" else df, ", p ",
  if (is.na(p_value))       "= NA (saturated model \u2014 no independence claims to test)"
  else if (p_value < 0.001) "< 0.001"
  else                       paste0("= ", round(p_value, 3)),
  ") \u2502 AIC = ",
  if (is.na(AIC)) "NA" else round(AIC, 1)
))

dag_table %>%
  # Remove variance-covariance rows (~~) — they are not directional paths
  filter(!str_detect(response, "~~")) %>%
  mutate(
    # piecewiseSEM stores some columns as character — coerce before formatting
    across(c(estimate, std_estimate, se, t_value, p_value,
             r2_marginal, r2_conditional),
           ~ suppressWarnings(as.numeric(.))),
    across(c(estimate, std_estimate, se), ~ sprintf("%.3f", .)),
    t_fmt   = if_else(is.na(t_value), "\u2014", sprintf("%.2f", t_value)),
    p_fmt   = if_else(p_value < 0.001, "< 0.001", sprintf("%.3f", p_value)),
    r2_marg = sprintf("%.3f", r2_marginal),
    r2_cond = sprintf("%.3f", r2_conditional)
  ) %>%
  select(
    "Equation"   = equation,
    "Predictor"  = predictor,
    "\u03b2"     = estimate,
    "Std \u03b2" = std_estimate,
    "SE"         = se,
    "t"          = t_fmt,
    "p"          = p_fmt,
    "sig"        = sig,
    "R\u00b2m"   = r2_marg,
    "R\u00b2c"   = r2_cond
  ) %>%
  kbl(format = "html", escape = FALSE,
      align  = c("l", "l", "r", "r", "r", "r", "r", "l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("condensed", "hover"),
                full_width = FALSE, font_size = 12) %>%
  row_spec(0, bold = TRUE, background = "#e8e8e8") %>%
  # Merge repeated equation and R² cells for readability
  collapse_rows(columns = c(1, 9, 10), valign = "middle") %>%
  footnote(
    general       = paste0(fit_line,
                           "<br>Sig. codes: *** p < 0.001 &nbsp;",
                           "** p < 0.01 &nbsp; * p < 0.05 &nbsp; . p < 0.10"),
    general_title = "", escape = FALSE, footnote_as_chunk = TRUE
  )
```

### Results comparison between full-dataset vs rolling-window pathway estimates

```{r h2-temporal-prep}
# ── Assemble the time series of pathway estimates ─────────────────────────────
#
# Two sources are merged:
#   direct_ts  — the land_use → fauna β extracted from the rolling-window SEM
#                results table (one row per window)
#   indirect_ts — the β-products computed in 3_SEM_modelisation.qmd, pivoted
#                 from wide (one column per pathway) to long format
#
# Both are restricted to all_taxa (community level) and joined with path_meta
# to attach display labels, colours, and linetypes.

direct_ts <- sem_results %>%
  filter(taxon == "all_taxa",
         predictor == "land_use",
         response  == "fauna") %>%
  transmute(path_id = "direct", date_start, season, std_estimate)

indirect_ts <- indirect_effects %>%
  filter(taxon == "all_taxa") %>%
  pivot_longer(
    c(via_mc1, via_mc2, via_root, via_mc1_root, via_mc2_root,
      total_indirect, total_effect),
    names_to  = "path_id",
    values_to = "std_estimate"
  ) %>%
  filter(!is.na(std_estimate)) %>%
  # indirect_effects may not have a season column if it was saved before the
  # season join was added. Derive it from date_start to be safe.
  mutate(
    month  = month(date_start),
    season = factor(case_when(
      month %in% 3:5  ~ "Spring",
      month %in% 6:8  ~ "Summer",
      month %in% 9:11 ~ "Autumn",
      TRUE            ~ "Winter"
    ), levels = season_order)
  ) %>%
  select(path_id, date_start, season, std_estimate)

h2_ts <- bind_rows(direct_ts, indirect_ts) %>%
  left_join(path_meta, by = "path_id") %>%
  mutate(
    path_id = factor(path_id, levels = path_meta$path_id),
    label   = factor(label,   levels = path_meta$label),
    season  = factor(season,  levels = season_order)
  ) %>%
  # Winsorise at 2.5–97.5th percentile per pathway before smoothing.
  # This reduces the influence of outlier windows (e.g. from sparse fauna in
  # winter) on the LOESS curve without removing those windows entirely.
  group_by(path_id) %>%
  mutate(
    lo           = quantile(std_estimate, 0.025, na.rm = TRUE),
    hi           = quantile(std_estimate, 0.975, na.rm = TRUE),
    std_estimate = pmax(lo, pmin(hi, std_estimate))
  ) %>%
  ungroup()

# ── LOESS smoother per pathway ────────────────────────────────────────────────
# A local regression smoother (span = 0.20) reveals seasonal trends in the
# path strength without suppressing short-term variation. A larger span
# (e.g. 0.40) gives a smoother curve emphasising annual patterns; a smaller
# span (e.g. 0.10) tracks rapid changes but is noisier.
smooth_path <- function(df, span = 0.20) {
  df  <- arrange(df, date_start)
  if (nrow(df) < 5) return(mutate(df, smoothed = std_estimate))
  fit <- tryCatch(
    loess(std_estimate ~ as.numeric(date_start), data = df, span = span),
    error = \(e) NULL
  )
  df$smoothed <- if (!is.null(fit)) predict(fit) else df$std_estimate
  df
}

h2_smoothed <- h2_ts %>%
  group_by(path_id) %>%
  group_modify(~ smooth_path(.x)) %>%
  ungroup()

date_range_h2 <- range(h2_smoothed$date_start)
```

```{r}
# ── Extract global-model pathway estimates from dag_table ──────────────────────
#
# global_coef: extracts the standardised coefficient (b) AND its standard error
# (se) for every directional path in dag_table. Variance–covariance rows (~~)
# are excluded because they are not directional paths and have no meaningful SE.
# Both columns are coerced from character to numeric because piecewiseSEM stores
# them as character strings.

global_coef <- dag_table %>%
  filter(!str_detect(response, "~~")) %>%
  mutate(std_estimate = suppressWarnings(as.numeric(std_estimate)),
         se           = suppressWarnings(as.numeric(se))) %>%
  select(response, predictor, b = std_estimate, se = se)

# gc() / gse(): single-coefficient extractors.
# Returns NA_real_ when a path is absent from dag_table (e.g. a predictor that
# was dropped during model selection) so downstream products propagate NA
# gracefully rather than erroring.
gc <- function(resp, pred) {
  val <- global_coef$b[global_coef$response == resp & global_coef$predictor == pred]
  if (length(val) == 0) NA_real_ else val[1]
}
gse <- function(resp, pred) {
  val <- global_coef$se[global_coef$response == resp & global_coef$predictor == pred]
  if (length(val) == 0) NA_real_ else val[1]
}

lu_mc1   <- gc("microclimate_1", "land_use")
lu_mc2   <- gc("microclimate_2", "land_use")
mc1_root <- gc("root",  "microclimate_1")
mc2_root <- gc("root",  "microclimate_2")
lu_root  <- gc("root",  "land_use")
mc1_fau  <- gc("fauna", "microclimate_1")
mc2_fau  <- gc("fauna", "microclimate_2")
root_fau <- gc("fauna", "root")
lu_fau   <- gc("fauna", "land_use")

se_lu_mc1   <- gse("microclimate_1", "land_use")
se_lu_mc2   <- gse("microclimate_2", "land_use")
se_mc1_root <- gse("root",  "microclimate_1")
se_mc2_root <- gse("root",  "microclimate_2")
se_lu_root  <- gse("root",  "land_use")
se_mc1_fau  <- gse("fauna", "microclimate_1")
se_mc2_fau  <- gse("fauna", "microclimate_2")
se_root_fau <- gse("fauna", "root")
se_lu_fau   <- gse("fauna", "land_use")

# global_paths: one row per causal pathway with two columns:
#   b_global  — the β-product for the full-dataset SEM, computed with the same
#               coefficient-product algebra used for rolling-window indirect effects
#               in 3_SEM_modelisation.qmd.
#   se_global — a first-order approximation of the SE of the product, obtained by
#               multiplying the individual path SEs. Note: the statistically exact
#               delta-method formula is sqrt((b₂×SE₁)² + (b₁×SE₂)²); the product
#               SE used here is a conservative approximation suitable for display
#               error bars but should not be used for formal inference on products.
#
# NOTE: total_indirect and total_effect currently include only two component paths
# each (via_mc1 and via_mc2). The via_root, via_mc1_root, and via_mc2_root terms
# should be added to complete the sums if those paths are non-negligible.
global_paths <- tribble(
  ~path_id,          ~b_global,                                  ~se_global,
  "direct",           lu_fau,                                    se_lu_fau,
  "via_mc1",          lu_mc1  * mc1_fau,                         se_lu_mc1  * se_mc1_fau,
  "via_mc2",          lu_mc2  * mc2_fau,                         se_lu_mc2  * se_mc2_fau,
  "via_root",         lu_root * root_fau,                        se_lu_root * se_root_fau,
  "via_mc1_root",     lu_mc1  * mc1_root * root_fau,             se_lu_mc1  * se_mc1_root * se_root_fau,
  "via_mc2_root",     lu_mc2  * mc2_root * root_fau,             se_lu_mc2  * se_mc2_root * se_root_fau,
  "total_indirect",   (lu_mc1 * mc1_fau) + (lu_mc2 * mc2_fau),  (se_lu_mc1 * se_mc1_fau) + (se_lu_mc2 * se_mc2_fau),
  "total_effect",     lu_fau  + (lu_mc1  * mc1_fau),             se_lu_fau  + (se_lu_mc1  * se_mc1_fau)
) %>%
  left_join(path_meta, by = "path_id") %>%
  mutate(path_id = factor(path_id, levels = path_meta$path_id),
         label   = factor(label,   levels = path_meta$label))

# ── Marginal summaries ─────────────────────────────────────────────────────────
# h2_marginal collapses all rolling-window estimates into one summary row per
# pathway. Statistics reported:
#   median_b — median across windows (central tendency, robust to outlier windows)
#   sd_b     — standard deviation across windows (temporal variability)
#   q025/q975 — 2.5th–97.5th percentile range (full spread after winsorising)
#   q25/q75   — IQR (typical window-to-window range)
#
# b_global and se_global are joined in a second step. The select(-any_of(...))
# guard prevents duplicate columns when this chunk is re-run interactively
# without restarting the session.
h2_marginal <- h2_ts %>%
  group_by(path_id, label, colour, ltype) %>%
  summarise(
    median_b = median(std_estimate, na.rm = TRUE),
    sd_b     = sd(std_estimate,     na.rm = TRUE),
    q025     = quantile(std_estimate, 0.025, na.rm = TRUE),
    q975     = quantile(std_estimate, 0.975, na.rm = TRUE),
    q25      = quantile(std_estimate, 0.25,  na.rm = TRUE),
    q75      = quantile(std_estimate, 0.75,  na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  select(-any_of(c("b_global", "se_global"))) %>%
  left_join(select(global_paths, path_id, b_global, se_global), by = "path_id")
```

```{r fig-h2-marginal}
#| label: fig-h2-marginal
#| fig-cap: >
#|   Marginal distribution of standardised β-products across all rolling
#|   windows (upper row per pathway), compared with the global (full
#|   time-series) piecewise SEM (lower row, nudged down by 0.3 units for
#|   readability). Upper row: open circle = rolling-window median; thick bar =
#|   IQR (25th–75th percentile); thin bar = 95% range (2.5th–97.5th
#|   percentile). Lower row: open diamond = global-model β-product; horizontal
#|   bar = ± 1 SE (first-order product approximation; see global_paths for
#|   caveats). A global estimate outside the rolling-window IQR suggests that
#|   temporal extremes pull the full-series fit away from the typical
#|   within-season relationship. Sign convention: A = 0, C = 1 — negative
#|   values indicate a stronger effect in the treed system.
#| fig-width: 7
#| fig-height: 5.5

y_order <- rev(levels(h2_ts$label))

# legend_dummy: two phantom rows at coordinates that fall outside any axis range
# (x = NA). They are never rendered as visible points but force ggplot's legend
# machinery to emit one entry per shape type. Without this layer, shapes mapped
# inside aes() across multiple geom_point() calls would not appear in the legend.
legend_dummy <- tibble(
  x     = c(NA,  NA),
  y     = factor(NA, levels = levels(h2_marginal$label)),
  shape = c("median", "global"),
  label = c("Rolling median (IQR / 95%)", "Global model (\u00b1 1 SE)")
)

shape_values <- c(median = 21, global = 23)

p_marginal <- ggplot(h2_marginal, aes(colour = path_id)) +

  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey55", linewidth = 0.35) +

  # ── Rolling-window distribution (upper row, y = label) ────────────────────
  # Thin bar: 95% range — captures rare seasonal extremes
  geom_linerange(aes(x = median_b, y = label,
                     xmin = q025, xmax = q975),
                 linewidth = 0.5, alpha = 0.30) +
  # Thick bar: IQR — the typical window-to-window spread
  geom_linerange(aes(x = median_b, y = label,
                     xmin = q25, xmax = q75),
                 linewidth = 2.0, alpha = 0.60) +
  # Open circle: median — robust central tendency across windows
  geom_point(aes(x = median_b, y = label, shape = "median"),
             size = 2.6, fill = "white", stroke = 1.2) +

  # ── Global model (lower row, nudged down by 0.3 units) ────────────────────
  # Nudging places the global estimate below its rolling-window counterpart so
  # the two rows are visually separable without adding a second y-axis or facet.
  # Thin bar: ± 1 SE of the β-product (product of individual path SEs)
  geom_linerange(aes(x = b_global, y = label,
                     xmin = b_global - se_global,
                     xmax = b_global + se_global),
                 position = position_nudge(y = -0.3),
                 linewidth = 0.5, alpha = 0.30) +
  # Open diamond: global-model β-product
  geom_point(aes(x = b_global, y = label, shape = "global"),
             position = position_nudge(y = -0.3),
             size = 3, fill = "white", stroke = 0.8,
             na.rm = TRUE) +

  # Phantom layer — legend entries only, no rendered points (x = NA)
  geom_point(data = legend_dummy,
             aes(x = x, y = y, shape = shape),
             colour = "grey30", fill = "white",
             size = 2.6, stroke = 1.1, na.rm = TRUE) +

  scale_shape_manual(
    name   = NULL,
    values = shape_values,
    labels = c(median = "Rolling median (IQR / 95%)",
               global = "Global model (\u00b1 1 SE)"),
    guide  = guide_legend(
      override.aes = list(
        fill   = c("white", "white"),
        colour = "grey30",
        size   = c(2.6, 2.6)
      )
    )
  ) +
  # Pathway colours encode the causal route (defined in path_meta); the colour
  # scale is hidden because it duplicates the y-axis label information.
  scale_colour_manual(values = path_colours, guide = "none") +
  scale_y_discrete(limits = y_order) +

  labs(x = "Standardised \u03b2", y = NULL) +
  theme_sem(legend_pos = "bottom") +
  theme(
    axis.text.y     = element_text(size = 6.5, hjust = 1),
    legend.text     = element_text(size = 7),
    legend.key.size = unit(14, "pt"),
    plot.margin     = margin(5, 8, 5, 5)
  )

p_marginal

tiff("../output/Fig2_H2_marginal.tiff",
     width = 12, height = 10, units = "cm", res = 300)
print(p_marginal)
invisible(dev.off())
```

### Temporal variation of rolling-window pathway estimates

```{r fig-h2-temporal}
#| fig-cap: >
#|   LOESS-smoothed time series of standardised β-products per
#|   causal pathway (community — all_taxa; span = 0.20). Each raw value is
#|   the product of standardised path coefficients from one rolling-window
#|   piecewise SEM; values are winsorised at the 2.5th–97.5th percentile
#|   before smoothing. Line colour and type encode the pathway: solid = direct
#|   or composite (total indirect / total effect); dashed = two-step indirect;
#|   dotted = three-step indirect. Season bands: Spring = Mar–May, etc.
#|   Sign convention: A = 0, C = 1 — negative = stronger effect in treed system.
#|   Peaks and troughs that align across pathways suggest a shared seasonal
#|   driver; diverging patterns suggest pathway-specific dynamics.
#| fig-width: 13
#| fig-height: 5.5

p_time <- ggplot(h2_smoothed,
                 aes(x        = date_start, y = smoothed,
                     colour   = path_id,    linetype = path_id)) +
  make_season_layers(date_range_h2,
                     y_label_pos = max(h2_smoothed$smoothed, na.rm = TRUE)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "grey55", linewidth = 0.35) +
  geom_line(linewidth = 0.65, alpha = 0.85) +
  scale_colour_manual( values = path_colours, labels = path_labels,
                       name = "Pathway") +
  scale_linetype_manual(values = path_ltypes,  labels = path_labels,
                        name = "Pathway") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months",
               expand = expansion(mult = 0.01)) +
  labs(
    x        = NULL,
    y        = "Standardised \u03b2 (LOESS-smoothed)",
    title    = "Temporal fluctuations of causal pathway estimates",
    subtitle = paste0(
      "LOESS smoother (span = 0.20) \u2022 community (all_taxa) \u2022 ",
      nrow(filter(h2_ts, path_id == "direct")), " rolling windows"
    )
  ) +
  theme_sem(legend_pos = "right") +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 6.5),
    legend.text      = element_text(size = 6.5),
    legend.title     = element_text(size = 7, face = "bold"),
    legend.key.width = unit(18, "pt"),
    plot.margin      = margin(5, 5, 5, 2)
  )

p_time

tiff("../output/Fig3_H2_temporal.tiff",
     width = 20, height = 9, units = "cm", res = 300)
print(p_time)
invisible(dev.off())
```
