---
title: |
  Results and visualisation
author: "Emma Belaud"
format:
  html:
    css: css/styles.css
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
    number-sections: true
execute:
  warning: false
  message: false
  echo: false
---

::: {.callout-note appearance="simple"}
## Purpose of this document

This is the **final step** of the SEM pipeline, following:

- `1_database_edition.qmd` — assembles the SEM database
- `2_SEM_diagnostic.qmd` — selects window duration, transformation, and valid windows
- `3_SEM_modelisation.qmd` — fits all models and writes result databases

**No modelling is performed here.** Every number in every figure comes directly
from pre-computed CSV files. Re-running this script is fast and safe — it never
touches a model.

| Input file | Contents |
|---|---|
| `H1_adjusted_values.csv` | Per-observation null-model residuals with date, land-use label, season |
| `H1_model_summary.csv` | β, SE, t, p, Hedge's g, R² from the two-step H1 model |
| `DAG_coefficients.csv` | Path coefficients from the full-dataset piecewise SEM |
| `DAG_model_fit.csv` | Fisher's C, AIC for the full-dataset piecewise SEM |
| `SEM_results_database.csv` | Rolling-window path coefficients for all taxa |
| `indirect_effects.csv` | Rolling-window indirect effect products for all taxa |
:::

---

# Setup

```{r setup}
# ══════════════════════════════════════════════════════════════════════════════
# SETUP
#
# Loads packages, reads shared pipeline metadata (microclimate variable names),
# and defines all visual constants and helper functions used throughout.
#
# Design philosophy
# ─────────────────
# All aesthetic choices — colours, linetypes, labels, font sizes — are defined
# once in this block and referenced by name everywhere else. Changing a value
# here propagates consistently to every figure without manual search-and-replace.
# ══════════════════════════════════════════════════════════════════════════════

invisible(Sys.setlocale("LC_TIME", "C"))    # Force English month names in plots
dir.create("../output", showWarnings = FALSE)

libs <- c("tidyverse", "lubridate", "patchwork",
          "kableExtra", "colorspace", "nlme")
invisible(lapply(libs, require, character.only = TRUE))


# ── Microclimate variable names ───────────────────────────────────────────────
# Written by 1_database_edition.qmd after PCA-based index selection.
# mc1_pred and mc2_pred are the two most informative, mutually orthogonal
# microclimate indices (e.g. "temp_soil_amplitude", "hum_soil_amplitude").
# They are used as axis labels and in pathway name strings throughout.
microclimate_index <- readLines("../output/microclimate_index.txt")
mc1_pred <- microclimate_index[1]   # First retained microclimate index (PC1)
mc2_pred <- microclimate_index[2]   # Second retained microclimate index (PC2)


# ── Land-use colour palette ───────────────────────────────────────────────────
# Applied to every figure that distinguishes land-use type.
# Dark olive-green = treed plots (A); warm orange = cultivated plots (C).
pos_colours <- c("Treed (A)" = "#717F55", "Cultivated (C)" = "#FF994E")


# ── Causal pathway metadata ───────────────────────────────────────────────────
# Single source of truth for all pathway display properties.
#
# Columns
# -------
#   path_id  : machine key used as join key in all downstream code
#   label    : human-readable string used in legends and axis labels
#   colour   : warm tones for root-mediated paths; cool/neutral for microclimate
#   ltype    : encodes causal depth
#              "solid"  → direct effect or composite (total indirect / total effect)
#              "dashed" → two-step indirect  (lu → X → fauna)
#              "dotted" → three-step indirect (lu → X → root → fauna)
path_meta <- tribble(
  ~path_id,         ~label,                                                 ~colour,   ~ltype,
  "direct",         "Direct (lu\u2192fauna)",                               "#F9BB05", "solid",
  "via_mc1",        paste0("lu\u2192", mc1_pred, "\u2192fauna"),            "#BCD094", "solid",
  "via_mc2",        paste0("lu\u2192", mc2_pred, "\u2192fauna"),            "#717F55", "solid",
  "via_root",       "lu\u2192root\u2192fauna",                              "#C45300", "solid",
  "via_mc1_root",   paste0("lu\u2192", mc1_pred, "\u2192root\u2192fauna"),  "#FF7E1F", "solid",
  "via_mc2_root",   paste0("lu\u2192", mc2_pred, "\u2192root\u2192fauna"),  "#FF994E", "solid",
  "total_indirect", "Total indirect",                                        "#999999", "solid",
  "total_effect",   "Total effect",                                          "#444444", "solid"
)

# Named vectors for ggplot scale_*_manual() calls (names = path_id values)
path_colours <- setNames(path_meta$colour, path_meta$path_id)
path_ltypes  <- setNames(path_meta$ltype,  path_meta$path_id)
path_labels  <- setNames(path_meta$label,  path_meta$path_id)


# ── Season background bands ───────────────────────────────────────────────────
# Alternating light-grey rectangles mark the four meteorological seasons.
# Boundaries: Spring = Mar–May | Summer = Jun–Aug | Autumn = Sep–Nov | Winter = Dec–Feb.
season_order <- c("Spring", "Summer", "Autumn", "Winter")
season_fills <- c(Spring = "#f9f9f9", Summer = "#eeeeee",
                  Autumn = "#f9f9f9", Winter = "#eeeeee")


# ── Agricultural and irrigation event landmarks ───────────────────────────────
# Used as vertical reference lines in all time-series figures.
# Red dotted = crop management events; green dotted = irrigation events.
interventions <- data.frame(
  date = as.Date(c(
    "2024-07-08", "2024-07-15",
    "2024-11-22", "2024-11-25",
    "2025-03-14", "2025-07-10"
  )),
  event = c(
    "Wheat Harvest", "Sorghum Sown",
    "Sorghum Mown",  "Tilled",
    "Chickpea Sown", "Chickpea Harvest"
  )
)

irrigation <- data.frame(
  date = as.Date(c("2024-07-17", "2024-07-22", "2024-08-12"))
)


# ── Helper: season annotation layers ─────────────────────────────────────────
#' Build season background rectangles and italic labels for a time-series plot.
#'
#' Returns a list of ggplot layers (annotate rectangles + geom_text labels)
#' that can be added with `+ make_season_layers(...)` to any date-scaled plot.
#'
#' @param date_range  Length-2 Date vector matching the plot x-axis range.
#' @param y_label_pos Vertical position of season name labels. `Inf` = top.
#' @param label_size  ggplot font size for season labels (default 3.5).
#' @return A list of ggplot annotation / geom layers.
make_season_layers <- function(date_range, y_label_pos = Inf, label_size = 3.5) {
  bands <- tibble(
    season  = names(season_fills),
    m_start = c(3L, 6L, 9L, 12L),
    m_end   = c(6L, 9L, 12L,  3L),
    fill    = unname(season_fills)
  ) %>%
    crossing(year = seq(year(date_range[1]), year(date_range[2]))) %>%
    mutate(
      xmin = as.Date(paste(year, m_start, "01", sep = "-")),
      xmax = as.Date(paste(
        if_else(m_end < m_start, year + 1L, year),
        m_end, "01", sep = "-"
      )),
      xmid = xmin + (xmax - xmin) / 2
    ) %>%
    filter(xmin < date_range[2], xmax > date_range[1]) %>%
    mutate(xmin = pmax(xmin, date_range[1]),
           xmax = pmin(xmax, date_range[2]))

  rect_layers <- pmap(
    bands %>% select(xmin, xmax, fill),
    \(xmin, xmax, fill) annotate(
      "rect", xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf,
      fill = fill, alpha = 1
    )
  )

  label_layer <- geom_text(
    data        = bands,
    aes(x = xmid, y = y_label_pos, label = season),
    inherit.aes = FALSE,
    vjust = 1.6, size = label_size,
    colour = "grey55", fontface = "italic", family = "sans"
  )

  c(rect_layers, list(label_layer))
}


# ── Base ggplot theme ─────────────────────────────────────────────────────────
#' Consistent theme for all SEM figures.
#'
#' Built on top of theme_classic() with a light panel border, subtle grid
#' lines, and no axis lines (replaced by panel.border). All font sizes scale
#' from `base_size` so that the whole figure can be rescaled by changing one
#' argument.
#'
#' @param base_size  Base font size in pt. Default 11 (increased for legibility
#'                   following reviewer request).
#' @param legend_pos Passed to `legend.position`. Default "none".
theme_sem <- function(base_size = 11, legend_pos = "none") {
  theme_classic(base_size = base_size) %+replace%
    theme(
      panel.border      = element_rect(colour = "grey20", fill = NA,
                                       linewidth = 0.4),
      panel.grid.major  = element_line(colour = "grey92", linewidth = 0.3),
      panel.grid.minor  = element_blank(),
      panel.background  = element_rect(fill = "white", colour = NA),
      axis.line         = element_blank(),
      axis.ticks        = element_line(colour = "grey40", linewidth = 0.35),
      axis.ticks.length = unit(3, "pt"),
      axis.title        = element_text(size = base_size,     colour = "grey20"),
      axis.text         = element_text(size = base_size - 1, colour = "grey30"),
      strip.text        = element_text(size = base_size,     face = "bold",
                                       colour = "grey15"),
      strip.background  = element_rect(fill = "grey96", colour = "grey96",
                                       linewidth = 0.3),
      plot.title        = element_text(size = base_size + 2, face = "bold",
                                       hjust = 0, colour = "grey10",
                                       margin = margin(b = 4)),
      plot.subtitle     = element_text(size = base_size - 1, colour = "grey45",
                                       hjust = 0, margin = margin(b = 6)),
      plot.caption      = element_text(size = base_size - 2, colour = "grey55",
                                       hjust = 0, margin = margin(t = 5)),
      legend.position   = legend_pos,
      legend.background = element_blank(),
      legend.key        = element_blank(),
      legend.title      = element_text(size = base_size,     face = "bold",
                                       colour = "grey20"),
      legend.text       = element_text(size = base_size - 1, colour = "grey30"),
      legend.key.size   = unit(12, "pt"),
      legend.spacing.x  = unit(4, "pt"),
      plot.margin       = margin(8, 8, 8, 8)
    )
}


# ── Load pre-computed result databases ────────────────────────────────────────
# All modelling was performed upstream in 3_SEM_modelisation.qmd.
# This script reads outputs only — no model is fitted here.

# H1: null-model residuals (one row per observation × scanner)
h1_adj <- read.csv("../output/H1_adjusted_values.csv",
                   na.strings = c("", "NA")) %>%
  mutate(
    position_label = factor(position_label,
                            levels = c("Treed (A)", "Cultivated (C)")),
    season         = factor(season, levels = season_order)
  )

# H1 model summary: β, SE, t, p, Hedge's g, R² from the two-step linear model
h1_summary <- read.csv("../output/H1_model_summary.csv",
                       na.strings = c("", "NA"))

# Full-dataset piecewise SEM: directional path coefficients
dag_table <- read.csv("../output/DAG_coefficients.csv",
                      na.strings = c("", "NA"))

# Full-dataset piecewise SEM: global fit statistics (Fisher's C, AIC)
dag_fit <- read.csv("../output/DAG_model_fit.csv",
                    na.strings = c("", "NA"))

# Rolling-window SEM: one row per (taxon × window × path)
sem_results <- read.csv("../output/SEM_results_database.csv",
                        na.strings = c("", "NA")) %>%
  mutate(season = factor(season, levels = season_order))

# Rolling-window indirect effects: β-products for all causal routes
indirect_effects <- read.csv("../output/indirect_effects.csv",
                             na.strings = c("", "NA"))

# Faunal transformation used during modelling (e.g. "log", "sqrt_plus_one")
# — stored by 3_SEM_modelisation.qmd and used in axis labels where relevant
transfo_fauna <- h1_summary$transformation[1]
```

---

# H1 — Direct effect of land-use type on faunal activity

> **Land-use type directly structures faunal activity, with treed habitats
> sustaining lower but more temporally stable activity than cultivated zones.**

## Two-step modelling approach

Figures in this section show **null-model residuals** — faunal activity with
the spatial grouping structure removed.

**Step 1 — Remove spatial structure.** A linear mixed model with only an
intercept is fitted on the full dataset. Random effects for orientation nested
within depth absorb all variation attributable to scanner placement, with an
AR(1) correlation structure for temporal autocorrelation:

$$y^* \sim 1 + (1 \mid \text{orientation} / \text{depth}) + \text{AR}(1)$$

**Step 2 — Test the land-use effect on residuals.**

$$\hat{\varepsilon}_{\text{Step 1}} \sim \text{land\_use}$$

β = mean(residuals at C) − mean(residuals at A). A positive β means faunal
activity is higher in cultivated plots after removing orientation and depth
effects.

```{r fig-h1-combined}
#| label: fig-h1-combined
#| fig-cap: >
#|   **Direct effect of land-use type on soil invertebrate activity.**
#|   (A) Distribution of random-effect-adjusted fauna values per land-use type.
#|   Values represent observed fauna with scanner orientation and depth effects
#|   removed, preserving the intercept so that the Y axis is in fauna units
#|   (log-transformed scale). (B) Temporal dynamics of random-effect-adjusted
#|   fauna values with LOESS smoothers (span = 0.18). Green = treed (A);
#|   orange = cultivated (C). Shaded bands = meteorological seasons. Dotted red
#|   vertical lines = agricultural interventions (wheat harvest, sorghum sowing,
#|   mowing, tillage, chickpea sowing). Dotted green vertical lines = irrigation
#|   events.
#| fig-width: 14
#| fig-height: 6

# ── Shared y-axis limits ──────────────────────────────────────────────────────
# Derived from the raw data so both panels use exactly the same scale.
# Rounded to the nearest 0.1 for clean tick positions.
shared_ylim <- c(
  floor(  min(h1_adj$adj_value, na.rm = TRUE) * 10) / 10,
  ceiling(max(h1_adj$adj_value, na.rm = TRUE) * 10) / 10
)


# ── Left panel: boxplot ───────────────────────────────────────────────────────
# Jitter (very transparent) shows the full data density; boxplot summarises
# the distribution. The right border is suppressed so the two panels join flush.
p_box <- ggplot(
  h1_adj,
  aes(x = position_label, y = adj_value,
      fill = position_label, colour = position_label)
) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.35,
             linetype = "dashed") +
  geom_jitter(alpha = 0.05, size = 0.4, shape = 20) +
  geom_boxplot(width = 0.5, outliers = FALSE,
               linewidth = 0.9, fatten = 1.2, alpha = 0.5) +
  scale_colour_manual(values = pos_colours) +
  scale_fill_manual(values   = pos_colours) +
  scale_x_discrete(labels = c("Treed", "Cultivated")) +
  coord_cartesian(ylim = shared_ylim) +
  labs(x = NULL, y = "Soil invertebrate activity (adjusted)") +
  guides(fill = "none", colour = "none") +
  theme_sem() +
  theme(
    # Bold x tick labels to distinguish the two land-use categories clearly
    axis.text.x      = element_text(size = 11, face = "bold"),
    panel.border     = element_blank(),
    axis.line.y.left = element_line(colour = "grey20", linewidth = 0.35),
    axis.line.x      = element_line(colour = "grey20", linewidth = 0.35),
    plot.margin      = margin(t = 5, r = 0, b = 5, l = 5, unit = "pt")
  )


# ── Right panel: LOESS time series ───────────────────────────────────────────
# Left border and y-axis are suppressed — they are carried by the left panel.
# Season bands are added first so all geoms are drawn on top.
p_temp <- ggplot(
  h1_adj %>% mutate(date = as.Date(date)),
  aes(x = date, colour = position_label, fill = position_label)
) +
  make_season_layers(range(as.Date(h1_adj$date), na.rm = TRUE),
                     y_label_pos = shared_ylim[2]) +
  geom_vline(data = interventions, aes(xintercept = date),
             colour = "darkred",   linetype = "dotted", alpha = 0.7) +
  geom_vline(data = irrigation,    aes(xintercept = date),
             colour = "darkgreen", linetype = "dotted", alpha = 0.7) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.35,
             linetype = "dashed") +
  # Raw observations: very transparent to show local density without clutter
  geom_point(aes(y = adj_value), size = 0.6, alpha = 0.15, shape = 16) +
  # LOESS smoother: span = 0.18 gives seasonal resolution without over-smoothing
  geom_smooth(aes(y = adj_value), method = "loess", formula = y ~ x,
              span = 0.18, linewidth = 0.85, se = FALSE) +
  scale_colour_manual(values = pos_colours, name = "Land-use type",
                      labels = c("Treed (A)", "Cultivated (C)")) +
  scale_fill_manual(values   = pos_colours, name = "Land-use type",
                    labels   = c("Treed (A)", "Cultivated (C)")) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months",
               expand = expansion(mult = 0.01)) +
  coord_cartesian(ylim = shared_ylim) +
  labs(x = NULL, y = NULL) +
  guides(fill = "none", colour = "none") +
  theme_sem(legend_pos = "bottom") +
  theme(
    axis.text.x       = element_text(angle = 35, hjust = 1, size = 10),
    legend.direction  = "horizontal",
    legend.text       = element_text(size = 11),
    axis.title.y      = element_blank(),
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    panel.border      = element_blank(),
    axis.line.y.right = element_line(colour = "grey20", linewidth = 0.35),
    axis.line.x       = element_line(colour = "grey20", linewidth = 0.35),
    plot.margin       = margin(t = 5, r = 5, b = 5, l = 0, unit = "pt")
  )


# ── Combine with patchwork ────────────────────────────────────────────────────
# widths = c(1, 3): temporal panel is 3× wider than the boxplot.
# guides = "collect" merges the two colour legends into one block below.
h1_combined <- (p_box + p_temp) +
  plot_layout(widths = c(1, 3), guides = "collect") &
  theme(legend.position = "bottom", legend.direction = "horizontal")

h1_combined <- h1_combined + plot_annotation(tag_levels = "A")
h1_combined

ggsave("../output/Fig1_H1_combined.tiff", h1_combined,
       width = 24, height = 12, units = "cm", dpi = 300, compression = "lzw")
```

---

# H2 — Indirect effects through contrasting causal pathways

> **Land-use effects are mediated through contrasting causal pathways —
> agricultural management drives faunal activity primarily through altered
> microclimate, whereas treed habitats sustain faunal activity more
> consistently through root resource supply.**

## Pathway algebra

The SEM estimates one standardised β per path per rolling window. An
**indirect effect** is the product of all β values along a causal route.
Because all continuous variables are z-scored within each window, these
products are dimensionless and directly comparable across windows and pathways.

| Path | Expression |
|---|---|
| Direct | $\beta_9$ |
| lu → mc₁ → fauna | $\beta_1 \times \beta_6$ |
| lu → mc₂ → fauna | $\beta_2 \times \beta_7$ |
| lu → root → fauna | $\beta_5 \times \beta_8$ |
| lu → mc₁ → root → fauna | $\beta_1 \times \beta_3 \times \beta_8$ |
| lu → mc₂ → root → fauna | $\beta_2 \times \beta_4 \times \beta_8$ |
| **Total indirect** | sum of the five routes above |
| **Total effect** | direct + total indirect |

**Sign convention:** A = 0, C = 1. A negative β-product means the effect
through that pathway is stronger in the treed system.

## Table 1 — Full-dataset piecewise SEM path coefficients

```{r tbl-h2-dag}
#| tbl-cap: >
#|   Path coefficients from the full-dataset piecewise SEM
#|   (all_taxa, complete time series, all depths). All continuous variables
#|   were z-scored after the calibration-selected transformation. β =
#|   unstandardised coefficient; Std β = standardised coefficient directly
#|   comparable to rolling-window values in Figures 2–3. R²m = marginal R²
#|   (fixed effects only); R²c = conditional R² (fixed + random effects).
#|   Fisher's C tests global model fit: p > 0.05 indicates that the
#|   conditional independence claims implied by the DAG are consistent
#|   with the data.

# Build the model-fit footnote string.
# A saturated DAG has no d-separation claims, so Fisher's C is NA — expected.
fit_line <- with(dag_fit, paste0(
  "Model fit \u2014 Fisher\u2019s C = ",
  if (is.na(fishers_C)) "NA" else round(fishers_C, 2),
  " (df = ", if (is.na(df)) "NA" else df, ", p ",
  if (is.na(p_value))
    "= NA (saturated model \u2014 no independence claims to test)"
  else if (p_value < 0.001) "< 0.001"
  else paste0("= ", round(p_value, 3)),
  ") \u2502 AIC = ",
  if (is.na(AIC)) "NA" else round(AIC, 1)
))

dag_table %>%
  # Exclude variance–covariance rows (~~): not directional paths
  filter(!str_detect(response, "~~")) %>%
  mutate(
    # piecewiseSEM may store numeric columns as character — coerce before formatting
    across(c(estimate, std_estimate, se, t_value, p_value,
             r2_marginal, r2_conditional),
           ~ suppressWarnings(as.numeric(.))),
    across(c(estimate, std_estimate, se), ~ sprintf("%.3f", .)),
    t_fmt   = if_else(is.na(t_value), "\u2014", sprintf("%.2f", t_value)),
    p_fmt   = if_else(p_value < 0.001, "< 0.001", sprintf("%.3f", p_value)),
    r2_marg = sprintf("%.3f", r2_marginal),
    r2_cond = sprintf("%.3f", r2_conditional)
  ) %>%
  select(
    "Equation"   = equation,
    "Predictor"  = predictor,
    "\u03b2"     = estimate,
    "Std \u03b2" = std_estimate,
    "SE"         = se,
    "t"          = t_fmt,
    "p"          = p_fmt,
    "sig"        = sig,
    "R\u00b2m"   = r2_marg,
    "R\u00b2c"   = r2_cond
  ) %>%
  kbl(format = "html", escape = FALSE,
      align  = c("l", "l", "r", "r", "r", "r", "r", "l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("condensed", "hover"),
                full_width = FALSE, font_size = 13) %>%
  row_spec(0, bold = TRUE, background = "#e8e8e8") %>%
  # Merge repeated equation and R² cells for readability
  collapse_rows(columns = c(1, 9, 10), valign = "middle") %>%
  footnote(
    general = paste0(
      fit_line,
      "<br>Sig. codes: *** p < 0.001 &nbsp;** p < 0.01 &nbsp;* p < 0.05 &nbsp;. p < 0.10"
    ),
    general_title = "", escape = FALSE, footnote_as_chunk = TRUE
  )
```

## Rolling-window vs. full-dataset pathway estimates

```{r h2-prep}
# ══════════════════════════════════════════════════════════════════════════════
# PREPARE H2 TIME SERIES
#
# Two data sources are assembled:
#   direct_ts   — the lu→fauna β extracted from sem_results (one row per window)
#   indirect_ts — β-products from indirect_effects, pivoted to long format
#
# Both are restricted to all_taxa (community level), then joined with path_meta
# to attach display labels, colours, and linetypes.
#
# Winsorisation at 2.5–97.5% per pathway before LOESS smoothing reduces the
# leverage of outlier windows (e.g. sparse fauna in winter) without discarding
# those windows from the raw data layer.
# ══════════════════════════════════════════════════════════════════════════════

direct_ts <- sem_results %>%
  filter(taxon == "all_taxa", predictor == "land_use", response == "fauna") %>%
  transmute(path_id = "direct", date_start, season, std_estimate)

indirect_ts <- indirect_effects %>%
  filter(taxon == "all_taxa") %>%
  pivot_longer(
    c(via_mc1, via_mc2, via_root, via_mc1_root, via_mc2_root,
      total_indirect, total_effect),
    names_to  = "path_id",
    values_to = "std_estimate"
  ) %>%
  filter(!is.na(std_estimate)) %>%
  # Derive season from date_start rather than relying on the saved column,
  # which may be absent in older pipeline runs
  mutate(
    month  = month(date_start),
    season = factor(case_when(
      month %in% 3:5  ~ "Spring",
      month %in% 6:8  ~ "Summer",
      month %in% 9:11 ~ "Autumn",
      TRUE            ~ "Winter"
    ), levels = season_order)
  ) %>%
  select(path_id, date_start, season, std_estimate)

h2_ts <- bind_rows(direct_ts, indirect_ts) %>%
  left_join(path_meta, by = "path_id") %>%
  mutate(
    path_id = factor(path_id, levels = path_meta$path_id),
    label   = factor(label,   levels = path_meta$label),
    season  = factor(season,  levels = season_order)
  ) %>%
  # Winsorise per pathway to reduce leverage of extreme windows on the smoother
  group_by(path_id) %>%
  mutate(
    lo           = quantile(std_estimate, 0.025, na.rm = TRUE),
    hi           = quantile(std_estimate, 0.975, na.rm = TRUE),
    std_estimate = pmax(lo, pmin(hi, std_estimate))
  ) %>%
  ungroup()


# ── LOESS smoother helper ─────────────────────────────────────────────────────
#' Fit a LOESS smoother to one pathway's time series.
#'
#' Used inside group_modify(). Falls back to the raw estimates when fewer than
#' 5 valid observations are available or when LOESS fails.
#'
#' @param df    Data frame with columns `date_start` (Date) and `std_estimate`.
#' @param span  LOESS bandwidth (default 0.30).
#' @return `df` with an added `smoothed` column.
smooth_path <- function(df, span = 0.30) {
  df <- df %>% filter(!is.na(std_estimate)) %>% arrange(date_start)
  if (nrow(df) < 5) return(mutate(df, smoothed = std_estimate))
  fit <- tryCatch(
    loess(std_estimate ~ as.numeric(date_start), data = df, span = span,
          control = loess.control(surface = "direct")),  # avoids edge artefacts
    error = \(e) NULL
  )
  df$smoothed <- if (!is.null(fit)) predict(fit) else df$std_estimate
  df
}

h2_smoothed <- h2_ts %>%
  mutate(date_start = as.Date(date_start)) %>%
  group_by(path_id) %>%
  group_modify(~ smooth_path(.x)) %>%
  ungroup()

date_range_h2 <- range(h2_smoothed$date_start)

# Shared y limits: derived from both raw and smoothed values simultaneously
shared_ylim_h2 <- c(
  floor(  min(c(h2_ts$std_estimate, h2_smoothed$smoothed), na.rm = TRUE) * 10) / 10,
  ceiling(max(c(h2_ts$std_estimate, h2_smoothed$smoothed), na.rm = TRUE) * 10) / 10
)

# Short pathway labels for the marginal panel x-axis (multi-line allowed)
path_labels_short <- setNames(
  c("Direct", "mc1\u2192fauna", "mc2\u2192fauna", "root\u2192fauna",
    "mc1\u2192root\n\u2192fauna", "mc2\u2192root\n\u2192fauna",
    "Total\nindirect", "Total\neffect"),
  path_meta$label   # names must match the full label strings used as breaks
)
```

```{r h2-global-coef}
# ══════════════════════════════════════════════════════════════════════════════
# EXTRACT GLOBAL-MODEL PATHWAY ESTIMATES
#
# global_coef: one row per directional path with standardised β (b) and SE.
# get_beta() / get_se(): single-value extractors that return NA when a path is
# absent, so downstream β-products propagate NA gracefully.
#
# global_paths: β-products and their approximate SEs for the full-dataset SEM.
# SEs use a first-order product approximation (product of individual SEs),
# which is conservative and appropriate for error bars only — not formal
# inference on products.
# ══════════════════════════════════════════════════════════════════════════════

global_coef <- dag_table %>%
  filter(!str_detect(response, "~~")) %>%
  mutate(std_estimate = suppressWarnings(as.numeric(std_estimate)),
         se           = suppressWarnings(as.numeric(se))) %>%
  select(response, predictor, b = std_estimate, se)

get_beta <- function(resp, pred) {
  val <- global_coef$b[global_coef$response == resp &
                        global_coef$predictor == pred]
  if (length(val) == 0) NA_real_ else val[1]
}
get_se <- function(resp, pred) {
  val <- global_coef$se[global_coef$response == resp &
                         global_coef$predictor == pred]
  if (length(val) == 0) NA_real_ else val[1]
}

# Individual path coefficients
lu_mc1  <- get_beta("microclimate_1", "land_use");  se_lu_mc1  <- get_se("microclimate_1", "land_use")
lu_mc2  <- get_beta("microclimate_2", "land_use");  se_lu_mc2  <- get_se("microclimate_2", "land_use")
mc1_root<- get_beta("root", "microclimate_1");      se_mc1_root<- get_se("root", "microclimate_1")
mc2_root<- get_beta("root", "microclimate_2");      se_mc2_root<- get_se("root", "microclimate_2")
lu_root <- get_beta("root", "land_use");            se_lu_root <- get_se("root", "land_use")
mc1_fau <- get_beta("fauna", "microclimate_1");     se_mc1_fau <- get_se("fauna", "microclimate_1")
mc2_fau <- get_beta("fauna", "microclimate_2");     se_mc2_fau <- get_se("fauna", "microclimate_2")
root_fau<- get_beta("fauna", "root");               se_root_fau<- get_se("fauna", "root")
lu_fau  <- get_beta("fauna", "land_use");           se_lu_fau  <- get_se("fauna", "land_use")

# β-products for each pathway in the full-dataset model
global_paths <- tribble(
  ~path_id,         ~b_global,                                 ~se_global,
  "direct",          lu_fau,                                    se_lu_fau,
  "via_mc1",         lu_mc1  * mc1_fau,                         se_lu_mc1  * se_mc1_fau,
  "via_mc2",         lu_mc2  * mc2_fau,                         se_lu_mc2  * se_mc2_fau,
  "via_root",        lu_root * root_fau,                        se_lu_root * se_root_fau,
  "via_mc1_root",    lu_mc1  * mc1_root * root_fau,             se_lu_mc1  * se_mc1_root * se_root_fau,
  "via_mc2_root",    lu_mc2  * mc2_root * root_fau,             se_lu_mc2  * se_mc2_root * se_root_fau,
  "total_indirect",  (lu_mc1 * mc1_fau) + (lu_mc2 * mc2_fau),  (se_lu_mc1 * se_mc1_fau) + (se_lu_mc2 * se_mc2_fau),
  "total_effect",    lu_fau  + (lu_mc1  * mc1_fau),             se_lu_fau  + (se_lu_mc1  * se_mc1_fau)
) %>%
  left_join(path_meta, by = "path_id") %>%
  mutate(path_id = factor(path_id, levels = path_meta$path_id),
         label   = factor(label,   levels = path_meta$label))


# ── Rolling-window marginal summaries per pathway ────────────────────────────
# Collapses all windows into one summary row per pathway:
#   median_b — robust central tendency across windows
#   sd_b     — temporal variability
#   q025/q975 — 2.5–97.5% range (after winsorising)
#   q25/q75  — IQR (typical window-to-window range)
h2_marginal <- h2_ts %>%
  group_by(path_id, label, colour, ltype) %>%
  summarise(
    median_b = median(std_estimate, na.rm = TRUE),
    sd_b     = sd(std_estimate,     na.rm = TRUE),
    q025     = quantile(std_estimate, 0.025, na.rm = TRUE),
    q975     = quantile(std_estimate, 0.975, na.rm = TRUE),
    q25      = quantile(std_estimate, 0.25,  na.rm = TRUE),
    q75      = quantile(std_estimate, 0.75,  na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  # Guard against duplicate columns on interactive re-runs
  select(-any_of(c("b_global", "se_global"))) %>%
  left_join(select(global_paths, path_id, b_global, se_global), by = "path_id")
```

## Figure 2

```{r fig-h2-combined}
#| label: fig-h2-combined
#| fig-cap: >
#|   **Causal pathway decomposition of the land-use effect on soil
#|   invertebrate activity.**
#|   (A) Marginal distribution of standardised β-products across all rolling
#|   windows per causal pathway. Thick bar = IQR; thin bar = 95% range (after
#|   winsorising); open circle = rolling-window median; filled circle =
#|   full time-series piecewise SEM estimate ± 1 SE.
#|   (B) LOESS-smoothed (span = 0.30) time series of standardised β-products.
#|   Shaded bands = meteorological seasons. Dotted red lines = agricultural
#|   interventions; dotted green lines = irrigation events. Sign convention:
#|   A = 0, C = 1 — negative values indicate a stronger effect in the treed
#|   system. Both panels share the same y-axis.
#| fig-width: 18
#| fig-height: 7

# Shared shape scale: circle = rolling median; filled circle = global estimate
shape_values <- c(median = 21, global = 21)

# Phantom data for the shape legend — coordinates are NA so nothing is drawn,
# but ggplot emits one legend entry per shape key
legend_dummy <- tibble(
  x     = c(NA_character_, NA_character_),
  y     = c(NA_real_,      NA_real_),
  shape = c("median",      "global")
)


# ── Left panel: marginal summary ─────────────────────────────────────────────
# One column per pathway. Thin bar = 95% range; thick bar = IQR;
# open circle = rolling median; filled circle = global SEM estimate.
# Right border suppressed so panels join flush.
p_marginal <- ggplot(h2_marginal, aes(colour = path_id, fill = path_id)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey55", linewidth = 0.4) +
  # 95% range (thin, transparent)
  geom_linerange(aes(x = label, ymin = q025, ymax = q975),
                 linewidth = 0.6, alpha = 0.30) +
  # IQR (thick, semi-transparent)
  geom_linerange(aes(x = label, ymin = q25, ymax = q75),
                 linewidth = 2.5, alpha = 0.60) +
  # Rolling-window median (open circle)
  geom_point(aes(x = label, y = median_b, shape = "median"),
             size = 2.8, fill = "white", stroke = 1.2) +
  # Global model ± 1 SE bar
  geom_linerange(aes(x = label,
                     ymin = b_global - se_global,
                     ymax = b_global + se_global),
                 linewidth = 0.6, alpha = 0.40, na.rm = TRUE) +
  # Global model estimate (filled circle, pathway colour)
  geom_point(aes(x = label, y = b_global, shape = "global", fill = path_id),
             size = 2.5, na.rm = TRUE) +
  # Phantom points for legend keys
  geom_point(data    = legend_dummy,
             mapping = aes(x = x, y = y, shape = shape),
             colour  = "grey30", fill = "white",
             size = 2.5, stroke = 1.0, na.rm = TRUE) +
  scale_shape_manual(
    name   = NULL,
    values = shape_values,
    breaks = c("median", "global"),
    labels = c(median = "Rolling median (IQR / 95%)",
               global = "Global model (\u00b1 1 SE)"),
    guide  = guide_legend(direction = "vertical",
      override.aes = list(fill   = c("white", "grey30"),
                          colour = "grey30",
                          size   = c(2.8, 2.5)))
  ) +
  scale_colour_manual(values = path_colours, guide = "none") +
  scale_fill_manual(values   = path_colours, guide = "none") +
  scale_x_discrete(limits = levels(h2_marginal$label),
                   labels  = path_labels_short) +
  coord_cartesian(ylim = shared_ylim_h2) +
  labs(x = NULL, y = "Standardised \u03b2") +
  theme_sem(legend_pos = "bottom") +
  theme(
    axis.text.x        = element_text(size = 9, angle = 40, hjust = 1,
                                      lineheight = 0.9),
    axis.text.y        = element_text(size = 10),
    axis.title.y       = element_text(size = 11),
    legend.text        = element_text(size = 10),
    legend.key.size    = unit(12, "pt"),
    panel.border       = element_blank(),
    axis.line.y.left   = element_line(colour = "grey20", linewidth = 0.35),
    axis.line.x.bottom = element_line(colour = "grey20", linewidth = 0.35),
    plot.margin        = margin(t = 5, r = 0, b = 5, l = 5, unit = "pt")
  )


# ── Right panel: LOESS temporal time series ───────────────────────────────────
# y-axis suppressed — carried entirely by the left panel.
# Left border and left plot margin zeroed to close the join flush.
p_time <- ggplot(
  h2_smoothed %>% mutate(date_start = as.Date(date_start)),
  aes(x = date_start, y = smoothed, colour = path_id, linetype = path_id)
) +
  make_season_layers(date_range_h2, y_label_pos = shared_ylim_h2[2] * 0.9) +
  geom_vline(data = interventions, aes(xintercept = date),
             colour = "darkred",   linetype = "dotted", alpha = 0.7) +
  geom_vline(data = irrigation,    aes(xintercept = date),
             colour = "darkgreen", linetype = "dotted", alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey55", linewidth = 0.4) +
  geom_line(linewidth = 0.75) +
  scale_colour_manual(values = path_colours, labels = path_labels,
                      name   = "Pathway") +
  scale_linetype_manual(values = path_ltypes, labels = path_labels,
                        name   = "Pathway") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months",
               expand = expansion(mult = 0.01)) +
  coord_cartesian(ylim = shared_ylim_h2) +
  labs(x = NULL, y = NULL) +
  theme_sem(legend_pos = "bottom") +
  theme(
    axis.text.x        = element_text(angle = 35, hjust = 1, size = 10),
    legend.text        = element_text(size = 9),
    legend.title       = element_text(size = 10, face = "bold"),
    legend.key.width   = unit(20, "pt"),
    legend.key.height  = unit(12, "pt"),
    axis.title.y       = element_blank(),
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    panel.border       = element_blank(),
    axis.line.y.right  = element_line(colour = "grey20", linewidth = 0.35),
    axis.line.x.bottom = element_line(colour = "grey20", linewidth = 0.35),
    plot.margin        = margin(t = 5, r = 5, b = 5, l = 0, unit = "pt")
  )


# ── Combine with patchwork ────────────────────────────────────────────────────
# guides = "keep" preserves each panel's own legend (shape below marginal;
# pathway legend below temporal). "collect" would merge them and break the
# spatial association with each panel.
h2_combined <- (p_marginal + p_time) +
  plot_layout(widths = c(1, 3), guides = "keep")
h2_combined <- h2_combined + plot_annotation(tag_levels = "A")
h2_combined

ggsave("../output/Fig2_H2_combined.tiff", h2_combined,
       width = 32, height = 16, units = "cm", dpi = 300, compression = "lzw")
```

---

# H3 — Land-use forcing compresses taxonomic response diversity

> **Land-use intensification compresses taxonomic response diversity by
> reducing the divergence among individual taxa in their sensitivity to
> the land-use contrast.**

## Analysis parameters

```{r h3-params}
# Pathways included in the H3 divergence analysis.
# "direct" and the five indirect routes + the two composites.
pathway_focus <- c(
  "direct", "total_indirect",
  "via_mc1", "via_mc2", "via_root",
  "via_mc1_root", "via_mc2_root"
)

eps           <- 1e-4    # Small positive guard against exact zeros in denominators
pathway_order <- pathway_focus   # Controls facet and legend ordering
```

## Data preparation

```{r h3-data}
# ══════════════════════════════════════════════════════════════════════════════
# H3 — ASSEMBLE PER-TAXON PER-WINDOW β ESTIMATES
#
# Two data sources are stacked:
#   h3_direct_win   — the lu→fauna β from sem_results (one row per window)
#   h3_indirect_win — β-products from indirect_effects, pivoted to long format
#
# Both are restricted to non-community taxa (taxon != "all_taxa").
# The community (all_taxa) β is extracted separately as comm_beta_win and used
# as the "forcing strength" covariate in the divergence analysis.
# ══════════════════════════════════════════════════════════════════════════════

h3_direct_win <- sem_results %>%
  filter(predictor == "land_use", response == "fauna") %>%
  transmute(taxon, date_start = as.Date(date_start),
            path_id = "direct", std_estimate)

h3_indirect_win <- indirect_effects %>%
  pivot_longer(
    cols      = c(via_mc1, via_mc2, via_root,
                  via_mc1_root, via_mc2_root,
                  total_indirect, total_effect),
    names_to  = "path_id", values_to = "std_estimate"
  ) %>%
  filter(!is.na(std_estimate), path_id %in% pathway_focus) %>%
  mutate(date_start = as.Date(date_start)) %>%
  select(taxon, path_id, date_start, std_estimate)

h3_win_raw <- bind_rows(h3_direct_win, h3_indirect_win) %>%
  filter(path_id %in% pathway_focus)

# Community-level β per (pathway × window): used as forcing-strength covariate
comm_beta_win <- h3_win_raw %>%
  filter(taxon == "all_taxa") %>%
  select(path_id, date_start, community_beta = std_estimate)

# Individual taxa only (all_taxa excluded from divergence computation)
h3_taxa_win <- h3_win_raw %>% filter(taxon != "all_taxa")
```

## Divergence computation

```{r h3-divergence}
# ══════════════════════════════════════════════════════════════════════════════
# H3 — ROSS ET AL. DIVERGENCE INDEX
#
# For a set of taxon β-values in a given (pathway × window):
#
#   D = max(0, range(β) − |max(β) + min(β)|) / range(β)
#
# Interpretation
# ──────────────
#   D = 1  → β-values span zero symmetrically (true response divergence:
#             some taxa positively affected, some negatively)
#   D = 0  → all taxa respond in the same direction (no divergence)
#   0 < D < 1 → intermediate divergence
#
# In the rolling-window SEM, each taxon's β is the slope of the activity–
# land_use relationship in that window, i.e. f′(land_use). D thus captures
# how spread across the sign boundary the taxon-specific land-use responses
# are within each temporal snapshot.
#
# IQR-fence outlier removal (1.5 × IQR) is applied per pathway on both the
# divergence values and the |community β| values before the regression step,
# to prevent rare extreme windows from dominating the GLS slope.
# ══════════════════════════════════════════════════════════════════════════════

compute_divergence <- function(df) {
  betas  <- df$std_estimate[!is.na(df$std_estimate)]
  n_taxa <- length(unique(df$taxon))
  if (length(betas) < 2) return(tibble(divergence = NA_real_, n_taxa = n_taxa))

  b_max <- max(betas)
  b_min <- min(betas)
  b_rng <- b_max - b_min

  if (b_rng == 0) return(tibble(divergence = 0, n_taxa = n_taxa))

  tibble(
    divergence = max(0, (b_rng - abs(b_max + b_min)) / b_rng),
    n_taxa     = n_taxa
  )
}

h3_div <- h3_taxa_win %>%
  group_by(path_id, date_start) %>%
  group_modify(~ compute_divergence(.x)) %>%
  ungroup() %>%
  left_join(comm_beta_win, by = c("path_id", "date_start")) %>%
  filter(!is.na(divergence), !is.na(community_beta)) %>%
  # Use absolute forcing strength on x-axis
  mutate(
    community_beta = abs(community_beta),
    path_id        = factor(path_id, levels = pathway_order)
  ) %>%
  # IQR-fence outliers per pathway on both axes
  group_by(path_id) %>%
  mutate(
    q1_d     = quantile(divergence,     0.25, na.rm = TRUE),
    q3_d     = quantile(divergence,     0.75, na.rm = TRUE),
    iqr_d    = q3_d - q1_d,
    q1_beta  = quantile(community_beta, 0.25, na.rm = TRUE),
    q3_beta  = quantile(community_beta, 0.75, na.rm = TRUE),
    iqr_beta = q3_beta - q1_beta
  ) %>%
  filter(
    divergence     >= (q1_d    - 1.5 * iqr_d)    &
    divergence     <= (q3_d    + 1.5 * iqr_d),
    community_beta >= (q1_beta - 1.5 * iqr_beta) &
    community_beta <= (q3_beta + 1.5 * iqr_beta)
  ) %>%
  ungroup()
```

## GLS regression: divergence ~ |community β|

```{r h3-gls}
# ══════════════════════════════════════════════════════════════════════════════
# H3 — GLS SLOPES WITH AR(1) TEMPORAL CORRELATION
#
# For each pathway, fit:
#   divergence ~ |community β|
# using GLS with a first-order autoregressive error structure (corAR1) to
# account for the temporal autocorrelation inherent in rolling-window estimates.
#
# A negative slope indicates that higher absolute land-use forcing is associated
# with *lower* response divergence — i.e. forcing compresses taxon responses.
# ══════════════════════════════════════════════════════════════════════════════

gls_results <- h3_div %>%
  filter(path_id %in% pathway_focus) %>%
  group_by(path_id) %>%
  group_modify(~ {
    fit <- tryCatch(
      gls(divergence ~ community_beta,
          data        = .x,
          correlation = corAR1(form = ~ as.numeric(date_start)),
          method      = "REML"),
      error = function(e) NULL
    )
    if (is.null(fit)) return(tibble())

    s <- summary(fit)$tTable

    # Build annotation string: slope + significance stars
    tibble(
      slope = s["community_beta", "Value"],
      se    = s["community_beta", "Std.Error"],
      t     = s["community_beta", "t-value"],
      p     = s["community_beta", "p-value"],
      label = paste0(
        "slope = ", round(s["community_beta", "Value"], 3),
        dplyr::case_when(
          s["community_beta", "p-value"] < 0.001 ~ "***",
          s["community_beta", "p-value"] < 0.01  ~ "**",
          s["community_beta", "p-value"] < 0.05  ~ "*",
          TRUE                                   ~ ""
        )
      )
    )
  }) %>%
  ungroup() %>%
  left_join(path_meta %>% select(path_id, colour), by = "path_id")

# Annotation position: top-right corner of each facet
annot_pos <- h3_div %>%
  group_by(path_id) %>%
  summarise(
    x = min(community_beta, na.rm = TRUE),
    y = max(divergence,     na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(gls_results, by = "path_id") %>%
  mutate(path_id = factor(path_id, levels = pathway_focus))
```

## Temporal dynamics per pathway

```{r h3-dual-panel-helper}
# ══════════════════════════════════════════════════════════════════════════════
# H3 — DUAL-AXIS TEMPORAL PANEL FACTORY
#
# make_dual_panel() produces one mini time-series panel for a single pathway.
# It overlays two LOESS-smoothed time series on a dual y-axis:
#   Primary axis (left, pathway colour) : divergence
#   Secondary axis (right, grey)        : |community β|
#
# Both axes are rescaled to the same coordinate space so the two lines share
# the same plotting area; the secondary axis label converts back to β units.
# Season bands and intervention / irrigation lines are added automatically.
# ══════════════════════════════════════════════════════════════════════════════

#' Build one dual-axis temporal panel for a given pathway.
#'
#' @param pid        Character. One of the pathway_focus IDs.
#' @param show_xaxis Logical. Whether to display x-axis labels (only TRUE for
#'                   the bottom panel in the stacked layout).
#' @return A ggplot object.
make_dual_panel <- function(pid, show_xaxis = TRUE) {
  dat        <- h3_div %>% filter(path_id == pid)
  path_col   <- path_colours[[pid]]
  path_ltype <- path_ltypes[[pid]]

  # Compute axis limits with 5% padding on each side
  d_rng    <- range(dat$divergence,     na.rm = TRUE)
  beta_rng <- range(dat$community_beta, na.rm = TRUE)
  pad_d    <- diff(d_rng)    * 0.05
  pad_beta <- diff(beta_rng) * 0.05
  d_rng    <- d_rng    + c(-pad_d,    pad_d)
  beta_rng <- beta_rng + c(-pad_beta, pad_beta)

  # Linear rescaling functions between the two axis domains
  rescale_b_to_d  <- function(x) d_rng[1] + (x - beta_rng[1]) *
                                    diff(d_rng) / diff(beta_rng)
  rescale_d_to_b  <- function(x) beta_rng[1] + (x - d_rng[1]) *
                                    diff(beta_rng) / diff(d_rng)

  dr <- range(dat$date_start, na.rm = TRUE)

  p <- ggplot(dat, aes(x = date_start)) +
    make_season_layers(dr, y_label_pos = d_rng[2]) +
    # Agricultural events (red dotted) and irrigation (green dotted)
    geom_vline(data = interventions, aes(xintercept = date),
               colour = "darkred",   linetype = "dotted",
               alpha = 0.6, inherit.aes = FALSE) +
    geom_vline(data = irrigation, aes(xintercept = date),
               colour = "darkgreen", linetype = "dotted",
               alpha = 0.6, inherit.aes = FALSE) +
    # Zero-β reference line (rescaled to divergence axis)
    geom_hline(yintercept = rescale_b_to_d(0),
               linetype = "dashed", colour = "grey55", linewidth = 0.3) +
    # Raw divergence points (very transparent)
    geom_point(aes(y = divergence), alpha = 0.20, size = 0.9, colour = path_col) +
    # LOESS: divergence (coloured, pathway linetype)
    geom_smooth(aes(y = divergence),
                method = "loess", span = 0.10, se = FALSE,
                colour = path_col, linetype = path_ltype, linewidth = 0.9) +
    # LOESS: |community β| (grey, solid, thinner)
    geom_smooth(aes(y = rescale_b_to_d(community_beta)),
                method = "loess", span = 0.10, se = FALSE,
                colour = "grey40", linetype = "solid", linewidth = 0.4) +
    scale_y_continuous(
      name     = "Divergence",
      limits   = d_rng,
      sec.axis = sec_axis(transform = rescale_d_to_b,
                          name = "|Community \u03b2|")
    ) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "3 months",
                 expand = expansion(mult = 0.01)) +
    labs(x = NULL) +
    theme_sem(base_size = 10) +
    theme(
      axis.title.y       = element_text(colour = path_col, size = 10),
      axis.title.y.right = element_text(colour = "grey35", size = 9),
      axis.text.y        = element_text(size = 8),
      axis.text.y.right  = element_text(colour = "grey35", size = 8),
      panel.border       = element_blank(),
      axis.line.y.left   = element_line(colour = path_col,  linewidth = 0.4),
      axis.line.y.right  = element_line(colour = "grey35",  linewidth = 0.4),
      axis.line.x.bottom = element_line(colour = "grey20",  linewidth = 0.3),
      plot.margin        = margin(t = 4, r = 5, b = 4, l = 2, unit = "pt")
    )

  if (!show_xaxis) {
    # Suppress x-axis text for all panels except the bottom one
    p <- p + theme(axis.text.x  = element_blank(),
                   axis.ticks.x = element_blank())
  } else {
    p <- p + theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 9))
  }
  p
}
```

## Figure 3 

```{r fig-h3-combined}
#| label: fig-h3-combined
#| fig-cap: >
#|   **Land-use forcing compresses taxonomic response diversity
#|   (Ross et al. divergence).**
#|   (A) Relationship between the Ross et al. divergence index and the absolute
#|   community-level land-use β (|community β|), per pathway. Divergence = 1
#|   when taxon β-values span zero symmetrically; = 0 when all taxa respond in
#|   the same direction. Each point = one rolling window. Lines = OLS fits;
#|   slopes estimated by GLS with AR(1) error structure and annotated with
#|   slope and significance (*** p < 0.001).
#|   (B–H) Temporal co-variation of divergence (coloured line; left axis) and
#|   |community β| (grey line; right axis) per pathway, as LOESS smoothers
#|   (span = 0.10). Shaded bands = meteorological seasons. Dotted red lines =
#|   agricultural interventions; dotted green lines = irrigation events.
#| fig-width: 18
#| fig-height: 30

# ── Left column: scatter of divergence vs. |community β| ─────────────────────
# One facet per pathway (7 facets stacked vertically).
# OLS line shown in grey with 70% CI ribbon; GLS slope annotated top-right.
p_scatter <- ggplot(
  h3_div %>%
    filter(path_id %in% pathway_focus) %>%
    mutate(path_id = factor(path_id, levels = pathway_focus)),
  aes(x = community_beta, y = divergence, colour = path_id)
) +
  facet_wrap(~ path_id, ncol = 1, labeller = as_labeller(path_labels)) +
  geom_point(alpha = 0.25, size = 1.2) +
  geom_smooth(method = "lm", formula = y ~ x,
              colour = "grey20", fill = "grey70",
              linewidth = 0.8, alpha = 0.20) +
  # GLS slope annotation: top-right of each facet
  geom_text(
    data        = annot_pos,
    mapping     = aes(label = label),
    x = Inf, y = Inf, hjust = 1.1, vjust = 1.5,
    inherit.aes = FALSE, size = 3.5,
    colour = "black", fontface = "italic"
  ) +
  scale_colour_manual(values = path_colours, guide = "none") +
  labs(x = "|Community \u03b2|", y = "Divergence (Ross et al.)") +
  theme_sem(base_size = 10) +
  theme(
    strip.text  = element_text(size = 9, face = "bold"),
    axis.text   = element_text(size = 8.5),
    axis.title  = element_text(size = 10),
    plot.margin = margin(t = 5, r = 8, b = 5, l = 5, unit = "pt")
  )

# ── Right column: temporal dual-axis panels per pathway ───────────────────────
# Stacked vertically; only the bottom panel shows x-axis labels.
p_direct       <- make_dual_panel("direct",        show_xaxis = FALSE)
p_indirect     <- make_dual_panel("total_indirect", show_xaxis = FALSE)
p_via_mc1      <- make_dual_panel("via_mc1",        show_xaxis = FALSE)
p_via_mc2      <- make_dual_panel("via_mc2",        show_xaxis = FALSE)
p_via_root     <- make_dual_panel("via_root",       show_xaxis = FALSE)
p_via_mc1_root <- make_dual_panel("via_mc1_root",   show_xaxis = FALSE)
p_via_mc2_root <- make_dual_panel("via_mc2_root",   show_xaxis = TRUE)

temporal_stack <- (p_direct / p_indirect / p_via_mc1 / p_via_mc2 /
                   p_via_root / p_via_mc1_root / p_via_mc2_root)

# ── Assemble final figure ─────────────────────────────────────────────────────
# scatter (width 1) | temporal stack (width 2.5)
# Panel tags: A = scatter; B–H = temporal panels
h3_final <- (p_scatter | temporal_stack) +
  plot_layout(widths = c(1, 2.5)) +
  plot_annotation(tag_levels = "A")

h3_final

ggsave("../output/Fig3_H3_final.tiff", h3_final,
       width = 22, height = 34, units = "cm",
       dpi = 300, compression = "lzw")
```
