---
title: |
  Results and visualisation
author: "Emma Belaud"
format:
  html:
    css: css/styles.css
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
    number-sections: true
execute:
  warning: false
  message: false
  echo: false
---

::: {.callout-note appearance="simple"}
## Purpose of this document

This is the **final step** of the SEM pipeline, following:

- `1_database_edition.qmd` — assembles the SEM database
- `2_SEM_diagnostic.qmd` — selects window duration, transformation, and valid windows
- `3_SEM_modelisation.qmd` — fits all models and writes result databases

**No modelling is performed here.** Every number in every figure comes directly
from pre-computed CSV files. Re-running this script is fast and safe — it never
touches a model.

| Input file | Contents |
|---|---|
| `H1_adjusted_values.csv` | Per-observation null-model residuals with date, land-use label, season |
| `H1_model_summary.csv` | β, SE, t, p, Hedge's g, R² from the two-step H1 model |
| `DAG_coefficients.csv` | Path coefficients from the full-dataset piecewise SEM |
| `DAG_model_fit.csv` | Fisher's C, AIC for the full-dataset piecewise SEM |
| `SEM_results_database.csv` | Rolling-window path coefficients for all taxa |
| `indirect_effects.csv` | Rolling-window indirect effect products for all taxa |
:::

---

# Setup

```{r setup}
# ══════════════════════════════════════════════════════════════════════════════
# SETUP
#
# Loads packages, reads shared pipeline metadata (microclimate variable names),
# and defines all visual constants and helper functions used across sections
# H1, H2, and H3. All aesthetic choices (colours, linetypes, labels) are
# centralised here so that changing one value propagates consistently.
# ══════════════════════════════════════════════════════════════════════════════

invisible(Sys.setlocale("LC_TIME", "C"))    # Force English month names in plots
dir.create("../output", showWarnings = FALSE)

libs <- c("tidyverse", "lubridate", "patchwork",
          "kableExtra", "colorspace", "nlme")
invisible(lapply(libs, require, character.only = TRUE))


# ── Microclimate variable names ───────────────────────────────────────────────
# Written by S1 (1_database_edition.qmd) via PCA-based index selection.
microclimate_index <- readLines("../output/microclimate_index.txt")
mc1_pred <- microclimate_index[1]
mc2_pred <- microclimate_index[2]


# ── Land-use colour palette ───────────────────────────────────────────────────
# Dark green = treed plots (A); burnt orange = cultivated plots (C).
pos_colours <- c("Treed (A)" = "#717F55", "Cultivated (C)" = "#FF994E")


# ── Causal pathway metadata ───────────────────────────────────────────────────
# Single source of truth for pathway labels, colours, and linetypes.
#
# ltype encodes causal depth:
#   solid  = direct or composite (total indirect / total effect)
#   dashed = two-step indirect   (lu → X → fauna)
#   dotted = three-step indirect (lu → X → root → fauna)
path_meta <- tribble(
  ~path_id,         ~label,                                                 ~colour,   ~ltype,
  "direct",         "Direct (lu\u2192fauna)",                               "#F9BB05", "solid",
  "via_mc1",        paste0("lu\u2192", mc1_pred, "\u2192fauna"),            "#BCD094", "solid",
  "via_mc2",        paste0("lu\u2192", mc2_pred, "\u2192fauna"),            "#717F55", "solid",
  "via_root",       "lu\u2192root\u2192fauna",                              "#C45300", "solid",
  "via_mc1_root",   paste0("lu\u2192", mc1_pred, "\u2192root\u2192fauna"),  "#FF7E1F", "solid",
  "via_mc2_root",   paste0("lu\u2192", mc2_pred, "\u2192root\u2192fauna"),  "#FF994E", "solid",
  "total_indirect", "Total indirect",                                        "#999999", "solid",
  "total_effect",   "Total effect",                                          "#444444", "solid"
)

path_colours <- setNames(path_meta$colour, path_meta$path_id)
path_ltypes  <- setNames(path_meta$ltype,  path_meta$path_id)
path_labels  <- setNames(path_meta$label,  path_meta$path_id)


# ── Season background bands ───────────────────────────────────────────────────
season_order <- c("Spring", "Summer", "Autumn", "Winter")
season_fills <- c(Spring = "#f9f9f9", Summer = "#eeeeee",
                  Autumn = "#f9f9f9", Winter = "#eeeeee")


# ── Agricultural intervention and irrigation landmarks ────────────────────────
interventions <- data.frame(
  date = as.Date(c(
    "2024-07-08", "2024-07-15",
    "2024-11-22", "2024-11-25",
    "2025-03-14", "2025-07-10"
  )),
  event = c(
    "Wheat Harvest", "Sorghum Sown",
    "Sorghum Mown",  "Tilled",
    "Chickpea Sown", "Chickpea Harvest"
  )
)

irrigation <- data.frame(
  date = as.Date(c("2024-07-17", "2024-07-22", "2024-08-12"))
)


# ── Helper: season annotation layers ─────────────────────────────────────────
#' Build season background rectangles and italic labels for a time-series plot.
#'
#' @param date_range  Length-2 Date vector matching the x-axis.
#' @param y_label_pos Vertical position of season labels (default = Inf).
#' @param label_size  ggplot font size for season labels. Default 2.2.
#' @return A list of ggplot annotation / geom layers.
make_season_layers <- function(date_range, y_label_pos = Inf, label_size = 2.2) {
  bands <- tibble(
    season  = names(season_fills),
    m_start = c(3L, 6L, 9L, 12L),
    m_end   = c(6L, 9L, 12L,  3L),
    fill    = unname(season_fills)
  ) %>%
    crossing(year = seq(year(date_range[1]), year(date_range[2]))) %>%
    mutate(
      xmin = as.Date(paste(year, m_start, "01", sep = "-")),
      xmax = as.Date(paste(
        if_else(m_end < m_start, year + 1L, year),
        m_end, "01", sep = "-"
      )),
      xmid = xmin + (xmax - xmin) / 2
    ) %>%
    filter(xmin < date_range[2], xmax > date_range[1]) %>%
    mutate(xmin = pmax(xmin, date_range[1]), xmax = pmin(xmax, date_range[2]))

  rect_layers <- pmap(
    bands %>% select(xmin, xmax, fill),
    \(xmin, xmax, fill) annotate(
      "rect", xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf,
      fill = fill, alpha = 1
    )
  )

  label_layer <- geom_text(
    data        = bands,
    aes(x = xmid, y = y_label_pos, label = season),
    inherit.aes = FALSE,
    vjust = 1.6, size = label_size,
    colour = "grey60", fontface = "italic", family = "sans"
  )

  c(rect_layers, list(label_layer))
}


# ── Base ggplot theme ─────────────────────────────────────────────────────────
#' Consistent theme for all SEM figures.
#'
#' @param base_size  Base font size in pt. Default 8.
#' @param legend_pos Passed to `legend.position`. Default "none".
theme_sem <- function(base_size = 8, legend_pos = "none") {
  theme_classic(base_size = base_size) %+replace%
    theme(
      panel.border      = element_rect(colour = "grey20", fill = NA, linewidth = 0.3),
      panel.grid.major  = element_line(colour = "grey94", linewidth = 0.25),
      panel.grid.minor  = element_blank(),
      panel.background  = element_rect(fill = "white", colour = NA),
      axis.line         = element_blank(),
      axis.ticks        = element_line(colour = "grey40", linewidth = 0.3),
      axis.ticks.length = unit(2, "pt"),
      axis.title        = element_text(size = base_size - 0.5, colour = "grey20"),
      axis.text         = element_text(size = base_size - 1.5, colour = "grey30"),
      strip.text        = element_text(size = base_size - 1, face = "bold",
                                       colour = "grey15"),
      strip.background  = element_rect(fill = "grey96", colour = "grey96",
                                       linewidth = 0.3),
      plot.title        = element_text(size = base_size + 0.5, face = "bold",
                                       hjust = 0, colour = "grey10",
                                       margin = margin(b = 3)),
      plot.subtitle     = element_text(size = base_size - 1, colour = "grey45",
                                       hjust = 0, margin = margin(b = 5)),
      plot.caption      = element_text(size = base_size - 2, colour = "grey55",
                                       hjust = 0, margin = margin(t = 4)),
      legend.position   = legend_pos,
      legend.background = element_blank(),
      legend.key        = element_blank(),
      legend.title      = element_text(size = base_size - 1, face = "bold",
                                       colour = "grey20"),
      legend.text       = element_text(size = base_size - 1.5, colour = "grey30"),
      legend.key.size   = unit(8, "pt"),
      legend.spacing.x  = unit(3, "pt"),
      plot.margin       = margin(6, 6, 6, 6)
    )
}


# ── Load pre-computed result databases ────────────────────────────────────────
h1_adj <- read.csv("../output/H1_adjusted_values.csv",
                   na.strings = c("", "NA")) %>%
  mutate(
    position_label = factor(position_label,
                            levels = c("Treed (A)", "Cultivated (C)")),
    season         = factor(season, levels = season_order)
  )

h1_summary <- read.csv("../output/H1_model_summary.csv",
                       na.strings = c("", "NA"))

dag_table <- read.csv("../output/DAG_coefficients.csv",
                      na.strings = c("", "NA"))

dag_fit <- read.csv("../output/DAG_model_fit.csv",
                    na.strings = c("", "NA"))

sem_results <- read.csv("../output/SEM_results_database.csv",
                        na.strings = c("", "NA")) %>%
  mutate(season = factor(season, levels = season_order))

indirect_effects <- read.csv("../output/indirect_effects.csv",
                             na.strings = c("", "NA"))

transfo_fauna <- h1_summary$transformation[1]
```

---

# H1 — Direct effect of land-use type on faunal activity

> **Land-use type directly structures faunal activity, with treed habitats sustaining lower but more temporally stable activity than cultivated zones.**

## Two-step modelling approach

Figures in this section show **null-model residuals** — faunal activity with
the spatial grouping structure removed.

**Step 1 — Remove spatial structure.** A linear mixed model with only an
intercept is fitted on the full dataset. Random effects for orientation nested
within depth absorb all variation attributable to scanner placement, with an
AR(1) correlation structure for temporal autocorrelation:

$$y^* \sim 1 + (1 \mid \text{orientation} / \text{depth}) + \text{AR}(1)$$

**Step 2 — Test the land-use effect on residuals.**

$$\hat{\varepsilon}_{\text{Step 1}} \sim \text{land\_use}$$

β = mean(residuals at C) − mean(residuals at A). A positive β means faunal
activity is higher in cultivated plots after removing orientation and depth
effects.

```{r fig-h1-combined}
#| label: fig-h1-combined
#| fig-cap: >
#|   **Figure 1. Direct effect of land-use type on soil invertebrate activity.**
#|   (A) Distribution of null-model residuals pooled across the monitoring
#|   period. A systematic offset from zero indicates a land-use effect after
#|   accounting for scanner orientation and depth. (B) Temporal dynamics of
#|   null-model residuals with LOESS smoothers (span = 0.18). Green = treed
#|   (A); orange = cultivated (C). Shaded bands = meteorological seasons.
#|   Dotted red vertical lines = agricultural interventions (wheat harvest,
#|   sorghum sowing, mowing, tillage, chickpea sowing). Dotted green vertical
#|   lines = irrigation events.
#| fig-width: 12
#| fig-height: 5

shared_ylim <- c(
  floor(  min(h1_adj$adj_value, na.rm = TRUE) * 10) / 10,
  ceiling(max(h1_adj$adj_value, na.rm = TRUE) * 10) / 10
)

p_box <- ggplot(
  h1_adj,
  aes(x = position_label, y = adj_value,
      fill = position_label, colour = position_label)
) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3,
             linetype = "dashed") +
  geom_jitter(alpha = 0.05, size = 0.3, shape = 20) +
  geom_boxplot(width = 0.5, outliers = FALSE,
               linewidth = 0.8, fatten = 1.2, alpha = 0.5) +
  scale_colour_manual(values = pos_colours) +
  scale_fill_manual(values   = pos_colours) +
  scale_x_discrete(labels = c("Treed", "Cultivated")) +
  coord_cartesian(ylim = shared_ylim) +
  labs(x = NULL, y = "Null-model residuals") +
  guides(fill = "none", colour = "none") +
  theme_sem() +
  theme(
    axis.text.x      = element_text(size = 6.5, face = "bold"),
    panel.border     = element_blank(),
    axis.line.y.left = element_line(colour = "grey20", linewidth = 0.3),
    axis.line.x      = element_line(colour = "grey20", linewidth = 0.3),
    plot.margin      = margin(t = 5, r = 0, b = 5, l = 5, unit = "pt")
  )

p_temp <- ggplot(
  h1_adj %>% mutate(date = as.Date(date)),
  aes(x = date, colour = position_label, fill = position_label)
) +
  make_season_layers(range(as.Date(h1_adj$date), na.rm = TRUE),
                     y_label_pos = shared_ylim[2]) +
  geom_vline(data = interventions, aes(xintercept = date),
             colour = "darkred", linetype = "dotted", alpha = 0.6) +
  geom_vline(data = irrigation, aes(xintercept = date),
             colour = "darkgreen", linetype = "dotted", alpha = 0.6) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3,
             linetype = "dashed") +
  geom_point(aes(y = adj_value), size = 0.5, alpha = 0.15, shape = 16) +
  geom_smooth(aes(y = adj_value), method = "loess", formula = y ~ x,
              span = 0.18, linewidth = 0.7, se = FALSE) +
  scale_colour_manual(values = pos_colours, name = "Land-use type",
                      labels = c("Treed (A)", "Cultivated (C)")) +
  scale_fill_manual(values   = pos_colours, name = "Land-use type",
                    labels   = c("Treed (A)", "Cultivated (C)")) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months",
               expand = expansion(mult = 0.01)) +
  coord_cartesian(ylim = shared_ylim) +
  labs(x = NULL, y = NULL) +
  guides(fill = "none", colour = "none") +
  theme_sem(legend_pos = "bottom") +
  theme(
    axis.text.x       = element_text(angle = 30, hjust = 1, size = 6.5),
    legend.direction  = "horizontal",
    axis.title.y      = element_blank(),
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    panel.border      = element_blank(),
    axis.line.y.right = element_line(colour = "grey20", linewidth = 0.3),
    axis.line.x       = element_line(colour = "grey20", linewidth = 0.3),
    plot.margin       = margin(t = 5, r = 5, b = 5, l = 0, unit = "pt")
  )

h1_combined <- (p_box + p_temp) +
  plot_layout(widths = c(1, 3), guides = "collect") &
  theme(legend.position = "bottom", legend.direction = "horizontal")

h1_combined <- h1_combined + plot_annotation(tag_levels = "A")
h1_combined

ggsave("../output/Fig1_H1_combined.tiff", h1_combined,
       width = 20, height = 10, units = "cm", dpi = 300, compression = "lzw")
```

---

# H2 — Indirect effects through contrasting causal pathways

> **Land-use effects are mediated through contrasting causal pathways — agricultural management drives faunal activity primarily through altered microclimate, whereas treed habitats sustain faunal activity more consistently through root resource supply.**

## Pathway algebra

The SEM estimates one standardised β per path per rolling window. An **indirect
effect** is the product of all β values along a causal route. Because all
continuous variables are z-scored within each window, these products are
dimensionless and directly comparable across windows and pathways.

| Path | Expression |
|---|---|
| Direct | $\beta_9$ |
| lu → mc₁ → fauna | $\beta_1 \times \beta_6$ |
| lu → mc₂ → fauna | $\beta_2 \times \beta_7$ |
| lu → root → fauna | $\beta_5 \times \beta_8$ |
| lu → mc₁ → root → fauna | $\beta_1 \times \beta_3 \times \beta_8$ |
| lu → mc₂ → root → fauna | $\beta_2 \times \beta_4 \times \beta_8$ |
| **Total indirect** | sum of the five routes above |
| **Total effect** | direct + total indirect |

**Sign convention:** A = 0, C = 1. A negative β-product means the effect
through that pathway is stronger in the treed system.

## Table 1 — Full-dataset piecewise SEM path coefficients

```{r tbl-h2-dag}
#| tbl-cap: >
#|   **Table 1.** Path coefficients from the full-dataset piecewise SEM
#|   (all_taxa, complete time series, all depths). All continuous variables
#|   were z-scored after the calibration-selected transformation. β =
#|   unstandardised coefficient; Std β = standardised coefficient directly
#|   comparable to rolling-window values in Figures 2–3. R²m = marginal R²
#|   (fixed effects only); R²c = conditional R² (fixed + random effects).
#|   Fisher's C tests global model fit: p > 0.05 indicates that the
#|   conditional independence claims implied by the DAG are consistent
#|   with the data.

fit_line <- with(dag_fit, paste0(
  "Model fit \u2014 Fisher\u2019s C = ",
  if (is.na(fishers_C)) "NA" else round(fishers_C, 2),
  " (df = ", if (is.na(df)) "NA" else df, ", p ",
  if (is.na(p_value))
    "= NA (saturated model \u2014 no independence claims to test)"
  else if (p_value < 0.001) "< 0.001"
  else paste0("= ", round(p_value, 3)),
  ") \u2502 AIC = ",
  if (is.na(AIC)) "NA" else round(AIC, 1)
))

dag_table %>%
  filter(!str_detect(response, "~~")) %>%
  mutate(
    across(c(estimate, std_estimate, se, t_value, p_value,
             r2_marginal, r2_conditional),
           ~ suppressWarnings(as.numeric(.))),
    across(c(estimate, std_estimate, se), ~ sprintf("%.3f", .)),
    t_fmt   = if_else(is.na(t_value), "\u2014", sprintf("%.2f", t_value)),
    p_fmt   = if_else(p_value < 0.001, "< 0.001", sprintf("%.3f", p_value)),
    r2_marg = sprintf("%.3f", r2_marginal),
    r2_cond = sprintf("%.3f", r2_conditional)
  ) %>%
  select(
    "Equation"   = equation,
    "Predictor"  = predictor,
    "\u03b2"     = estimate,
    "Std \u03b2" = std_estimate,
    "SE"         = se,
    "t"          = t_fmt,
    "p"          = p_fmt,
    "sig"        = sig,
    "R\u00b2m"   = r2_marg,
    "R\u00b2c"   = r2_cond
  ) %>%
  kbl(format = "html", escape = FALSE,
      align  = c("l", "l", "r", "r", "r", "r", "r", "l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("condensed", "hover"),
                full_width = FALSE, font_size = 12) %>%
  row_spec(0, bold = TRUE, background = "#e8e8e8") %>%
  collapse_rows(columns = c(1, 9, 10), valign = "middle") %>%
  footnote(
    general = paste0(
      fit_line,
      "<br>Sig. codes: *** p < 0.001 &nbsp;** p < 0.01 &nbsp;* p < 0.05 &nbsp;. p < 0.10"
    ),
    general_title = "", escape = FALSE, footnote_as_chunk = TRUE
  )
```

## Rolling-window vs. full-dataset pathway estimates

```{r h2-prep}
direct_ts <- sem_results %>%
  filter(taxon == "all_taxa", predictor == "land_use", response == "fauna") %>%
  transmute(path_id = "direct", date_start, season, std_estimate)

indirect_ts <- indirect_effects %>%
  filter(taxon == "all_taxa") %>%
  pivot_longer(
    c(via_mc1, via_mc2, via_root, via_mc1_root, via_mc2_root,
      total_indirect, total_effect),
    names_to  = "path_id",
    values_to = "std_estimate"
  ) %>%
  filter(!is.na(std_estimate)) %>%
  mutate(
    month  = month(date_start),
    season = factor(case_when(
      month %in% 3:5  ~ "Spring",
      month %in% 6:8  ~ "Summer",
      month %in% 9:11 ~ "Autumn",
      TRUE            ~ "Winter"
    ), levels = season_order)
  ) %>%
  select(path_id, date_start, season, std_estimate)

h2_ts <- bind_rows(direct_ts, indirect_ts) %>%
  left_join(path_meta, by = "path_id") %>%
  mutate(
    path_id = factor(path_id, levels = path_meta$path_id),
    label   = factor(label,   levels = path_meta$label),
    season  = factor(season,  levels = season_order)
  ) %>%
  group_by(path_id) %>%
  mutate(
    lo           = quantile(std_estimate, 0.025, na.rm = TRUE),
    hi           = quantile(std_estimate, 0.975, na.rm = TRUE),
    std_estimate = pmax(lo, pmin(hi, std_estimate))
  ) %>%
  ungroup()

smooth_path <- function(df, span = 0.30) {
  df <- df %>% filter(!is.na(std_estimate)) %>% arrange(date_start)
  if (nrow(df) < 5) return(mutate(df, smoothed = std_estimate))
  fit <- tryCatch(
    loess(std_estimate ~ as.numeric(date_start), data = df, span = span,
          control = loess.control(surface = "direct")),
    error = \(e) NULL
  )
  df$smoothed <- if (!is.null(fit)) predict(fit) else df$std_estimate
  df
}

h2_smoothed <- h2_ts %>%
  mutate(date_start = as.Date(date_start)) %>%
  group_by(path_id) %>%
  group_modify(~ smooth_path(.x)) %>%
  ungroup()

date_range_h2 <- range(h2_smoothed$date_start)

shared_ylim_h2 <- c(
  floor(  min(c(h2_ts$std_estimate, h2_smoothed$smoothed), na.rm = TRUE) * 10) / 10,
  ceiling(max(c(h2_ts$std_estimate, h2_smoothed$smoothed), na.rm = TRUE) * 10) / 10
)

path_labels_short <- setNames(
  c("Direct", "mc1\u2192fauna", "mc2\u2192fauna", "root\u2192fauna",
    "mc1\u2192root\n\u2192fauna", "mc2\u2192root\n\u2192fauna",
    "Total\nindirect", "Total\neffect"),
  path_meta$label
)
```

```{r h2-global-coef}
global_coef <- dag_table %>%
  filter(!str_detect(response, "~~")) %>%
  mutate(std_estimate = suppressWarnings(as.numeric(std_estimate)),
         se           = suppressWarnings(as.numeric(se))) %>%
  select(response, predictor, b = std_estimate, se)

get_beta <- function(resp, pred) {
  val <- global_coef$b[global_coef$response == resp & global_coef$predictor == pred]
  if (length(val) == 0) NA_real_ else val[1]
}
get_se <- function(resp, pred) {
  val <- global_coef$se[global_coef$response == resp & global_coef$predictor == pred]
  if (length(val) == 0) NA_real_ else val[1]
}

lu_mc1 <- get_beta("microclimate_1","land_use"); se_lu_mc1 <- get_se("microclimate_1","land_use")
lu_mc2 <- get_beta("microclimate_2","land_use"); se_lu_mc2 <- get_se("microclimate_2","land_use")
mc1_root<- get_beta("root","microclimate_1");   se_mc1_root<- get_se("root","microclimate_1")
mc2_root<- get_beta("root","microclimate_2");   se_mc2_root<- get_se("root","microclimate_2")
lu_root <- get_beta("root","land_use");         se_lu_root <- get_se("root","land_use")
mc1_fau <- get_beta("fauna","microclimate_1");  se_mc1_fau <- get_se("fauna","microclimate_1")
mc2_fau <- get_beta("fauna","microclimate_2");  se_mc2_fau <- get_se("fauna","microclimate_2")
root_fau<- get_beta("fauna","root");            se_root_fau<- get_se("fauna","root")
lu_fau  <- get_beta("fauna","land_use");        se_lu_fau  <- get_se("fauna","land_use")

global_paths <- tribble(
  ~path_id,         ~b_global,                                 ~se_global,
  "direct",          lu_fau,                                    se_lu_fau,
  "via_mc1",         lu_mc1  * mc1_fau,                         se_lu_mc1  * se_mc1_fau,
  "via_mc2",         lu_mc2  * mc2_fau,                         se_lu_mc2  * se_mc2_fau,
  "via_root",        lu_root * root_fau,                        se_lu_root * se_root_fau,
  "via_mc1_root",    lu_mc1  * mc1_root * root_fau,             se_lu_mc1  * se_mc1_root * se_root_fau,
  "via_mc2_root",    lu_mc2  * mc2_root * root_fau,             se_lu_mc2  * se_mc2_root * se_root_fau,
  "total_indirect",  (lu_mc1 * mc1_fau) + (lu_mc2 * mc2_fau),  (se_lu_mc1 * se_mc1_fau) + (se_lu_mc2 * se_mc2_fau),
  "total_effect",    lu_fau  + (lu_mc1  * mc1_fau),             se_lu_fau  + (se_lu_mc1  * se_mc1_fau)
) %>%
  left_join(path_meta, by = "path_id") %>%
  mutate(path_id = factor(path_id, levels = path_meta$path_id),
         label   = factor(label,   levels = path_meta$label))

h2_marginal <- h2_ts %>%
  group_by(path_id, label, colour, ltype) %>%
  summarise(
    median_b = median(std_estimate, na.rm = TRUE),
    sd_b     = sd(std_estimate,     na.rm = TRUE),
    q025     = quantile(std_estimate, 0.025, na.rm = TRUE),
    q975     = quantile(std_estimate, 0.975, na.rm = TRUE),
    q25      = quantile(std_estimate, 0.25,  na.rm = TRUE),
    q75      = quantile(std_estimate, 0.75,  na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  select(-any_of(c("b_global", "se_global"))) %>%
  left_join(select(global_paths, path_id, b_global, se_global), by = "path_id")
```

```{r fig-h2-combined}
#| label: fig-h2-combined
#| fig-cap: >
#|   **Figure 2. Causal pathway decomposition of the land-use effect on soil
#|   invertebrate activity.**
#|   (A) Marginal distribution of standardised β-products across all rolling
#|   windows per causal pathway. Thick bar = IQR; thin bar = 95% range (after
#|   winsorising); open circle = rolling-window median; filled circle =
#|   full time-series piecewise SEM estimate ± 1 SE.
#|   (B) LOESS-smoothed (span = 0.30) time series of standardised β-products.
#|   Shaded bands = meteorological seasons. Dotted red lines = agricultural
#|   interventions; dotted green lines = irrigation events. Sign convention:
#|   A = 0, C = 1 — negative values indicate a stronger effect in the treed
#|   system. Both panels share the same y-axis.
#| fig-width: 16
#| fig-height: 6

shape_values <- c(median = 21, global = 21)

legend_dummy <- tibble(
  x     = c(NA_character_, NA_character_),
  y     = c(NA_real_,      NA_real_),
  shape = c("median",      "global")
)

p_marginal <- ggplot(h2_marginal, aes(colour = path_id, fill = path_id)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey55", linewidth = 0.35) +
  geom_linerange(aes(x = label, ymin = q025, ymax = q975),
                 linewidth = 0.5, alpha = 0.30) +
  geom_linerange(aes(x = label, ymin = q25, ymax = q75),
                 linewidth = 2.2, alpha = 0.60) +
  geom_point(aes(x = label, y = median_b, shape = "median"),
             size = 2.4, fill = "white", stroke = 1.1) +
  geom_linerange(aes(x = label, ymin = b_global - se_global,
                     ymax = b_global + se_global),
                 linewidth = 0.5, alpha = 0.35, na.rm = TRUE) +
  geom_point(aes(x = label, y = b_global, shape = "global", fill = path_id),
             size = 2.2, na.rm = TRUE) +
  geom_point(data    = legend_dummy,
             mapping = aes(x = x, y = y, shape = shape),
             colour  = "grey30", fill = "white",
             size = 2.2, stroke = 1.0, na.rm = TRUE) +
  scale_shape_manual(
    name   = NULL,
    values = shape_values,
    breaks = c("median", "global"),
    labels = c(median = "Rolling median (IQR / 95%)",
               global = "Global model (\u00b1 1 SE)"),
    guide  = guide_legend(
      override.aes = list(fill   = c("white", "grey30"),
                          colour = "grey30",
                          size   = c(2.4, 2.2)))
  ) +
  scale_colour_manual(values = path_colours, guide = "none") +
  scale_fill_manual(values   = path_colours, guide = "none") +
  scale_x_discrete(limits = levels(h2_marginal$label),
                   labels  = path_labels_short) +
  coord_cartesian(ylim = shared_ylim_h2) +
  labs(x = NULL, y = "Standardised \u03b2") +
  theme_sem(legend_pos = "bottom") +
  theme(
    axis.text.x        = element_text(size = 5.5, angle = 40, hjust = 1,
                                      lineheight = 0.85),
    axis.text.y        = element_text(size = 6.5),
    legend.text        = element_text(size = 6.5),
    legend.key.size    = unit(10, "pt"),
    panel.border       = element_blank(),
    axis.line.y.left   = element_line(colour = "grey20", linewidth = 0.3),
    axis.line.x.bottom = element_line(colour = "grey20", linewidth = 0.3),
    plot.margin        = margin(t = 5, r = 0, b = 5, l = 5, unit = "pt")
  )

p_time <- ggplot(
  h2_smoothed %>% mutate(date_start = as.Date(date_start)),
  aes(x = date_start, y = smoothed, colour = path_id, linetype = path_id)
) +
  make_season_layers(date_range_h2, y_label_pos = shared_ylim_h2[2] * 0.9) +
  geom_vline(data = interventions, aes(xintercept = date),
             colour = "darkred", linetype = "dotted", alpha = 0.6) +
  geom_vline(data = irrigation, aes(xintercept = date),
             colour = "darkgreen", linetype = "dotted", alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey55", linewidth = 0.35) +
  geom_line(linewidth = 0.65) +
  scale_colour_manual(values = path_colours, labels = path_labels, name = "Pathway") +
  scale_linetype_manual(values = path_ltypes, labels = path_labels, name = "Pathway") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months",
               expand = expansion(mult = 0.01)) +
  coord_cartesian(ylim = shared_ylim_h2) +
  labs(x = NULL, y = NULL) +
  theme_sem(legend_pos = "bottom") +
  theme(
    axis.text.x        = element_text(angle = 30, hjust = 1, size = 6.5),
    legend.text        = element_text(size = 6.0),
    legend.title       = element_text(size = 6.5, face = "bold"),
    legend.key.width   = unit(16, "pt"),
    legend.key.height  = unit(10, "pt"),
    axis.title.y       = element_blank(),
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    panel.border       = element_blank(),
    axis.line.y.right  = element_line(colour = "grey20", linewidth = 0.3),
    axis.line.x.bottom = element_line(colour = "grey20", linewidth = 0.3),
    plot.margin        = margin(t = 5, r = 5, b = 5, l = 0, unit = "pt")
  )

h2_combined <- (p_marginal + p_time) +
  plot_layout(widths = c(1, 3), guides = "keep")
h2_combined <- h2_combined + plot_annotation(tag_levels = "A")
h2_combined

ggsave("../output/Fig2_H2_combined.tiff", h2_combined,
       width = 24, height = 12, units = "cm", dpi = 300, compression = "lzw")
```

---

# H3 — Land-use forcing compresses taxonomic response diversity

> **Land-use intensification compresses taxonomic response diversity by reducing the divergence among individual taxa in their sensitivity to the land-use contrast.**

## Analysis parameters

```{r h3-params}
pathway_focus <- c(
  "direct", "total_indirect",
  "via_mc1", "via_mc2", "via_root",
  "via_mc1_root", "via_mc2_root"
)
eps           <- 1e-4
pathway_order <- pathway_focus
```

## Data preparation

```{r h3-data}
h3_direct_win <- sem_results %>%
  filter(predictor == "land_use", response == "fauna") %>%
  transmute(taxon, date_start = as.Date(date_start),
            path_id = "direct", std_estimate)

h3_indirect_win <- indirect_effects %>%
  pivot_longer(
    cols      = c(via_mc1, via_mc2, via_root,
                  via_mc1_root, via_mc2_root,
                  total_indirect, total_effect),
    names_to  = "path_id", values_to = "std_estimate"
  ) %>%
  filter(!is.na(std_estimate), path_id %in% pathway_focus) %>%
  mutate(date_start = as.Date(date_start)) %>%
  select(taxon, path_id, date_start, std_estimate)

h3_win_raw <- bind_rows(h3_direct_win, h3_indirect_win) %>%
  filter(path_id %in% pathway_focus)

comm_beta_win <- h3_win_raw %>%
  filter(taxon == "all_taxa") %>%
  select(path_id, date_start, community_beta = std_estimate)

h3_taxa_win <- h3_win_raw %>% filter(taxon != "all_taxa")
```

## CV computation

```{r h3-cv}
# CV = SD(β_taxa) / (mean|β_taxa| + eps)
# Relative dispersion: immune to the mathematical artefact that inflates
# pairwise distances in high-forcing windows.
compute_cv <- function(df) {
  beta_vec <- df$std_estimate
  tibble(cv     = sd(beta_vec, na.rm = TRUE) /
                    (mean(abs(beta_vec), na.rm = TRUE) + eps),
         n_taxa = length(unique(df$taxon)))
}

h3_cv <- h3_taxa_win %>%
  group_by(path_id, date_start) %>%
  group_modify(~ compute_cv(.x)) %>%
  ungroup() %>%
  left_join(comm_beta_win, by = c("path_id", "date_start")) %>%
  filter(!is.na(cv), !is.na(community_beta)) %>%
  mutate(community_beta = abs(community_beta),
         path_id        = factor(path_id, levels = pathway_order)) %>%
  # Remove outliers: 1.5 × IQR fence per pathway on both CV and |community β|
  group_by(path_id) %>%
  mutate(
    q1_cv    = quantile(cv,             0.25, na.rm = TRUE),
    q3_cv    = quantile(cv,             0.75, na.rm = TRUE),
    iqr_cv   = q3_cv - q1_cv,
    q1_beta  = quantile(community_beta, 0.25, na.rm = TRUE),
    q3_beta  = quantile(community_beta, 0.75, na.rm = TRUE),
    iqr_beta = q3_beta - q1_beta
  ) %>%
  filter(cv             >= (q1_cv   - 1.5 * iqr_cv)   &
         cv             <= (q3_cv   + 1.5 * iqr_cv),
         community_beta >= (q1_beta - 1.5 * iqr_beta) &
         community_beta <= (q3_beta + 1.5 * iqr_beta)) %>%
  ungroup()
```

## GLS: CV ~ |community β|

```{r h3-gls}
# GLS with AR(1) error structure: accounts for temporal autocorrelation.
# Negative slope = stronger forcing → lower inter-taxon response diversity.
gls_results <- h3_cv %>%
  filter(path_id %in% pathway_focus) %>%
  group_by(path_id) %>%
  group_modify(~ {
    fit <- tryCatch(
      gls(cv ~ community_beta,
          data        = .x,
          correlation = corAR1(form = ~ as.numeric(date_start)),
          method      = "REML"),
      error = function(e) NULL
    )
    if (is.null(fit)) return(tibble())
    s <- summary(fit)$tTable
    tibble(
      slope = s["community_beta", "Value"],
      se    = s["community_beta", "Std.Error"],
      t     = s["community_beta", "t-value"],
      p     = s["community_beta", "p-value"],
      label = paste0(
        "slope = ", round(s["community_beta", "Value"], 3),
        dplyr::case_when(
          s["community_beta", "p-value"] < 0.001 ~ "***",
          s["community_beta", "p-value"] < 0.01  ~ "**",
          s["community_beta", "p-value"] < 0.05  ~ "*",
          TRUE                                   ~ ""
        )
      )
    )
  }) %>%
  ungroup() %>%
  left_join(path_meta %>% select(path_id, colour), by = "path_id")

annot_pos <- h3_cv %>%
  group_by(path_id) %>%
  summarise(x = min(community_beta, na.rm = TRUE),
            y = max(cv,             na.rm = TRUE), .groups = "drop") %>%
  left_join(gls_results, by = "path_id") %>%
  mutate(path_id = factor(path_id, levels = pathway_focus))
```

## Dual-axis temporal panel helper

```{r h3-dual-panel-helper}
make_dual_panel <- function(pid, show_xaxis = TRUE) {
  dat        <- h3_cv %>% filter(path_id == pid)
  path_col   <- path_colours[[pid]]
  path_ltype <- path_ltypes[[pid]]

  cv_rng   <- range(dat$cv,             na.rm = TRUE)
  beta_rng <- range(dat$community_beta, na.rm = TRUE)
  pad_cv   <- diff(cv_rng)   * 0.05
  pad_beta <- diff(beta_rng) * 0.05
  cv_rng   <- cv_rng   + c(-pad_cv,   pad_cv)
  beta_rng <- beta_rng + c(-pad_beta, pad_beta)

  rescale_b    <- function(x) cv_rng[1] + (x - beta_rng[1]) *
                                 diff(cv_rng) / diff(beta_rng)
  rescale_cv_b <- function(x) beta_rng[1] + (x - cv_rng[1]) *
                                 diff(beta_rng) / diff(cv_rng)

  dr <- range(dat$date_start, na.rm = TRUE)

  p <- ggplot(dat, aes(x = date_start)) +
    make_season_layers(dr, y_label_pos = cv_rng[2]) +
    geom_vline(data = interventions, aes(xintercept = date),
               colour = "darkred", linetype = "dotted",
               alpha = 0.6, inherit.aes = FALSE) +
    geom_vline(data = irrigation, aes(xintercept = date),
               colour = "darkgreen", linetype = "dotted",
               alpha = 0.6, inherit.aes = FALSE) +
    geom_hline(yintercept = rescale_b(0),
               linetype = "dashed", colour = "grey55", linewidth = 0.3) +
    geom_point(aes(y = cv), alpha = 0.20, size = 0.8, colour = path_col) +
    geom_smooth(aes(y = cv),
                method = "loess", span = 0.10, se = FALSE,
                colour = path_col, linetype = path_ltype, linewidth = 0.85) +
    geom_smooth(aes(y = rescale_b(community_beta)),
                method = "loess", span = 0.10, se = FALSE,
                colour = "grey40", linetype = "solid", linewidth = 0.35) +
    scale_y_continuous(
      name     = "CV",
      limits   = cv_rng,
      sec.axis = sec_axis(transform = rescale_cv_b,
                          name = "|Community \u03b2|")
    ) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "3 months",
                 expand = expansion(mult = 0.01)) +
    labs(x = NULL) +
    theme_sem(base_size = 8) +
    theme(
      axis.title.y       = element_text(colour = path_col, size = 9),
      axis.title.y.right = element_text(colour = "grey35", size = 7),
      axis.text.y        = element_text(size = 6),
      axis.text.y.right  = element_text(colour = "grey35", size = 6),
      panel.border       = element_blank(),
      axis.line.y.left   = element_line(colour = path_col, linewidth = 0.4),
      axis.line.y.right  = element_line(colour = "grey35", linewidth = 0.4),
      axis.line.x.bottom = element_line(colour = "grey20", linewidth = 0.3),
      plot.margin        = margin(t = 4, r = 5, b = 4, l = 2, unit = "pt")
    )

  if (!show_xaxis) {
    p <- p + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  } else {
    p <- p + theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 6.5))
  }
  p
}
```

## Figure 3 — Taxonomic response diversity

```{r fig-h3-combined}
#| label: fig-h3-combined
#| fig-cap: >
#|   **Figure 3. Land-use forcing compresses taxonomic response diversity.**
#|   (A) Relationship between the coefficient of variation (CV) of individual
#|   taxon β-estimates and the absolute community-level land-use coefficient
#|   (|community β|), per pathway. Each point = one rolling window. Lines =
#|   OLS fits; slopes estimated by GLS with AR(1) error structure and annotated
#|   with slope and significance (*** p < 0.001). Low CV = convergence of
#|   taxon-specific responses; high CV = divergence.
#|   (B–H) Temporal co-variation of CV (coloured line; left axis) and
#|   |community β| (grey line; right axis) per pathway, as LOESS smoothers
#|   (span = 0.10). Shaded bands = meteorological seasons. Dotted red lines =
#|   agricultural interventions; dotted green lines = irrigation events.
#| fig-width: 16
#| fig-height: 8

p_scatter <- ggplot(
  h3_cv %>%
    filter(path_id %in% pathway_focus) %>%
    mutate(path_id = factor(path_id, levels = pathway_focus)),
  aes(x = community_beta, y = cv, colour = path_id)
) +
  facet_wrap(~ path_id, ncol = 1, labeller = as_labeller(path_labels)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(method = "lm", formula = y ~ x,
              colour = "grey20", fill = "grey70",
              linewidth = 0.75, alpha = 0.20) +
  geom_text(
    data        = annot_pos,
    mapping     = aes(label = label),
    x = Inf, y = Inf, hjust = 1.1, vjust = 1.5,
    inherit.aes = FALSE, size = 3,
    colour = "black", fontface = "italic"
  ) +
  scale_colour_manual(values = path_colours, guide = "none") +
  labs(x = "|Community \u03b2|", y = "CV across taxa") +
  theme_sem(base_size = 8) +
  theme(strip.text  = element_text(size = 7, face = "bold"),
        axis.text   = element_text(size = 6.5),
        plot.margin = margin(t = 5, r = 8, b = 5, l = 5, unit = "pt"))

p_direct      <- make_dual_panel("direct",         show_xaxis = FALSE)
p_indirect    <- make_dual_panel("total_indirect",  show_xaxis = FALSE)
p_via_mc1     <- make_dual_panel("via_mc1",         show_xaxis = FALSE)
p_via_mc2     <- make_dual_panel("via_mc2",         show_xaxis = FALSE)
p_via_root    <- make_dual_panel("via_root",        show_xaxis = FALSE)
p_via_mc1_root<- make_dual_panel("via_mc1_root",    show_xaxis = FALSE)
p_via_mc2_root<- make_dual_panel("via_mc2_root",    show_xaxis = TRUE)

temporal_stack <- (p_direct / p_indirect / p_via_mc1 / p_via_mc2 /
                   p_via_root / p_via_mc1_root / p_via_mc2_root)

h3_final <- (p_scatter | temporal_stack) +
  plot_layout(widths = c(1, 2.5)) +
  plot_annotation(tag_levels = "A")

h3_final

ggsave("../output/Fig3_H3_final.tiff", h3_final,
       width = 20, height = 30, units = "cm",
       dpi = 300, compression = "lzw")
```
