---
title: "From pixels to patterns: high-throughput in-situ imaging unveils soil fauna dynamics over a year in agroforestry systems"
format: html
editor: visual
---

# Data preparation

```{r}
#Package
library(tidyverse)
library(lubridate)
library(ade4)
```

## Fauna data

```{r}
df_raw <- read.csv("../data/fauna_data.csv", sep = ",", na.strings = c("", "NA"))

# Data cleaning
df <- df_raw %>%
  filter(!is.na(Predicted.class))%>%
  filter(!Predicted.class %in% c("under_threshold", "nodule", "SOM", "root_myccorhized", "fungi","unknown", "multi_taxa"))%>%
  mutate(
    position = substr(scanner, 4, 4),  # A, B, ou C
    depth = substr(scanner, 5, 6),      # 20 ou 50
    orientation = substr(scanner, 8, 8)     # O ou E
  )%>%
  mutate(date = ymd_hms(date), 
         date = if_else(minute(date) == 1,
                        date - minutes(1),date), 
         hour = hour(date),minute = minute(date)) %>%
  filter(date> "2024-03-01 00:00:00")%>%
  filter(minute == 0, hour %in% c(0, 6, 12, 18))

taxon_list <- unique(df$Predicted.class)
scanner_list <- unique(df$scanner)

position_palette <- c("A" = "#385438", "B" = "#609060", "C" = "#a8c090")
```

Selection of the most observed taxa (\> 3000 occurrences)

```{r}
taxa_recode <- df %>% mutate(taxa_levelup = case_when( Predicted.class %in% c("pseudosinella", "poduromorpha") ~ "collembola",Predicted.class %in% c("trombidiformes", "mesostigmata") ~ "acari", TRUE ~ Predicted.class))

relevant_taxa <- df %>% 
  group_by(Predicted.class) %>% 
  summarise(total_occurrences = n(), .groups = "drop") %>% 
  filter(total_occurrences >= 3000) %>% 
  select(Predicted.class) 

df_filtered <- df %>% 
  filter(Predicted.class %in% relevant_taxa$Predicted.class)

cat(nrow(relevant_taxa), "selected taxa :", relevant_taxa$Predicted.class)
```

Completion of occurrence data with 0 for non-observations of invertebrates in images and NA for missing images.

```{r}
seq_time <- seq(min(df$date), max(df$date), by = "6 hours")
#Creation of complete grid of dates 
complete_grid <- expand_grid(
  date = seq_time,
  scanner = scanner_list)
#Transition from df to abundance matrix
df_complete <- df_filtered %>%
  group_by(Predicted.class, date, scanner, depth, position, orientation)%>%
  reframe(count= n())%>%
  mutate(presence = 1) %>% 
  pivot_wider(names_from = Predicted.class, values_from = count, values_fill = 0)
#Completion of the matrix
df_complete <- complete_grid %>%
  mutate(position = as.factor(substr(scanner, 4, 4)),  # A, B, ou C
    depth = as.factor(substr(scanner, 5, 6)),      # 20 ou 50
    orientation = as.factor(substr(scanner, 8, 8))     # O ou E
  )%>%
  left_join(df_complete, by = c("date", "scanner", "depth", "position", "orientation"))%>%   
  mutate(presence = replace_na(presence, 0))

#Creation of a total abundance column, but only for the most frequently observed taxa
df_complete <- df_complete %>%
  rowwise() %>%
  mutate(all_taxa = sum(c_across(all_of(relevant_taxa$Predicted.class)), na.rm = FALSE)) %>%
  ungroup()
```

```{r}
df_complete %>%
  mutate(
    year = as.factor(year(date)),
    month = as.factor(lubridate::month(date, label = TRUE, abbr = TRUE))) %>%
  group_by(scanner, year, month, presence) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(scanner, year, month) %>%
  mutate(prop = n / sum(n))%>%
  ggplot(aes(x = month, y = prop, fill = as.factor(presence))) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(scanner ~ year) +
  scale_fill_manual(values = c("0" = "grey80", "1" = "#F4631E"),
                    labels = c("0" = "Missing data", "1" = "Current data")) +
  labs(title = "Proportion of missing data per month and per scanner",x = "month",y = "Proportion",fill = "") +theme_minimal() +
  theme(strip.text = element_text(face = "bold"),axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "bottom")+theme(plot.title = element_text(size = 16))
```

Selection of the aggregate variable total abundance of invertebrates per image.

### Land-use

Two types of land use are compared: under tree cover and herbaceous species, and under cultivated cover. In order to integrate this categorical variable into the structural equation model, it is recoded as a binary variable:

-   0 corresponds to surveys carried out under tree cover and herbaceous species, generally associated with position "A"

-   1 corresponds to surveys carried out under cultivated cover, generally associated with position "C"

The intermediate position "B" is removed.

```{r}
data_sem <- df_complete %>%
  filter(position != "B") %>%
  mutate(depth = as.factor(depth), 
         scanner=as.factor(scanner),
         year = as.factor(year(date)),
         month = floor_date(date, unit = "month"),
         week = floor_date(date, unit = "week")) 
```

### Missing data

We select the dates (time + day of capture) for which we have images from the eight scanners studied.

```{r}
filter_data <- function(data, min_images = 6, apply_filter = TRUE) {
  
  data <- data %>%
    group_by(date) %>%
    mutate(
      images_count = sum(presence),
      dates_to_keep = as.factor(as.integer(images_count > min_images)),
      presence = as.factor(presence)
    ) %>%
    ungroup() %>%
    mutate(
      year = as.factor(year(date)),
      month = as.factor(month(date, label = TRUE, abbr = TRUE))
    )
  
  data_for_plot <- data
  if(apply_filter) {
    data_for_plot <- data_for_plot %>% filter(dates_to_keep == 1)
  }
  
  data_for_plot %>%
    group_by(scanner, year, month, dates_to_keep) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(scanner, year, month) %>%
    mutate(prop = n / sum(n)) %>%
    ggplot(aes(x = month, y = prop, fill = dates_to_keep)) +
    geom_bar(stat = "identity", position = "stack") +
    facet_grid(~ year) +
    scale_fill_manual(values = c("0" = "grey80", "1" = "#00A86B"),
                      labels = c("0" = "Missing data", "1" = "Current data")) +
    labs(title = "Proportion of data retained per month and per scanner",
         x = "Month", y = "Proportion", fill = "") +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom",
      plot.title = element_text(size = 16)
    )
  
    return(data)
}

data_sem <- filter_data(data_sem, min_images = 6, apply_filter = FALSE)
```

## Environmental data

```{r}
soil_probe_data <- read.csv("../data/Diams_AF1W_soil.csv") %>% 
  mutate(date = as.POSIXct(date, format = "%Y-%m-%d %H:%M:%S"), 
         depth=as.factor(depth), 
         position=as.factor(position))

air_probe_data <- read.csv("../data/Diams_AF1W_air.csv")  %>% 
  mutate(date = as.POSIXct(date, format = "%Y-%m-%d %H:%M:%S"), 
         position=as.factor(position))
```

### Micro-climat

Selection of humidity and temperature data

```{r}
data_sem  <- data_sem %>% 
  left_join(soil_probe_data %>% rename(hum=hum_soil, temp=temp_soil) %>% select(depth, position,date, hum, temp),by = c("date" , "depth", "position")) %>%
  filter(dates_to_keep == 1)
```

**But here is the procedure for calculating other microclimatic indices**

Calculation of microclimatic indices according to Grooss et al., in prep: median (represents the average without sensitivity to extremes), amplitude (variability over the period), rate of change (variability in the moment), 97.5% maxima (hot or wet extremes) and 2.5% minima (cold or dry extremes).

```{r}
# Function for calculating climate indices
microclimate_index_compute <- function(vec) {
  if (sum(!is.na(vec)) < 2) {
    return(setNames(rep(NA, 5), c("median", "amplitude", "change_ratio", "max", "min")))
  }
  med <- median(vec, na.rm = TRUE)
  amp <- unname(quantile(vec, 0.95, na.rm = TRUE) - quantile(vec, 0.05, na.rm = TRUE))
  change <- median(diff(round(vec, 6)), na.rm = TRUE)
  max_val <- unname(quantile(vec, 0.975, na.rm = TRUE))
  min_val <- unname(quantile(vec, 0.025, na.rm = TRUE))

  result <- list(
    median = med,
    amplitude = amp,
    change_ratio = change,
    max = max_val,
    min = min_val
  )
  return(list(result))
}
```

```{r}
time_scale = c("year", "month","hour") # day scale
soil_probe_data %>%
  mutate(year = as.factor(year(date)),
         month = as.factor(month(date)),
         week = as.factor(isoweek(date)),
         day = date(date),
         hour = as.factor(hour(date)),
         minutes = as.factor(minute(date)))%>%
  group_by(across(all_of(c("depth", "position", time_scale)))) %>% 
  reframe(across(all_of(c("temp_soil", "hum_soil")), ~microclimate_index_compute(.), .names = "{.col}")) %>%
  unnest_wider(temp_soil, names_sep = "_") %>%
  unnest_wider(hum_soil, names_sep = "_") %>% 
  mutate(across(where(is.numeric), ~ scale(.)[, 1]))%>%
  pivot_longer(cols = where(is.numeric),names_to = "variable",values_to = "value")%>%
  group_by(position, variable, .groups = "drop")%>%
  reframe(mean_value=mean(value))%>%
  ggplot(aes(variable, mean_value, group=position, color = position))+
  geom_point(size=3)+
  geom_line(linewidth=2)+
   ylim(-1, 1)+
  geom_hline(yintercept = 0, linetype = "dashed")+
  scale_color_manual(values = position_palette)+
  labs(y = "centred and reduced mean values", title = paste0("comparison of positional modalities according to different microclimatic indices calculated by ", paste(time_scale,collapse = " ")))+
  theme_minimal()
```

```{r}
microclimate_soil_index <- soil_probe_data %>% 
  mutate(year = as.factor(year(date)),
         month = as.factor(month(date)),
         week = as.factor(isoweek(date)),
         day = date(date),
         hour = as.factor(hour(date)),
         minutes = as.factor(minute(date))) %>%
  group_by(across(all_of(c("position", "depth", time_scale)))) %>% 
  reframe(across(all_of(c("temp_soil", "hum_soil")), ~microclimate_index_compute(.), .names = "{.col}")) %>% 
  unnest_wider(temp_soil, names_sep = "_") %>% 
  unnest_wider(hum_soil, names_sep = "_")
```

✏️️ *Les indices utilisés sont ceux mesurés à l'échelle de la journée. Le graphique présentant les indices mesurés à une autre échelle montrent des résultats différents \> approfondir la question de l'échelle temporelle d’agrégation des données et ses implications écologiques.*

Performing PCA with the 10 microclimatic indices to identify correlation structures between variables and extract a small number of uncorrelated indices that summarise most of the variability of the indices:

```{r}
acp <- dudi.pca(microclimate_soil_index[,6:15], scannf = F, nf = 2)

co <- as.data.frame(acp$co)
co$var <- rownames(co)
eig_percent <- round(100 * acp$eig / sum(acp$eig), 1)
x_lab <- paste0("Dimension 1 (", eig_percent[1], " %)")
y_lab <- paste0("Dimension 2 (", eig_percent[2], " %)")
circle <- data.frame(x = cos(seq(0, 2 * pi, length.out = 100)),
                     y = sin(seq(0, 2 * pi, length.out = 100)))

# Fusion des modalités avec les scores individuels
ind_coords <- as.data.frame(cbind(acp$li, microclimate_soil_index[, 1:2]))

# Moyennes par modalité
depth_means <- aggregate(cbind(Axis1 = Axis1, Axis2 = Axis2) ~ depth, data = ind_coords, FUN = mean)
position_means <- aggregate(cbind(Axis1 = Axis1, Axis2 = Axis2) ~ position, data = ind_coords, FUN = mean)

ggplot() +
  geom_path(data = circle, aes(x = x, y = y), color = "gray70") +
  geom_segment(data = co, aes(x = 0, y = 0, xend = Comp1, yend = Comp2),
               arrow = arrow(length = unit(0.2, "cm")), color = "black") +
  ggrepel::geom_text_repel(data = co, aes(x = Comp1, y = Comp2, label = var),
                  size = 4, color = "black") +
  geom_point(data = depth_means, aes(x = Axis1, y = Axis2), color = "#d89060", size = 3) +
  ggrepel::geom_text_repel(data = depth_means, aes(x = Axis1, y = Axis2, label = depth), color = "#d89060") +
  geom_point(data = position_means, aes(x = Axis1, y = Axis2), color = "#a8c090", size = 3) +
  ggrepel::geom_text_repel(data = position_means, aes(x = Axis1, y = Axis2, label = position), color = "#a8c090") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  coord_fixed() +
  theme_minimal()+
  labs(x = x_lab, y = y_lab, title = "Correlation circle")
```

According to PCA, two highly orthogonal variables that we can select are: daily temperature range (temp_soil_amplitude) and daily minimum temperatures (temp_soil_min).

```{r}
#data_sem  <- data_sem %>% mutate(day = date(date))%>% left_join(microclimate_soil_index %>% select(depth, position,day, temp_soil_amplitude, temp_soil_min),by = c("day" , "depth", "position"))
```

## Roots data

```{r}
dpi <- 1200 
pixels_per_cm <- dpi / 2.54 
root_density <- read.csv("../data/root_pixels_count.csv") 
root_density<- root_density %>%
  mutate(image_name = sub("\\.png$", "", root_density$file)) %>%  
  separate(file, into = c("bloc", "position", "orientation","year","month", "day", "hour", "minute", "fileextension1"), 
           sep = "_|\\.|H|M.| |-", remove = FALSE)%>%
  mutate(scanner = as.factor(paste(position, orientation, sep = "_")),
         depth = as.factor(substr(scanner, 2, 3)),
         position = as.factor(substr(scanner, 1, 1)),  
         minute = if_else(minute == "01", "00", minute))%>%
  filter(hour %in% c("00", "06", "12", "18")) %>%
  mutate(date = ymd_hm(paste(year, month, day, hour, minute)))

root_density <- root_density %>%
  arrange(scanner, date) %>%  
  group_by(scanner) %>%
  mutate(lag_days = 9,
         date_target = date - days(9),
         date_target_24h = date - days(1)) %>%
  ungroup() %>%
  left_join(root_density %>% select(scanner, date, root_pixels_count) %>%
              rename(date_target = date,
                     root_pixels_count_lag = root_pixels_count),
            by = c("scanner", "date_target"), relationship = "many-to-many") %>%
  left_join(root_density %>% select(scanner, date, root_pixels_count) %>%
              rename(date_target_24h = date,
                     root_pixels_count_lag_6h = root_pixels_count),
            by = c("scanner", "date_target_24h"), relationship = "many-to-many") %>%
  mutate(root_growth_pixel_count = root_pixels_count - root_pixels_count_lag,
         root_mature_pixel_count = root_pixels_count - root_growth_pixel_count,
         root_growth_pixel_count_6h = root_pixels_count - root_pixels_count_lag_6h) %>%
  mutate(root_cm2_count = root_pixels_count / (pixels_per_cm^2),
         root_growth_cm2_count = root_growth_pixel_count / (pixels_per_cm^2),
         root_mature_cm2_count = root_mature_pixel_count / (pixels_per_cm^2),
         root_growth_cm2_count_24h = (root_growth_pixel_count_6h / (pixels_per_cm^2))/24)%>%
  mutate(root_growth_cm2_count_24h = case_when(
    root_growth_cm2_count_24h < 0 ~ 0,
    TRUE ~ root_growth_cm2_count_24h
  ))

data_sem <- data_sem %>%   
  left_join(root_density %>% select(date, position, depth, orientation, root_cm2_count, root_growth_cm2_count, root_mature_cm2_count,root_growth_cm2_count_24h), by = c("date", "position", "depth", "orientation"))
```

## Season variable

```{r}
years <- 2023:2025
season <- c("spring", "summer", "autumn", "winter")
seasons <- bind_rows(lapply(years, function(an) {
  tibble(
    season = season,
    date_debut = as.Date(paste0(an, c("-03-21", "-06-21", "-09-21", "-12-21"))),
    date_fin = as.Date(paste0(c(an, an, an, an + 1), c("-06-20", "-09-20", "-12-20", "-03-20"))))
}))

data_sem <- data_sem %>%
  mutate(day = as.Date(date))%>%
  left_join(seasons %>% rowwise() %>% 
  mutate(day = list(seq(date_debut, date_fin, by = "day"))) %>% 
  unnest(cols = c(day)) %>% 
  dplyr::select(day, season), by = "day")
```

## Data standardisation

```{r}
data_sem <- data_sem %>% 
  mutate(across(where(is.numeric), ~ as.numeric(scale(.))))
```

## CSV generation

```{r}
data_sem$date <- format(data_sem$date, "%Y-%m-%d %H:%M:%S")
write.csv(data_sem, "../data/SEM_database.csv", row.names = FALSE, quote = TRUE)
```
