---
title: "From pixels to patterns"
format:
  html:
    css: css/styles.css  
    embed-resources: true
    code-tools: true
    
execute: 
  warning: false
  message: false
  echo: false
---

## Packages and data

```{r}
invisible(Sys.setlocale("LC_TIME", "C"))
if (!dir.exists("../output")) {
  dir.create("../output")
}
invisible(lapply(c("tidyverse", "lubridate", "ade4", "piecewiseSEM", "performance", "patchwork", "slider", "kableExtra", "phylopath", "e1071", "nlme", "zoo"),require,character.only = TRUE))

position_palette <- c("A" = "#5C6F2B",  "C" = "#DE802B")
season_palette <- c('autumn' = '#CA7842', "summer" = "#DDA853", 'winter' = '#8ACCD5', "spring" = "#A4B465" )

# Fauna data
data_sem_results <- read.csv("../output/SEM_results_database.csv", sep = ",", na.strings = c("", "NA")) %>%
  mutate(taxon = case_when(
      taxon == "fauna" ~ "null model",
      TRUE ~ taxon))

# taxa list 
fauna_vars <- readLines("../output/fauna_vars.txt")

# microclimate index 
microclimate_index <- readLines("../output/microclimate_index.txt")
```

Graphical representation of estimates

```{r}
data_sem_results %>%
  filter(response == "fauna")%>%
ggplot(aes(x = day, y = std_estimate, group = position ,color = position)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.4, se = FALSE, size = 1) +
  facet_grid(path_id ~ taxon, scale = "free_y") +
  scale_color_manual(values = position_palette)+
  labs(title = "Seasons of Ecological Decoupling",
    x = "Date", y = "Estimates") +
  theme_minimal() +
  theme(legend.position = "bottom", 
        strip.text = element_text(size = 7, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Temporal stability of ecological "rules"

How consistently environmental factors drive soil fauna activity over time. By calculating the Coefficient of Variation (CV) for each SEM path, we distinguish between "steady" forcing (low CV), where the environment has a predictable impact, and "pulsed" or "unstable" forcing (high CV), where the relationship fluctuates. Comparing these across management positions allows us to test if human intensification destabilizes the ecosystem by replacing natural, steady rhythms with erratic, unpredictable biological responses.

```{r}
# 1. Aggregate Variability with explicit index mapping
cv_global <- data_sem_results %>%
  filter(taxon != "null model")%>%
  group_by(position, path_id) %>%
  summarise(
    n_windows = n(),
    mean_beta = mean(std_estimate, na.rm = TRUE),
    sd_beta = sd(std_estimate, na.rm = TRUE),
    cv_beta = (sd_beta / (abs(mean_beta) + 0.01)) * 100,
    .groups = "drop"
  ) %>%
  filter(n_windows >= 10)

# 2. Lollipop Graph
ggplot(cv_global, aes(x = reorder(path_id, cv_beta), y = cv_beta, color = position)) +
  # Use geom_linerange for perfect dodging with points
  geom_linerange(aes(ymin = 0, ymax = cv_beta), 
                 position = position_dodge(width = 0.6), 
                 linewidth = 1, alpha = 0.4) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  coord_flip() +
  scale_color_manual(values = position_palette) +
  labs(
    title = "Stability of Ecological Forcing Rules",
    subtitle = "Higher CV indicates 'pulsed' or unstable forcing dynamics",
    x = "SEM Pathways",
    y = "Volatility (Coefficient of Variation %)",
    color = "Management Position"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10),
    legend.position = "bottom")
```

> **Pistes amélioration ou développement ?**
>
> générer CV selon plus de facteurs pour pouvoir faire tests de comparaison
>
> est-ce le CV est le meilleur indicateur de variation/stabilité temporelle ?
>
> ne sélectionner au préalable que les relations significatives ?
>
> décortiquer davantage les variations temporelles en identifiant les transitions abruptes ?

## Spatial Synchrony of Forcing

hen Temperature becomes a strong driver in the "Natural" zone, does it also become a strong driver in the "Anthropized" zone?

-   High Correlation ($r \approx 1$): The management hasn't changed the "timing" of the rule. The forcing remains governed by regional climate.

-   Low/No Correlation ($r \approx 0$): Management has "decoupled" the site from the natural rhythm. The rules of the game in the anthropized zone are now independent of the natural ones.

```{r}
# 1. Calculate the "Synchrony Index" (Correlation A vs C)
sync_results <- data_sem_results %>%
  select(day, position, taxon, path_id, std_estimate)%>%
  pivot_wider(names_from = position, values_from = std_estimate) %>%
  group_by(path_id, taxon) %>%
  summarise(
    r_A_vs_C = cor(A, C, use = "pairwise.complete.obs"),
    n_obs = n()
  ) %>%
  filter(!is.na(r_A_vs_C))
print(sync_results, n=15)
# 2. Visualization: Forcing Decoupling Lollipop
ggplot(sync_results, aes(x = reorder(path_id, r_A_vs_C), y = r_A_vs_C)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "#385438", size = 0.5) +
  geom_segment(aes(xend = reorder(path_id, r_A_vs_C), y = 0, yend = r_A_vs_C), color = "grey70", size = 1) +
  # The point color changes based on synchrony strength
  geom_point(aes(color = r_A_vs_C), size = 5) +
  scale_color_gradient2(low = "#B2182B", mid = "#F7F7F7", high = "#2166AC", midpoint = 0) +
  coord_flip() +
  labs(
    title = "Ecological Decoupling: Pearson correlation (r) of path coefficients over time",
    x = "Forcing Pathway",
    y = "Synchrony Index",
    color = "Synchrony"
  ) +
  theme_minimal() +
  facet_wrap(~taxon) +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(face = "bold"))
```

```{r}
# Function to calculate rolling correlation between A and C
calculate_rolling_sync <- function(df, window_size = 21) {
  df %>%
    arrange(day) %>%
    mutate(roll_cor = slider::slide2_dbl(A, C, 
                                         ~cor(.x, .y, use = "pairwise.complete.obs"), 
                                         .before = window_size))
}

# Assurez-vous que results_all_taxa est bien chargé
position_sync_binary <- data_sem_results %>%
  # Création du nom du chemin (ex: temp_soil_max -> fauna)
  mutate(path_id = paste(predictor, "->", response)) %>%
  # Remplacement des noms génériques par les vrais noms du microclimat
  mutate(path_id = str_replace(path_id, "microclimate_index_1", microclimate_index[1]),
         path_id = str_replace(path_id, "microclimate_index_2", microclimate_index[2])) %>%
  # Sélection et passage en format large pour avoir A et C côte à côte
  select(day, position, taxon, path_id, estimate) %>%
  pivot_wider(names_from = position, values_from = estimate)


# Apply to your paths
rolling_sync_data <- position_sync_binary %>%
  group_by(path_id, taxon) %>%
  nest() %>%
  mutate(rolling_cor = map(data, calculate_rolling_sync)) %>%
  select(-data) %>%
  unnest(rolling_cor)

# Visualization of the flip
# Filter out the NAs generated by zero-variance windows and smooth the lines
ggplot(rolling_sync_data %>% filter(!is.na(roll_cor)), 
       aes(x = day, y = roll_cor, group = taxon,color = taxon)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", alpha = 0.5) +
  # Use geom_smooth to see the trend through the noise of daily fluctuations
  geom_smooth(method = "loess", span = 0.2, se = FALSE, size = 1) +
  # Add a rug to see data density
  geom_rug(alpha = 0.1) +
  facet_wrap(~ path_id, scale = "free") +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Seasons of Ecological Decoupling",
    subtitle = "Positive r: Coordinated forcing | Negative r: Antagonistic forcing (A vs C)",
    x = "Date", y = "Rolling Correlation (r)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", 
        strip.text = element_text(size = 7, face = "bold"),
        axis.text.x = element_blank())
```
